---
title: "Build reduced data set"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "00_build_pasture_data" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "78b66e9f-b0b1-4206-8f3d-e2c578a062bf")
```

In this script, we load the original PASTURE data (not published) and build the reduced data set, which is the basis for all PASTURE-related analyses conducted in this project. 

```{r packages}
library(conflicted)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(kableExtra)
library(pheatmap)
library(ggrepel)
library(patchwork)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Load data & set parameters

```{r load_data}
phylo_genus <- readRDS(path_source("PASTURE_data", paste0("bact_2m_genus.rds")))

phylo_genus
```

## Select variables of interest

We keep only the IDs and the variable on duration of exclusive breast feeding.
The sample IDs are not the original ones, but randomly generated numbers. 

```{r}
keep_vars  <- c("SampleID", "idx", "breast_excl_cat1")

sample_data(phylo_genus) <- sample_data(phylo_genus) |>
  as("data.frame") |>
  select(all_of(keep_vars))

# Extract sampdata if needed downstream
sampdata <- sample_data(phylo_genus) |> as("data.frame")
head(sampdata)
```

### Genus level

```{r}
phylo_genus
```


```{r}
# Save data
saveRDS(phylo_genus, path_target("bact_2m_genus.rds"))
```

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
