---
title: "Differential distribution analysis of single-cell gene expression using `waddR` and `permApprox`"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
always_allow_html: true
params:
  name: "01_sc_diff_distr" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "3f0a7f60-7a25-4421-89f9-e2d5fe4071a7")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 10, 
  fig.height = 6, 
  dpi = 200,
  fig.path = paste0('figures/', params, "/app_ddistr_sc_"),
  echo = FALSE)
```


```{r packages, message=FALSE, warning=FALSE}
# Load packages
library("conflicted")

library(waddR)
library(SingleCellExperiment)
library(BiocFileCache)
library(BiocParallel)

library(ggplot2)
library(patchwork)
library(pheatmap)
library(tibble)
library(dplyr)
library(tidyr)
library(parallel)
library(arm)
library(eva)
library(furrr)
library(progressr)
library(kableExtra)

#remotes::install_github("stefpeschel/permApprox")
library(permApprox)

conflicts_prefer(dplyr::count, dplyr::filter, dplyr::setdiff, dplyr::select,
                 dplyr::desc)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

## Description

Single-cell RNA sequencing (scRNA-seq) data often exhibit complex distributional differences between experimental conditions. Standard differential expression methods can miss these patterns, especially when they involve changes in the shape or spread of the distribution, not just the mean.

The `waddR` package provides a semi-parametric test based on the 2-Wasserstein distance, designed to detect such differences. However, its permutation-based p-values can be inaccurate when the number of permutations is limited, leading to zero p-values.

In this application study, we analyze a subset of the Vento-Tormo et al. (2018) dataset to demonstrate how `permApprox` can be used to accurately approximate small permutation p-values from the `waddR` method, particularly addressing zero p-values and improving inference robustness.

In `waddR`, the one-stage (OS) approach compares the complete distributions between groups in a single step, using the squared Wasserstein distance on all observed values, including zeros. In contrast, the two-stage (TS) approach separates sparsity from magnitude differences: in the first stage it tests for differences in the proportion of zero values between groups, and in the second stage it computes the Wasserstein distance only on the non-zero values. The evidence from both stages is then combined into one permutation-based p-value.

We concentrate on the two-stage (TS) approach here.

## Own waddR implementation

We needed to adapt the waddR package to export the permutation test statistics. 

```{r, message=FALSE, warning=FALSE}
# Load waddR functions
proj_path <- rprojroot::find_root(rprojroot::has_file("README.md"))
waddR_path <- file.path(proj_path, "functions/waddR/R")
waddR_functions <- list.files(waddR_path, pattern = "\\.R$", full.names = TRUE)
invisible(sapply(waddR_functions, source))
```

## Data and Preprocessing

We analyze decidua and blood samples from a normalized replicate of the Vento-Tormo dataset. The gene expression data are available as two `SingleCellExperiment` objects. 

We select the 1000 most prevalent genes for our analysis.


```{r}
# Loading and subsetting data
data_url <- "https://github.com/goncalves-lab/waddR-data/blob/master/data/"
sce_blood_url <- paste0(data_url, "data_blood.rda?raw=true")
sce_decidua_url <- paste0(data_url, "data_decidua.rda?raw=true")

getCachedPath <- function(url, rname){
  bfc <- BiocFileCache() # fire up cache
  res <- bfcquery(bfc, url, field="fpath", exact=TRUE)
  if (bfccount(res) == 0L)
    cachedFilePath <- bfcadd(bfc, rname=rname, fpath=url)
  else
    cachedFilePath <- bfcpath(bfc, res[["rid"]])
  cachedFilePath
}

load(getCachedPath(sce_blood_url, "data_blood"))
load(getCachedPath(sce_decidua_url, "data_decidua"))
```

```{r}
counts_blood <- counts(data_blood)
counts_decidua <- counts(data_decidua)

cat("Dimensions of blood counts: ", dim(counts_blood))
cat("\nDimensions of decidua counts: ", dim(counts_decidua))
```

```{r}
# Calculate prevalence (number of non-zero counts) per gene in each dataset
prev_blood <- Matrix::rowSums(counts_blood > 0)
prev_decidua <- Matrix::rowSums(counts_decidua > 0)

# Total prevalence across both conditions
prev_total <- prev_blood + prev_decidua

# Select top 1000 most prevalent genes
top_genes <- names(sort(prev_total, decreasing = TRUE))[1:1000]

# Subset SCE objects
sce_blood <- data_blood[top_genes, ]
sce_decidua <- data_decidua[top_genes, ]

# Number of cells
n_cells_blood <- dim(colData(sce_blood))[1]
n_cells_decidua <- dim(colData(sce_decidua))[1]
```

To eliminate potential effects of different sample sizes, we subsample from the data set with the larger number of samples (decidua) so that both data sets have the same number of cells. 

```{r}
# Number of cells per condition
n_cells_blood    <- ncol(sce_blood)
n_cells_decidua  <- ncol(sce_decidua)

# Target size = min of the two
n_target <- min(n_cells_blood, n_cells_decidua)

# We subsample from the larger group (decidua)
set.seed(123) 
keep_decidua <- sample(ncol(sce_decidua), n_target)
sce_decidua  <- sce_decidua[, keep_decidua, drop = FALSE]

dim(sce_blood)    # genes x cells
dim(sce_decidua)  # genes x cells
```

```{r}
# Clean up
rm(data_blood, data_decidua, counts_blood, counts_decidua)
gc()
```


## Exploratory Analysis

We visualize selected genes to explore differences in expression distributions between conditions.

```{r}
## Global color palette (consistent across all plots)
col_group1 <- "darkorange2"  # blood
col_group2 <- "#1f77b4"  # decidua
palette_groups <- c("blood" = col_group1, "decidua" = col_group2)
```


```{r}
# Utility function to shorten gene names
make_short_gene <- function(gene_vec) {
  # If the string contains "-", remove the trailing Ensembl ID
  sub("-ENSG[0-9]+.*$", "", gene_vec)
}
```

### Heatmap

```{r heatmap}
# Select top 20 random genes
topgenes <- top_genes[1:20]

# Combine expression matrices (logcounts)
expr_mat <- cbind(
  assay(sce_blood)[topgenes, ],
  assay(sce_decidua)[topgenes, ]
)

# Add cell labels
col_ann <- data.frame(
  tissue = factor(rep(c("blood", "decidua"),
                      c(ncol(sce_blood), ncol(sce_decidua))))
)
rownames(col_ann) <- colnames(expr_mat)

# Shorten gene names
rownames(expr_mat) <- make_short_gene(rownames(expr_mat))

# Correct color list
annotation_colors = list(tissue = palette_groups)

pheatmap(
  log1p(expr_mat),
  annotation_col = col_ann,
  annotation_colors = annotation_colors,
  show_rownames = TRUE,
  show_colnames = FALSE,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = "Gene expression (Top 20 genes)"
)
```

### Density plots

```{r}
# Select a few genes (e.g., 5)
genes_to_plot <- top_genes[1:6]

# Create a tidy data frame with log1p-transformed expression values
plot_df <- dplyr::bind_rows(
  lapply(genes_to_plot, function(gene) {
    tibble(
      gene = gene,
      expression = log1p(assay(sce_blood)[gene, ]),
      tissue = "blood"
    ) %>%
      dplyr::bind_rows(tibble(
        gene = gene,
        expression = log1p(assay(sce_decidua)[gene, ]),
        tissue = "decidua"
      )) %>%
      mutate(gene = make_short_gene(gene))
  })
)
```


```{r density_top6, fig.width=12, fig.height=5}
# Density
ggplot(plot_df, aes(x = expression, fill = tissue)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = palette_groups) +
  facet_wrap(~ gene, scales = "free") +
  labs(title = "Gene expression density", 
       y = "Density",
       x = "Expression") +
  theme_minimal() +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 16),
        strip.text = element_text(size = 14)
  )
```

### Histograms

```{r hist_top6, fig.width=12, fig.height=5}
ggplot(plot_df, aes(x = expression, fill = tissue)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  scale_fill_manual(values = palette_groups) +
  facet_wrap(~ gene, scales = "free") +
  labs(#title = "Gene expression histograms", 
    y = "Count",
    x = "Expression") +
  theme_minimal() +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 16),
        strip.text = element_text(size = 14)
  )
```

### Quantile plot

```{r quantf_top6, fig.width=12, fig.height=5}
# Quantile function
ggplot(plot_df, aes(x = expression, color = tissue)) +
  stat_ecdf(geom = "step", size = 1) +
  scale_color_manual(values = palette_groups) +
  facet_wrap(~ gene, scales = "free") +
  labs(
    #title = "Quantile Function",
    x = "Expression",
    y = "Cumulative Probability"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 16),
        strip.text = element_text(size = 14)
  )
```

## Differential Testing (two-stage approach)

### waddR - 1000 permutations

We apply the two-stage (`TS`) approach in `waddR`, which separately tests for differences in non-zero expression values and in the proportion of zero values.

```{r, eval=FALSE}
register(SerialParam())  # Disable parallelism

waddr_res_ts <- wassersteinEdit.sc(sce_blood, sce_decidua, method = "TS", 
                                   permnum = 1000, seed = 42)

saveRDS(waddr_res_ts, file = path_target("waddr_results_ts_1000.rds"))
```


```{r}
waddr_res_ts_1k <- readRDS(path_target("waddr_results_ts_1000.rds"))
perm_stats_ts_1k <- t(attributes(waddr_res_ts_1k)$wass.values)
colnames(perm_stats_ts_1k) <- make_short_gene(colnames(perm_stats_ts_1k))


waddr_res_ts_1k <- waddr_res_ts_1k %>%
  as_tibble() %>%
  dplyr::mutate(idx = row_number(),
         gene = row.names(waddr_res_ts_1k)) %>%
  dplyr::select(idx, gene, everything()) %>%
  mutate(gene = make_short_gene(gene))

head(waddr_res_ts_1k, n=10L)
```


```{r}
res <- waddr_res_ts_1k %>% dplyr::filter(!is.na(pval))

cat("Number of zero p-values with waddR: ", 
    sum(res$pval == 0, na.rm = TRUE))
cat("\nNumber of non-NA p-values: ", nrow(res))
cat("\nProportion of zero p-values with waddR: ", 
    sum(res$pval == 0, na.rm = TRUE) / nrow(res))
cat("\nNumber of significant tests (unadjusted): ", 
    sum(res$pval < 0.05, na.rm = TRUE))
cat("\nProportion of significant tests (unadjusted): ", 
    sum(res$pval < 0.05, na.rm = TRUE) / nrow(res))
cat("\nNumber of significant tests (adjusted): ", 
    sum(res$p.adj.nonzero < 0.05, na.rm = TRUE))
cat("\nProportion of significant tests (adjusted): ", 
    sum(res$p.adj.nonzero < 0.05, na.rm = TRUE) / nrow(res))
```

### waddR - 10000 permutations

We repeat the waddR run with 10000 permutations to compare the accuracy with the 1000 permuations later. 

```{r, eval=FALSE}
register(SerialParam())  # Disable parallelism

waddr_res_ts <- wassersteinEdit.sc(sce_blood, sce_decidua, method = "TS", 
                                   permnum = 10000, seed = 42)

saveRDS(waddr_res_ts, file = path_target("waddr_results_ts_10000.rds"))
```


```{r}
waddr_res_ts_10k <- readRDS(path_target("waddr_results_ts_10000.rds"))
perm_stats_ts_10k <- t(attributes(waddr_res_ts_10k)$wass.values)
colnames(perm_stats_ts_10k) <- make_short_gene(colnames(perm_stats_ts_10k))

waddr_res_ts_10k <- waddr_res_ts_10k %>%
  as_tibble() %>%
  dplyr::mutate(idx = row_number(),
         gene = row.names(waddr_res_ts_10k)) %>%
  dplyr::select(idx, gene, everything()) %>%
  mutate(gene = make_short_gene(gene))
```


```{r}
res <- waddr_res_ts_10k %>% dplyr::filter(!is.na(pval))

cat("Number of zero p-values with waddR: ", 
    sum(res$pval == 0, na.rm = TRUE))
cat("\nNumber of non-NA p-values: ", nrow(res))
cat("\nProportion of zero p-values with waddR: ", 
    sum(res$pval == 0, na.rm = TRUE) / nrow(res))
cat("\nNumber of significant tests (unadjusted): ", 
    sum(res$pval < 0.05, na.rm = TRUE))
cat("\nProportion of significant tests (unadjusted): ", 
    sum(res$pval < 0.05, na.rm = TRUE) / nrow(res))
cat("\nNumber of significant tests (adjusted): ", 
    sum(res$p.adj.nonzero < 0.05, na.rm = TRUE))
cat("\nProportion of significant tests (adjusted): ", 
    sum(res$p.adj.nonzero < 0.05, na.rm = TRUE) / nrow(res))
```

### waddR - 50000 permutations

We repeat the waddR run with 50000 permutations to compare the accuracy with the 1000 permuations later. 

```{r, eval=FALSE}
register(SerialParam())  # Disable parallelism

waddr_res_ts <- wassersteinEdit.sc(sce_blood, sce_decidua, method = "TS", 
                                   permnum = 50000, seed = 42)

saveRDS(waddr_res_ts, file = path_target("waddr_results_ts_50000.rds"))
```


```{r}
waddr_res_ts_50k <- readRDS(path_target("waddr_results_ts_50000.rds"))
perm_stats_ts_50k <- t(attributes(waddr_res_ts_50k)$wass.values)
colnames(perm_stats_ts_50k) <- make_short_gene(colnames(perm_stats_ts_50k))

waddr_res_ts_50k <- waddr_res_ts_50k %>%
  as_tibble() %>%
  dplyr::mutate(idx = row_number(),
         gene = row.names(waddr_res_ts_50k)) %>%
  dplyr::select(idx, gene, everything()) %>%
  mutate(gene = make_short_gene(gene))
```


```{r}
res <- waddr_res_ts_50k %>% dplyr::filter(!is.na(pval))

cat("Number of zero p-values with waddR: ", 
    sum(res$pval == 0, na.rm = TRUE))
cat("\nNumber of non-NA p-values: ", nrow(res))
cat("\nProportion of zero p-values with waddR: ", 
    sum(res$pval == 0, na.rm = TRUE) / nrow(res))
cat("\nNumber of significant tests (unadjusted): ", 
    sum(res$pval < 0.05, na.rm = TRUE))
cat("\nProportion of significant tests (unadjusted): ", 
    sum(res$pval < 0.05, na.rm = TRUE) / nrow(res))
cat("\nNumber of significant tests (adjusted): ", 
    sum(res$p.adj.nonzero < 0.05, na.rm = TRUE))
cat("\nProportion of significant tests (adjusted): ", 
    sum(res$p.adj.nonzero < 0.05, na.rm = TRUE) / nrow(res))
```

### waddR - 100000 permutations

We repeat the waddR run with 100000 permutations to compare the accuracy with the 1000 permuations later. 

```{r, eval=FALSE}
register(SerialParam())  # Disable parallelism

waddr_res_ts <- wassersteinEdit.sc(sce_blood, sce_decidua, method = "TS", 
                                   permnum = 100000, seed = 42)

saveRDS(waddr_res_ts, file = path_target("waddr_results_ts_100000.rds"))
```


```{r}
waddr_res_ts_100k <- readRDS(path_target("waddr_results_ts_100000.rds"))
perm_stats_ts_100k <- t(attributes(waddr_res_ts_100k)$wass.values)
colnames(perm_stats_ts_100k) <- make_short_gene(colnames(perm_stats_ts_100k))

waddr_res_ts_100k <- waddr_res_ts_100k %>%
  as_tibble() %>%
  dplyr::mutate(idx = row_number(),
         gene = row.names(waddr_res_ts_100k)) %>%
  dplyr::select(idx, gene, everything()) %>%
  mutate(gene = make_short_gene(gene))
```


```{r}
res <- waddr_res_ts_100k %>% dplyr::filter(!is.na(pval))

cat("Number of zero p-values with waddR: ", 
    sum(res$pval == 0, na.rm = TRUE))
cat("\nNumber of non-NA p-values: ", nrow(res))
cat("\nProportion of zero p-values with waddR: ", 
    sum(res$pval == 0, na.rm = TRUE) / nrow(res))
cat("\nNumber of significant tests (unadjusted): ", 
    sum(res$pval < 0.05, na.rm = TRUE))
cat("\nProportion of significant tests (unadjusted): ", 
    sum(res$pval < 0.05, na.rm = TRUE) / nrow(res))
cat("\nNumber of significant tests (adjusted): ", 
    sum(res$p.adj.nonzero < 0.05, na.rm = TRUE))
cat("\nProportion of significant tests (adjusted): ", 
    sum(res$p.adj.nonzero < 0.05, na.rm = TRUE) / nrow(res))
```

### Empirical p-values

```{r}
fname_ts_emp_1k <- path_target(paste0("permapprox_results_ts_emp_1k.rds"))
fname_ts_emp_10k <- path_target(paste0("permapprox_results_ts_emp_10k.rds"))
fname_ts_emp_50k <- path_target(paste0("permapprox_results_ts_emp_50k.rds"))
fname_ts_emp_100k <- path_target(paste0("permapprox_results_ts_emp_100k.rds"))

if (!file.exists(fname_ts_emp_1k)) {
  papprox_ts_emp_1k <- perm_approx(
    obs_stats = waddr_res_ts_1k$`d.wass^2`,
    perm_stats = perm_stats_ts_1k,
    method = "empirical",
    null_center = "mean"
  )
  saveRDS(papprox_ts_emp_1k, file = fname_ts_emp_1k)
} else {
  papprox_ts_emp_1k <- readRDS(fname_ts_emp_1k)
}

if (!file.exists(fname_ts_emp_10k)) {
  papprox_ts_emp_10k <- perm_approx(
    obs_stats = waddr_res_ts_10k$`d.wass^2`,
    perm_stats = perm_stats_ts_10k,
    method = "empirical",
    null_center = "mean"
  )
  saveRDS(papprox_ts_emp_10k, file = fname_ts_emp_10k)
} else {
  papprox_ts_emp_10k <- readRDS(fname_ts_emp_10k)
}

if (!file.exists(fname_ts_emp_50k)) {
  papprox_ts_emp_50k <- perm_approx(
    obs_stats = waddr_res_ts_50k$`d.wass^2`,
    perm_stats = perm_stats_ts_50k,
    method = "empirical",
    null_center = "mean"
  )
  saveRDS(papprox_ts_emp_50k, file = fname_ts_emp_50k)
} else {
  papprox_ts_emp_50k <- readRDS(fname_ts_emp_50k)
}

if (!file.exists(fname_ts_emp_100k)) {
  papprox_ts_emp_100k <- perm_approx(
    obs_stats = waddr_res_ts_100k$`d.wass^2`,
    perm_stats = perm_stats_ts_100k,
    method = "empirical",
    null_center = "mean"
  )
  saveRDS(papprox_ts_emp_100k, file = fname_ts_emp_100k)
} else {
  papprox_ts_emp_100k <- readRDS(fname_ts_emp_100k)
}
```

### permApprox - unconstrained

```{r}
fname_ts_unconstr1k <- path_target(paste0("permapprox_results_ts_unconstr_1k.rds"))

if (!file.exists(fname_ts_unconstr1k)) {
  
  gpd_ctrl <- make_gpd_ctrl(
    fit_method = "lme",
    tol = 1e-7,
    include_obs = FALSE,
    constraint = "unconstrained",
    thresh_method = "rob_ftr",
    exceed0 = 0.25,
    exceed0_min = 250,
    exceed_min = 10,
    thresh_step = 10,
    gof_test = "ad",
    gof_alpha = 0.05
  )
  
  papprox_ts_unconstr_1k <- perm_approx(
    obs_stats = waddr_res_ts_1k$`d.wass^2`,
    perm_stats = perm_stats_ts_1k,
    null_center = "mean",
    alternative = "greater",
    method = "gpd",
    approx_thresh = 1,
    gpd_ctrl = gpd_ctrl,
    adjust_method = "BH",
    cores = 10,
    verbose = TRUE
  )
  
  saveRDS(papprox_ts_unconstr_1k, file = fname_ts_unconstr1k)
} else {
  papprox_ts_unconstr_1k <- readRDS(fname_ts_unconstr1k)
}

summary(papprox_ts_unconstr_1k)
```


```{r}
fname_ts_unconstr10k <- path_target(paste0("permapprox_results_ts_unconstr_10k.rds"))

if (!file.exists(fname_ts_unconstr10k)) {
papprox_ts_unconstr_10k <- perm_approx(
  obs_stats = waddr_res_ts_10k$`d.wass^2`,
  perm_stats = perm_stats_ts_10k,
  null_center = "mean",
  alternative = "greater",
  method = "gpd",
  approx_thresh = 1,
  gpd_ctrl = gpd_ctrl,
  adjust_method = "BH",
  cores = 10,
  verbose = TRUE
)

  saveRDS(papprox_ts_unconstr_10k, file = fname_ts_unconstr10k)
} else {
  papprox_ts_unconstr_10k <- readRDS(fname_ts_unconstr10k)
}

summary(papprox_ts_unconstr_10k)
```

### permApprox - constrained

We use the `permApprox` package to approximate p-values from permutation distributions of `waddR` test statistics.

```{r}
sampsize <- min(n_cells_blood, n_cells_decidua)

fname_ts_constr1k <- path_target(paste0("permapprox_results_ts_constr_1k.rds"))

if (!file.exists(fname_ts_constr1k)) {
  
  gpd_ctrl <- make_gpd_ctrl(
    fit_method = "lme",
    tol = 1e-7,
    include_obs = FALSE,
    constraint = "support_at_obs",
    sample_size = sampsize,
    eps_fun = eps_slls,
    thresh_method = "rob_ftr",
    exceed0 = 0.25,
    exceed0_min = 250,
    exceed_min = 10,
    thresh_step = 10,
    gof_test = "ad",
    gof_alpha = 0.05
  )
  
  papprox_ts_constr_1k <- perm_approx(
    obs_stats = waddr_res_ts_1k$`d.wass^2`,
    perm_stats = perm_stats_ts_1k,
    null_center = "mean",
    alternative = "greater",
    method = "gpd",
    approx_thresh = 1,
    gpd_ctrl = gpd_ctrl,
    adjust_method = "BH",
    cores = 10,
    verbose = TRUE
  )
  
  saveRDS(papprox_ts_constr_1k, file = fname_ts_constr1k)
} else {
  papprox_ts_constr_1k <- readRDS(fname_ts_constr1k)
}

cat("Final tuning parameter:")
papprox_ts_constr_1k$fit_result$eps_tune
cat("\n")
summary(papprox_ts_constr_1k)
```


```{r}
fname_ts_constr10k <- path_target("permapprox_results_ts_constr_10k.rds")

if (!file.exists(fname_ts_constr10k)) {
  papprox_ts_constr_10k <- perm_approx(
    obs_stats = waddr_res_ts_10k$`d.wass^2`,
    perm_stats = perm_stats_ts_10k,
    null_center = "mean",
    alternative = "greater",
    method = "gpd",
    approx_thresh = 1,
    gpd_ctrl = gpd_ctrl,
    adjust_method = "BH",
    cores = 10,
    verbose = TRUE
  )
  saveRDS(papprox_ts_constr_10k, file = fname_ts_constr10k)
} else {
  papprox_ts_constr_10k <- readRDS(fname_ts_constr10k)
}

cat("Final tuning parameter:")
papprox_ts_constr_10k$fit_result$eps_tune
cat("\n")
summary(papprox_ts_constr_10k)
```

## Combine results

```{r}
# Collect rows method-by-method (only if the corresponding objects exist)
rows <- list()

## empirical (B = 1k)
rows[["emp_1k"]] <- tibble::tibble(
  idx    = seq_along(papprox_ts_emp_1k$p_unadjusted),
  gene   = waddr_res_ts_1k$gene,
  stat   = waddr_res_ts_1k$`d.wass^2`,
  pval   = papprox_ts_emp_1k$p_unadjusted,
  shape  = NA,
  scale  = NA,
  constraint = NA,
  source = "empirical",
  fit    = "empirical",
  B      = 1000L,
  method_key = "empirical_1000",
  method_label = "Empirical",
  method_label_long = "Empirical (B = 1k)"
)

## empirical (B = 10k)
rows[["emp_10k"]] <- tibble::tibble(
  idx    = seq_along(papprox_ts_emp_10k$p_unadjusted),
  gene   = waddr_res_ts_10k$gene,
  stat   = waddr_res_ts_10k$`d.wass^2`,
  pval   = papprox_ts_emp_10k$p_unadjusted,
  shape  = NA,
  scale  = NA,
  constraint = NA,
  source = "empirical",
  fit    = "empirical",
  B      = 10000L,
  method_key = "empirical_10000",
  method_label = "Empirical",
  method_label_long = "Empirical (B = 10k)"
)

## empirical (B = 50k)
rows[["emp_50k"]] <- tibble::tibble(
  idx    = seq_along(papprox_ts_emp_50k$p_unadjusted),
  gene   = waddr_res_ts_50k$gene,
  stat   = waddr_res_ts_50k$`d.wass^2`,
  pval   = papprox_ts_emp_50k$p_unadjusted,
  shape  = NA,
  scale  = NA,
  constraint = NA,
  source = "empirical",
  fit    = "empirical",
  B      = 50000L,
  method_key = "empirical_50000",
  method_label = "Empirical",
  method_label_long = "Empirical (B = 50k)"
)

## empirical (B = 100k)
rows[["emp_100k"]] <- tibble::tibble(
  idx    = seq_along(papprox_ts_emp_100k$p_unadjusted),
  gene   = waddr_res_ts_100k$gene,
  stat   = waddr_res_ts_100k$`d.wass^2`,
  pval   = papprox_ts_emp_100k$p_unadjusted,
  shape  = NA,
  scale  = NA,
  constraint = NA,
  source = "empirical",
  fit    = "empirical",
  B      = 100000L,
  method_key = "empirical_100000",
  method_label = "Empirical",
  method_label_long = "Empirical (B = 100k)"
)

## waddR (B = 1k)
rows[["waddR_1k"]] <- tibble::tibble(
  idx    = seq_len(nrow(waddr_res_ts_1k)),
  gene   = waddr_res_ts_1k$gene,
  stat   = waddr_res_ts_1k$`d.wass^2`,
  pval   = waddr_res_ts_1k$pval,
  shape  = waddr_res_ts_1k$shape,
  scale  = NA,
  constraint = NA,
  source = "waddR",
  fit    = ifelse(is.na(waddr_res_ts_1k$gpd.shape), "empirical", "GPD"),
  B      = 1000L,
  method_key = "waddR_1000",
  method_label = "waddR",
  method_label_long = "waddR (B = 1k)"
)

## waddR (B = 10k)
rows[["waddR_10k"]] <- tibble::tibble(
  idx    = seq_len(nrow(waddr_res_ts_10k)),
  gene   = waddr_res_ts_10k$gene,
  stat   = waddr_res_ts_10k$`d.wass^2`,
  pval   = waddr_res_ts_10k$pval,
  shape  = waddr_res_ts_10k$shape,
  scale  = NA,
  constraint = NA,
  source = "waddR",
  fit    = ifelse(is.na(waddr_res_ts_10k$gpd.shape), "empirical", "GPD"),
  B      = 10000L,
  method_key = "waddR_10000",
  method_label = "waddR",
  method_label_long = "waddR (B = 10k)"
)


## permApprox unconstrained (B = 1k)
rows[["pA_unconstr_1k"]] <- tibble::tibble(
  idx    = seq_along(papprox_ts_unconstr_1k$p_unadjusted),
  gene   = waddr_res_ts_1k$gene,
  stat   = waddr_res_ts_1k$`d.wass^2`,
  pval   = papprox_ts_unconstr_1k$p_unadjusted,
  shape  = papprox_ts_unconstr_1k$gpd_fit$shape,
  scale  = papprox_ts_unconstr_1k$gpd_fit$scale,
  constraint = "unconstrained",
  source = "permApprox (unconstrained)",
  fit    = ifelse(papprox_ts_unconstr_1k$method_used == "empirical", 
                  "empirical", "GPD"),
  B      = 1000L,
  method_key = "permApprox_unconstrained_1000",
  method_label = "waddR + permApprox(U)",
  method_label_long = "waddR + permApprox(U)\n(B = 1k)"
)


## permApprox unconstrained (B = 10k)
rows[["pA_unconstr_10k"]] <- tibble::tibble(
  idx    = seq_along(papprox_ts_unconstr_10k$p_unadjusted),
  gene   = waddr_res_ts_10k$gene,
  stat   = waddr_res_ts_10k$`d.wass^2`,
  pval   = papprox_ts_unconstr_10k$p_unadjusted,
  shape  = papprox_ts_unconstr_10k$gpd_fit$shape,
  scale  = papprox_ts_unconstr_10k$gpd_fit$scale,
  constraint = "unconstrained",
  source = "permApprox (unconstrained)",
  fit    = ifelse(papprox_ts_unconstr_10k$method_used == "empirical", 
                  "empirical", "GPD"),
  B      = 10000L,
  method_key = "permApprox_unconstrained_10000",
  method_label = "waddR + permApprox(U)",
  method_label_long = "waddR + permApprox(U)\n(B = 10k)"
)


## permApprox constrained (B = 1k)
rows[["pA_constr_1k"]] <- tibble::tibble(
  idx    = seq_along(papprox_ts_constr_1k$p_unadjusted),
  gene   = waddr_res_ts_1k$gene,
  stat   = waddr_res_ts_1k$`d.wass^2`,
  pval   = papprox_ts_constr_1k$p_unadjusted,
  shape  = papprox_ts_constr_1k$gpd_fit$shape,
  scale  = papprox_ts_constr_1k$gpd_fit$scale,
  constraint = "constrained",
  source = "permApprox (constrained)",
  fit    = ifelse(papprox_ts_constr_1k$method_used == "empirical", 
                  "empirical", "GPD"),
  B      = 1000L,
  method_key = "permApprox_constrained_1000",
  method_label = "waddR + permApprox(C)",
  method_label_long = "waddR + permApprox(C)\n(B = 1k)"
)

## permApprox constrained (B = 10k)
rows[["pA_constr_10k"]] <- tibble::tibble(
  idx    = seq_along(papprox_ts_constr_10k$p_unadjusted),
  gene   = waddr_res_ts_10k$gene,
  stat   = waddr_res_ts_10k$`d.wass^2`,
  pval   = papprox_ts_constr_10k$p_unadjusted,
  shape  = papprox_ts_constr_10k$gpd_fit$shape,
  scale  = papprox_ts_constr_10k$gpd_fit$scale,
  constraint = "constrained",
  source = "permApprox (constrained)",
  fit    = ifelse(papprox_ts_constr_10k$method_used == "empirical", 
                  "empirical", "GPD"),
  B      = 10000L,
  method_key = "permApprox_constrained_10000",
  method_label = "waddR + permApprox(C)",
  method_label_long = "waddR + permApprox(C)\n(B = 10k)"
)

# Bind + finalize a global tidy DF
df_pvals_global <- dplyr::bind_rows(rows) %>%
  dplyr::mutate(
    # stable keys & preferred global ordering
    method_key = factor(
      method_key,
      levels = c(
        "empirical_1000", 
        "empirical_10000", 
        "empirical_50000", 
        "empirical_100000",
        "waddR_1000", 
        "waddR_10000",
        "permApprox_unconstrained_1000", 
        "permApprox_unconstrained_10000",
        "permApprox_constrained_1000",
        "permApprox_constrained_10000"
      )
    ),
    # legend grouping
    ptype = ifelse(fit == "empirical", "empirical", "GPD fit")
  )
```

## Plots

```{r}
# returns a clean data.frame (no styling)
sig_table_for_methods_df <- function(df, methods, alphas = c(0.05, 0.01, 0.001),
                                     use_long_labels = TRUE) {
  stopifnot(all(methods %in% df$method_key))
  dd <- dplyr::filter(df, method_key %in% methods)

  lab_col <- if (isTRUE(use_long_labels)) "method_label_long" else "method_label"
  key_labels <- dd |>
    dplyr::group_by(method_key) |>
    dplyr::summarise(method = dplyr::first(.data[[lab_col]]), .groups = "drop")

  base_counts <- dd |>
    dplyr::group_by(method_key) |>
    dplyr::summarise(
      n_tests = sum(!is.na(pval)),
      n_zero  = sum(pval == 0, na.rm = TRUE),
      .groups = "drop"
    )

  alpha_counts <- purrr::map_dfr(alphas, function(a) {
    dd |>
      dplyr::filter(!is.na(pval)) |>
      dplyr::group_by(method_key) |>
      dplyr::summarise(n_sig = sum(pval <= a), .groups = "drop") |>
      dplyr::mutate(alpha = a)
  }) |>
    tidyr::pivot_wider(
      id_cols = method_key,
      names_from = alpha,
      values_from = n_sig,
      names_prefix = "alpha_"
    )

  out <- base_counts |>
    dplyr::left_join(alpha_counts, by = "method_key") |>
    dplyr::left_join(key_labels,   by = "method_key") |>
    dplyr::mutate(method_key = factor(method_key, levels = methods)) |>
    dplyr::arrange(method_key) |>
    dplyr::select(method, n_tests, n_zero, dplyr::starts_with("alpha_"))

  # Ensure numeric columns stay numeric; replace NAs visibly (prevents blank cells)
  out |>
    dplyr::mutate(dplyr::across(where(is.numeric), ~ .x)) |>
    dplyr::mutate(dplyr::across(dplyr::everything(),
                                ~ ifelse(is.na(.), NA, .)))
}

render_sig_table_kable <- function(tbl, caption = NULL) {
  # Optional: relabel alpha columns nicely
  coln <- names(tbl)
  coln <- gsub("^alpha_0.05$", "\u03B1 = 0.05", coln)
  coln <- gsub("^alpha_0.01$", "\u03B1 = 0.01", coln)
  coln <- gsub("^alpha_0.001$", "\u03B1 = 0.001", coln)

  knitr::kable(
    tbl,
    format = "html",            # be explicit
    digits = 0,                 # integers
    align  = c("l", rep("r", ncol(tbl) - 1)),
    format.args = list(scientific = FALSE, trim = TRUE),
    col.names = coln,
    caption   = caption %||% "Significant features and zero p-values"
  ) |>
    kableExtra::kable_styling(full_width = FALSE, position = "left")
}
```

### Simple p-value plots

```{r}
plot_pval_vs_stat <- function(df,
                              title = "",
                              p_col = "pval",
                              stat_col = "stat",
                              alpha = 0.6,
                              point_size = 1.4,
                              add_p05 = TRUE,
                              ybreaks = c(1e-2, 1e-25, 1e-50, 1e-75, 1e-100, 
                                          1e-150, 1e-200),
                              ylabels = c("1e-2", "1e-25", "1e-50", "1e-75", 
                                          "1e-100", "1e-150", "1e-200")) {

  # Count zeros / non-finite for transparency in the caption
  n_all <- nrow(df)
  n_zero <- sum(df[[p_col]] == 0, na.rm = TRUE)
  n_na <- sum(is.na(df[[p_col]]) | is.na(df[[stat_col]]), na.rm = TRUE)

  df_plot <- df %>%
    dplyr::mutate(
      p_for_plot = dplyr::if_else(is.na(.data[[p_col]]) | .data[[p_col]] <= 0,
                                  NA_real_, .data[[p_col]])
    )
  
  ggplot(df_plot, aes(x = .data[[stat_col]], y = p_for_plot)) +
    geom_point(alpha = alpha, size = point_size, na.rm = TRUE) +
    {if (add_p05) geom_hline(yintercept = 0.01, linetype = 2) } +
    scale_y_log10(
      breaks = ybreaks,
      labels = ylabels
    ) +
    labs(
      title = title,
      x = expression(d[W]^2),
      y = "p-value"
    ) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14),
      plot.title = element_text(size = 14),
      plot.subtitle = element_text(size = 10)
    )
}

# Subset: only B=1000 and the three methods you want
df_statplot_1k <- df_pvals_global %>%
  dplyr::filter(
    B == 1000L,
    method_key %in% c("empirical_1000", "waddR_1000", "permApprox_constrained_1000")
  )

p_emp <- plot_pval_vs_stat(
  df_statplot_1k %>% dplyr::filter(method_key == "empirical_1000"),
  title = "Empirical p-values"
)  + coord_cartesian(ylim = c(1, 1e-200))

p_waddr <- plot_pval_vs_stat(
  df_statplot_1k %>% dplyr::filter(method_key == "waddR_1000"),
  title = "waddR p-values "
)  + coord_cartesian(ylim = c(1, 1e-200))

p_constr <- plot_pval_vs_stat(
  df_statplot_1k %>% dplyr::filter(method_key == "permApprox_constrained_1000"),
  title = "permApprox p-values"
)  + coord_cartesian(ylim = c(1, 1e-200))
```

```{r pval_vs_stat_1k_emp_permApprox, fig.width=8, fig.height=4.2, dpi = 200}
(p_emp |  p_constr) +
  patchwork::plot_layout(guides = "collect")
```

```{r pval_vs_stat_1k_emp_waddR_permApprox, fig.width=12, fig.height=4.2, dpi = 200}
(p_emp | p_waddr |  p_constr) +
  patchwork::plot_layout(guides = "collect")
```

```{r pval_vs_stat_1k_emp, fig.width=5, fig.height=4.2, dpi = 200}
p_emp <- plot_pval_vs_stat(
  df_statplot_1k %>% dplyr::filter(method_key == "empirical_1000"),
  title = "Empirical p-values",
  ybreaks = c(1e-2, 1e-5, 1e-10),
  ylabels = c("1e-2", "1e-5", "1e-10")
)  + coord_cartesian(ylim = c(1, 1e-10))

p_emp
```

### P-values (waddR vs. permApprox)

```{r}
plot_pvals_three_zero_colored <- function(df,
                                       methods = c("waddR_1000",
                                                   "permApprox_unconstrained_1000",
                                                   "permApprox_constrained_1000"),
                                       label = c("short", "long"),
                                       legend_position = "bottom",
                                       legend_text_size = 13,
                                       zero_repl = 1e-40,
                                       add_zero  = TRUE,
                                       log_breaks = c(1e-3, 1e-5, 1e-10,
                                                      1e-20, 1e-30, 1e-40),
                                       annotate_genes_c = NULL,
                                       annotate_genes_u = NULL,
                                       annotate_size  = 3,
                                       box_padding_u  = 0.2,
                                       box_padding_c  = 0.2,
                                       point_padding_u = 0.3,
                                       point_padding_c = 0.3,
                                       label_force_u   = 2,
                                       label_force_c   = 2) {

  label <- match.arg(label)

  dd <- df %>%
    dplyr::filter(method_key %in% methods) %>%
    dplyr::mutate(
      method_key = factor(method_key, levels = methods)
    )

  dd <- dd %>%
    dplyr::mutate(
      pval_plot = dplyr::case_when(
        is.na(pval)      ~ NA_real_,
        pval == 0        ~ zero_repl,
        pval < zero_repl ~ NA_real_,
        TRUE             ~ pval
      )
    )

  lab_col <- if (label == "long") "method_label_long" else "method_label"

  zero_tbl <- dd %>%
    dplyr::group_by(method_key) %>%
    dplyr::summarise(
      n_total = sum(!is.na(pval)),
      n_zero  = sum(pval == 0, na.rm = TRUE),
      .groups = "drop"
    )

  base_labs <- dd %>%
    dplyr::group_by(method_key) %>%
    dplyr::summarise(lbl = dplyr::first(.data[[lab_col]]), .groups = "drop")

  lab_map <- setNames(
    paste0(
      base_labs$lbl,
      "\n(zeros: ",
      zero_tbl$n_zero[match(base_labs$method_key, zero_tbl$method_key)],
      " / ",
      zero_tbl$n_total[match(base_labs$method_key, zero_tbl$method_key)],
      ")"
    ),
    as.character(base_labs$method_key)
  )

  log_breaks <- sort(unique(log_breaks))
  if (isTRUE(add_zero)) log_breaks <- sort(unique(c(log_breaks, zero_repl)))
  lower_limit <- min(log_breaks)

  label_fun <- function(breaks) {
    out <- breaks
    out[breaks > 1e-5] <- scales::label_scientific()(breaks[breaks > 1e-5])
    if (isTRUE(add_zero)) {
      out[breaks <= zero_repl * 10] <- "0"
    }
    out
  }

  ## aesthetics for the 3 classes
  color_values <- c(
    "nonzero_waddR"   = "grey40",
    "zero_waddR_only" = "#0072B2",  # blue
    "zero_both"       = "#D55E00", # orange
    "na_waddR"        = "gray10"
  )
  shape_values <- c(
    "nonzero_waddR"   = 1,   # open circle
    "zero_waddR_only" = 17,  # triangle
    "zero_both"       = 16,   # filled circle
    "na_waddR"        = 7  # filled circle
  )
  labels_values <- c(
    "nonzero_waddR"   = "Non-zero in waddR",
    "zero_waddR_only" = "Zero in waddR → >0 in permApprox(U)",
    "zero_both"       = "Zero in waddR → 0 in permApprox(U)",
    "na_waddR"        = "NA in waddR"
  )

  p <- ggplot(dd, aes(x = stat, y = pval_plot)) +
    ## 4 layers so we can control stroke/size
    geom_point(
      data = dd %>% dplyr::filter(zero_class == "nonzero_waddR"),
      aes(color = zero_class, shape = zero_class),
      size = 2.7, alpha = 0.9, stroke = 1
    ) +
    geom_point(
      data = dd %>% dplyr::filter(zero_class == "zero_waddR_only"),
      aes(color = zero_class, shape = zero_class),
      size = 2.7, alpha = 0.9, stroke = 0.4
    ) +
    geom_point(
      data = dd %>% dplyr::filter(zero_class == "zero_both"),
      aes(color = zero_class, shape = zero_class),
      size = 2, alpha = 0.9, stroke = 0.3
    ) +
    geom_point(
      data = dd %>% dplyr::filter(zero_class == "na_waddR"),
      aes(color = zero_class, shape = zero_class),
      size = 2, alpha = 0.9, stroke = 0.6
    ) +
    geom_hline(yintercept = 0.05, linetype = "dashed", color = "black") +
    facet_wrap(~ method_key, nrow = 1, labeller = as_labeller(lab_map)) +
    scale_y_log10(breaks = log_breaks, labels = label_fun) +
    coord_cartesian(ylim = c(lower_limit / 2, 1)) +
    scale_color_manual(
      values = color_values,
      labels = labels_values,
      name   = "zero type"
    ) +
    scale_shape_manual(
      values = shape_values,
      labels = labels_values,
      name   = "zero type"
    ) +
    guides(
      color = guide_legend(override.aes = list(size = 3)),
      shape = guide_legend(override.aes = list(size = 3))
    ) +
    labs(
      x = expression(d[W]^2),
      y = expression(p)
    ) +
    theme_minimal() +
    theme(
      axis.text        = element_text(size = 15),
      axis.title       = element_text(size = 16),
      strip.text       = element_text(size = 15),
      legend.text      = element_text(size = legend_text_size),
      legend.title     = element_blank(),
      legend.position  = legend_position,
      legend.background    = element_rect(fill = "white", colour = "black"),
      legend.key           = element_rect(fill = "white", colour = NA),
      legend.box.background = element_rect(fill = "white", colour = "black"),
      panel.spacing    = unit(2, "lines"),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(color = "gray80")
    )

  ## ---- optional annotation ONLY in permApprox(C) panel ----
  if (!is.null(annotate_genes_c)) {
    ann_df <- dd %>%
      dplyr::filter(
        method_key == "permApprox_constrained_1000",
        gene %in% annotate_genes_c,
        !is.na(pval_plot)
      )

    if (nrow(ann_df) > 0) {
      p <- p +
        ggrepel::geom_text_repel(
          data               = ann_df,
          aes(label = gene),
          size               = annotate_size,
          min.segment.length = 0,
          max.overlaps       = Inf,
          box.padding        = box_padding_c,
          point.padding      = point_padding_c,
          force              = label_force_c,
          force_pull         = 0,
          segment.alpha      = 0.5
        )
    }
  }
  
    if (!is.null(annotate_genes_u)) {
    ann_df <- dd %>%
      dplyr::filter(
        method_key == "permApprox_unconstrained_1000",
        gene %in% annotate_genes_u,
        !is.na(pval_plot)
      )

    if (nrow(ann_df) > 0) {
      p <- p +
        ggrepel::geom_text_repel(
          data               = ann_df,
          aes(label = gene),
          size               = annotate_size,
          min.segment.length = 0,
          max.overlaps       = Inf,
          box.padding        = box_padding_u,
          point.padding      = point_padding_u,
          force              = label_force_u,
          force_pull         = 0,
          segment.alpha      = 0.5
        )
    }
  }

  p
}

```

#### B = 1000

```{r}
zero_class_df_1000 <- df_pvals_global %>%
  dplyr::filter(method_key %in% c("waddR_1000",
                                  "permApprox_unconstrained_1000")) %>%
  dplyr::select(gene, method_key, pval) %>%
  tidyr::pivot_wider(
    id_cols = gene,
    names_from = method_key,
    values_from = pval
  ) %>%
  dplyr::mutate(
    zero_class = dplyr::case_when(
      # NA in waddR
      is.na(`waddR_1000`) ~ "na_waddR",
      
      # non-zero waddR → grey
      `waddR_1000` > 0 ~ "nonzero_waddR",

      # waddR == 0 & permApprox(U) == 0 → red
      `waddR_1000` == 0 & `permApprox_unconstrained_1000` == 0 ~ "zero_both",

      # waddR == 0 but permApprox(U) > 0 → blue
      `waddR_1000` == 0 & (`permApprox_unconstrained_1000` > 0 |
                           is.na(`permApprox_unconstrained_1000`)) ~ "zero_waddR_only"
    )
  )
df_pvals_global_1000 <- df_pvals_global %>%
  dplyr::left_join(zero_class_df_1000, by = "gene")

# 1) create a single zero_class column from the first existing one
df_pvals_global_1000 <- df_pvals_global_1000 %>%
  dplyr::mutate(
    zero_class = dplyr::coalesce(
      !!!rlang::syms(grep("^zero_class", names(df_pvals_global_1000), value = TRUE))
    )
  ) %>%
  # 2) drop all the old suffixed versions
  dplyr::select(-dplyr::matches("^zero_class\\."))

df_pvals_global_1000 <- df_pvals_global_1000 %>%
  dplyr::mutate(
    zero_class = factor(
      zero_class,
      levels = c("nonzero_waddR",     # grey 
                 "zero_waddR_only",   # blue triangles 
                 "zero_both",         # orange
                 "na_waddR")
      
    )
  ) %>%
  arrange(zero_class)
```

```{r pvals_vs_dw2_wad_pau_pac_1k, fig.width=10, fig.height=4.5, dpi=200}
plot_pvals_three_zero_colored(
  df_pvals_global_1000,
  methods = c("waddR_1000",
              "permApprox_unconstrained_1000",
              "permApprox_constrained_1000"),
  label = "short",
  zero_repl = 1e-318,
  log_breaks = c(1e-3, 1e-25, 1e-50, 1e-75, 1e-100, 1e-200, 1e-300),
  add_zero = TRUE,
  legend_position = c(0.2,0.37),
  legend_text_size = 11
)
```

```{r pvals_vs_dw2_wad_pau_pac_1k_ann, fig.width=10, fig.height=4.5, dpi=200}
genes_to_annotate_c <- df_pvals_global_1000 %>% 
  filter(B == 1000, source == "permApprox (constrained)") %>% 
  filter(pval < 1e-140 | stat > 2) %>%
  dplyr::pull(gene)

genes_to_annotate_u <- df_pvals_global_1000 %>% 
  filter(B == 1000, source == "permApprox (unconstrained)") %>% 
  filter(pval < 1e-110, pval > 0) %>%
  dplyr::pull(gene)

plot_pvals_three_zero_colored(
  df_pvals_global_1000,
  methods = c("waddR_1000",
              "permApprox_unconstrained_1000",
              "permApprox_constrained_1000"),
  label = "short",
  zero_repl = 1e-318,
  log_breaks = c(1e-3, 1e-25, 1e-50, 1e-75, 1e-100, 1e-200, 1e-300),
  add_zero = TRUE,
  legend_position = c(0.2,0.37),
  legend_text_size = 11,
  annotate_genes_c = genes_to_annotate_c,
  #annotate_genes_u = genes_to_annotate_u,
  box_padding_c = 0.2,
  box_padding_u = 0.3,
  point_padding_u = 2,
  label_force_u = 4,
  annotate_size = 3
)
```

```{r}
tbl_1k <- sig_table_for_methods_df(
  df_pvals_global_1000,
  methods = c("waddR_1000",
              "permApprox_unconstrained_1000",
              "permApprox_constrained_1000")
)
render_sig_table_kable(tbl_1k, caption = "Counts for B = 1k")
```


#### B = 10000

```{r}
zero_class_df_10000 <- df_pvals_global %>%
  dplyr::filter(method_key %in% c("waddR_10000",
                                  "permApprox_unconstrained_10000")) %>%
  dplyr::select(gene, method_key, pval) %>%
  tidyr::pivot_wider(
    id_cols = gene,
    names_from = method_key,
    values_from = pval
  ) %>%
  dplyr::mutate(
    zero_class = dplyr::case_when(
      # non-zero waddR → grey
      is.na(`waddR_10000`) | `waddR_10000` > 0 ~ "nonzero_waddR",

      # waddR == 0 & permApprox(U) == 0 → red
      `waddR_10000` == 0 & `permApprox_unconstrained_10000` == 0 ~ "zero_both",

      # waddR == 0 but permApprox(U) > 0 → blue
      `waddR_10000` == 0 & (`permApprox_unconstrained_10000` > 0 |
                           is.na(`permApprox_unconstrained_10000`)) ~ "zero_waddR_only"
    )
  )
df_pvals_global_10000 <- df_pvals_global %>%
  dplyr::left_join(zero_class_df_10000, by = "gene")

# 1) create a single zero_class column from the first existing one
df_pvals_global_10000 <- df_pvals_global_10000 %>%
  dplyr::mutate(
    zero_class = dplyr::coalesce(
      !!!rlang::syms(grep("^zero_class", names(df_pvals_global_10000), value = TRUE))
    )
  ) %>%
  # 2) drop all the old suffixed versions
  dplyr::select(-dplyr::matches("^zero_class\\."))

df_pvals_global_10000 <- df_pvals_global_10000 %>%
  dplyr::mutate(
    zero_class = factor(
      zero_class,
      levels = c("nonzero_waddR",     # grey 
                 "zero_waddR_only",   # blue triangles 
                 "zero_both")         # orange
      
    )
  ) %>%
  arrange(zero_class)
```

```{r pvals_vs_dw2_wad_pau_pac_10k, fig.width=10, fig.height=5, dpi=200}
# tmp <- df_pvals_global %>% filter(method_key == "permApprox_unconstrained_10000")
# sum(tmp$pval == 0, na.rm = TRUE)

plot_pvals_three_zero_colored(
  df_pvals_global_10000,
  methods = c("waddR_10000",
              "permApprox_unconstrained_10000",
              "permApprox_constrained_10000"),
  label = "short",
  zero_repl = 1e-120,
  log_breaks = c(1e-3, 1e-25, 1e-50, 1e-75, 1e-100),
  add_zero = TRUE,
  legend_position = c(0.19,0.37),
  legend_text_size = 11
)
```

```{r}
tbl_1k <- sig_table_for_methods_df(
  df_pvals_global,
  methods = c("waddR_10000",
              "permApprox_unconstrained_10000",
              "permApprox_constrained_10000")
)
render_sig_table_kable(tbl_1k, caption = "Counts for B = 10k")
```


### P-values with annotations

```{r}
plot_pvals_single_method <- function(df_long,
                                     method_key,
                                     label        = c("short", "long"),
                                     zero_repl    = 1e-50,
                                     extra_space_factor = 1, # more padding at bottom if > 1
                                     alpha_level  = 0.01,
                                     add_zero     = TRUE,
                                     log_breaks   = c(1e-3, 1e-10, 1e-25, 1e-50),
                                     annotate_taxa   = NULL,
                                     annotate_size   = 3,
                                     label_force     = 2,
                                     box_padding     = 0.2,
                                     legend_position = "bottom") {

  label <- match.arg(label)

  # subset for one method_key (e.g., "waddR_1000", "permApprox_constrained_1000")
  dd <- df_long %>%
    dplyr::filter(method_key == !!method_key)

  if (nrow(dd) == 0L) {
    stop("No rows found for method_key = ", method_key)
  }

  # handle zeros / very small p-values (clamp tiny ones)
  dd <- dd %>%
    dplyr::mutate(
      pval_plot = dplyr::case_when(
        is.na(pval)      ~ NA_real_,
        pval == 0        ~ zero_repl,
        pval < zero_repl ~ zero_repl,
        TRUE             ~ pval
      )
    )

  # choose label column (short vs long) and build title with zero counts
  lab_col <- if (label == "long") "method_label_long" else "method_label"

  base_lab <- dd %>%
    dplyr::summarise(lbl = dplyr::first(.data[[lab_col]]), .groups = "drop") %>%
    dplyr::pull(lbl)

  zero_info <- dd %>%
    dplyr::summarise(
      n_total = sum(!is.na(pval)),
      n_zero  = sum(pval == 0, na.rm = TRUE),
      .groups = "drop"
    )

  title_text <- paste0(
    base_lab,
    "\n(zeros: ", zero_info$n_zero, " / ", zero_info$n_total, ")"
  )

  # y-axis breaks and labels
  log_breaks <- sort(unique(log_breaks))
  if (isTRUE(add_zero)) log_breaks <- sort(unique(c(log_breaks, zero_repl)))

  lower_limit       <- min(log_breaks)
  lower_plot_limit  <- lower_limit / extra_space_factor

  label_fun <- function(breaks) {
    out <- breaks
    out[breaks > 1e-5] <- scales::label_scientific()(breaks[breaks > 1e-5])
    if (isTRUE(add_zero)) {
      out[breaks <= zero_repl * 10] <- "0"
    }
    out
  }

  p <- ggplot(dd, aes(x = stat, y = pval_plot)) +
    geom_point(aes(fill = ptype),
               shape = 21, size = 2.5, alpha = 0.7,
               color = "gray10", stroke = 0.2) +
    geom_hline(yintercept = alpha_level, linetype = "dashed", color = "black") +
    scale_y_log10(breaks = log_breaks, labels = label_fun) +
    coord_cartesian(ylim = c(lower_plot_limit, 1)) +
    scale_fill_manual(
      values = c("GPD fit" = "#F8766D", "empirical" = "gray"),
      name   = "p-value type"
    ) +
    labs(
      title = title_text,
      x     = expression(d[W]^2),
      y     = expression(p)
    ) +
    theme_minimal() +
    theme(
      axis.text            = element_text(size = 14),
      axis.title           = element_text(size = 14),
      plot.title           = element_text(size = 14, hjust = 0.5),
      legend.title         = element_text(size = 14),
      legend.text          = element_text(size = 14),
      legend.background    = element_rect(fill = "white", colour = "black"),
      legend.key           = element_rect(fill = "white", colour = NA),
      legend.box.background = element_rect(fill = "white", colour = "black"),
      legend.position      = legend_position,
      panel.grid.minor     = element_blank(),
      panel.grid.major.y   = element_line(color = "gray80")
    )

  # optional annotation
  if (!is.null(annotate_taxa)) {
    ann_df <- dd %>%
      dplyr::filter(gene %in% annotate_taxa, !is.na(pval_plot))

    if (nrow(ann_df) > 0) {
      p <- p +
        ggrepel::geom_text_repel(
          data               = ann_df,
          aes(label = gene),
          size               = annotate_size,
          min.segment.length = 0,
          max.overlaps       = Inf,
          box.padding        = box_padding,
          point.padding      = 0.3,
          force              = label_force,
          force_pull         = 0,
          segment.alpha      = 0.5
        )
    }
  }

  p
}
```

```{r}
## Restrict to B = 1000 and the three methods of interest
df_ts_long <- df_pvals_global %>%
  dplyr::filter(method_key %in% c("waddR_1000",
                                  "permApprox_unconstrained_1000",
                                  "permApprox_constrained_1000"))

## For the paper, you want a cleaner label for the constrained permApprox
df_ts_long <- df_ts_long %>%
  dplyr::mutate(
    method_label = dplyr::if_else(
      method_key == "permApprox_constrained_1000",
      "waddR + permApprox",          # short label
      method_label
    ),
    method_label_long = dplyr::if_else(
      method_key == "permApprox_constrained_1000",
      "waddR + permApprox (constrained)",  # long label if you ever use it
      method_label_long
    )
  )

## Annotation sets:
## - genes with zero waddR p-values
## - top few genes by smallest waddR p-values
genes_to_annotate_perma <- df_pvals_global %>% 
  filter(B == 1000, source == "permApprox (constrained)") %>% 
  arrange(pval) %>% 
  dplyr::slice(1:7) %>%
  dplyr::pull(gene)

## Common y-axis settings for the plots
log_breaks <- c(1e-3, 1e-25, 1e-50, 1e-75, 1e-100, 1e-200, 1e-300)
zero_repl  <- 1e-319
extra_space_factor <- 10000
```

```{r, fig.width=5, fig.height=4.5, dpi=200}
p_waddR <- plot_pvals_single_method(
  df_ts_long,
  method_key        = "waddR_1000",
  label             = "short",
  zero_repl         = zero_repl,
  log_breaks        = log_breaks,
  box_padding       = 0.5,
  label_force       = 1.7,
  extra_space_factor = extra_space_factor,
  alpha_level       = 0.01,
  legend_position   = c(0.8, 0.4)
)

p_pApp_u <- plot_pvals_single_method(
  df_ts_long,
  method_key        = "permApprox_unconstrained_1000",
  label             = "short",
  zero_repl         = zero_repl,
  log_breaks        = log_breaks,
  box_padding       = 0.5,
  label_force       = 1.7,
  extra_space_factor = extra_space_factor,
  alpha_level       = 0.01
)

p_pApp <- plot_pvals_single_method(
  df_ts_long,
  method_key        = "permApprox_constrained_1000",
  label             = "short",
  zero_repl         = zero_repl,
  log_breaks        = log_breaks,
  annotate_taxa     = genes_to_annotate_perma,
  annotate_size     = 3,
  box_padding       = 0.3,
  label_force       = 1.8,
  extra_space_factor = extra_space_factor,
  alpha_level       = 0.01
)

```

```{r pvals_vs_dw2_ts_wad_pac_1k, fig.width=12, fig.height=4.5, dpi=200}
p_waddR +
  p_pApp_u +
  theme(legend.position = "none") +
  p_pApp +
  theme(legend.position = "none") +
  patchwork::plot_layout(nrow = 1, guides = "keep") &
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```


### P-values vs. ranked tests

```{r}
# Ranked p-value plot with empirical_100000 as fixed reference (open black circles)
# Always sort by decreasing p_ref; within the p_ref floor (<= 1/(B+1)) sort by increasing stat.
plot_pvals_ranked <- function(df,
                              methods = c("waddR_1000",
                                          "permApprox_unconstrained_1000",
                                          "permApprox_constrained_1000"),
                              label      = c("short", "long"),
                              pval_min   = NULL,
                              n_top      = NULL,   # keep first n after ordering
                              n_bottom   = NULL,   # keep last n after ordering
                              zero_repl  = 1e-320,
                              add_zero   = TRUE,
                              emp_to_na  = TRUE,
                              log_breaks = c(1e-3, 1e-10, 1e-25, 1e-50,
                                             1e-75, 1e-100, 1e-200, 1e-300),
                              add_grid   = TRUE,
                              ref_key    = "empirical_100000",
                              B_ref      = 100000,
                              floor_tol  = 1e-12,      # small tolerance around the p-floor
                              drop_na_stat = c("reference", "any", "none"),
                              ref_size   = 2.2,   # size of reference dots
                              base_size  = 1.8, # start size for other methods
                              method_colors = NULL,
                              show_ref_in_legend = TRUE) {
  
  label <- match.arg(label)
  drop_na_stat <- match.arg(drop_na_stat)

  if (!is.null(n_top) && !is.null(n_bottom)) 
    stop("Specify at most one of `n_top` or `n_bottom`.")
  if (!ref_key %in% df$method_key) 
    stop("Reference method_key '", ref_key, "' not found in df$method_key.")

  # ensure ref isn't duplicated in methods
  methods  <- unique(setdiff(methods, ref_key))
  sel_keys <- c(ref_key, methods)

  # subset to ref + methods
  dd <- df %>%
    dplyr::filter(method_key %in% sel_keys) %>%
    dplyr::mutate(method_key = factor(method_key, levels = sel_keys))

  # Drop NA stats early (prevents missing ranks at start)
  if (drop_na_stat == "reference") {
    bad_genes <- dd %>% dplyr::filter(method_key == ref_key, 
                                      is.na(stat) | !is.finite(stat)) %>%
      dplyr::pull(gene) %>% unique()
    dd <- dplyr::filter(dd, !gene %in% bad_genes)
  } else if (drop_na_stat == "any") {
    dd <- dd %>%
      dplyr::group_by(gene) %>%
      dplyr::filter(!any(is.na(stat) | !is.finite(stat))) %>%
      dplyr::ungroup()
  }

  # keep only genes present in ALL selected series
  dd <- dd %>%
    dplyr::group_by(gene) %>%
    dplyr::filter(dplyr::n_distinct(method_key) == length(sel_keys)) %>%
    dplyr::ungroup()

  # optional p-value floor across all selected series
  if (!is.null(pval_min)) {
    stopifnot(is.numeric(pval_min), length(pval_min) == 1, !is.na(pval_min))
    keep_genes <- dd %>%
      dplyr::group_by(gene) %>%
      dplyr::summarise(n_ok = sum(!is.na(pval) & pval > pval_min), 
                       .groups = "drop") %>%
      dplyr::filter(n_ok == length(sel_keys)) %>%
      dplyr::pull(gene)
    dd <- dplyr::filter(dd, gene %in% keep_genes)
  }

  if (nrow(dd) == 0L) {
    return(ggplot() +
             annotate("text", .5, .5, 
                      label = "No points after filtering.", size = 5) +
             theme_void())
  }

  lab_col <- if (label == "long") "method_label_long" else "method_label"

  # --- Reference order --------------------------------------------------------
  p_floor <- 1 / (B_ref + 1)  # min empirical p for B permutations with +1 correction

  ref_order <- dd %>%
    dplyr::filter(method_key == ref_key) %>%
    dplyr::transmute(
      gene,
      p_ref = pval,
      stat_ref = stat,
      # treat 0s as floor for ordering; keep NA last
      p_sort = dplyr::case_when(
        is.na(p_ref) ~ -Inf,     # NA will go to the very end (after slice it likely drops)
        p_ref == 0   ~ zero_repl,
        TRUE         ~ p_ref
      ),
      is_floor = !is.na(p_ref) & (p_ref <= p_floor * (1 + floor_tol))
    ) %>%
    # Decreasing by p, then for floor-tied ones sort by increasing stat
    dplyr::arrange(dplyr::desc(p_sort),
                   dplyr::if_else(is_floor, stat_ref, Inf)) %>%
    dplyr::select(gene)

  # slice top/bottom on the reference order
  if (!is.null(n_top))     ref_order <- ref_order %>% dplyr::slice_head(n = n_top)
  if (!is.null(n_bottom))  ref_order <- ref_order %>% dplyr::slice_tail(n = n_bottom)

  ref_order <- ref_order %>% dplyr::mutate(rank = dplyr::row_number())
  max_rank  <- nrow(ref_order)

  # x axis: start at 1, pad ±0.5 so rank 1 dot isn't cut; force ticks at 1 and max
  x_limits <- c(0.5, max_rank + 0.5)
  x_breaks <- sort(unique(c(1, scales::pretty_breaks(n = 6)(c(1, max_rank)), max_rank)))
  
  #join single rank to all methods
  dd <- dd %>%
  dplyr::semi_join(ref_order, by = "gene") %>%
  dplyr::left_join(ref_order, by = "gene") %>%
  dplyr::mutate(
    legend_label = .data[[lab_col]],
    # force NA if method is not "empirical" but ptype says "empirical"
    pval = dplyr::case_when(
      isTRUE(emp_to_na) & !grepl("empirical", method_key) & ptype == "empirical" ~ NA_real_,
      TRUE ~ pval
    ),
    pval_plot = dplyr::case_when(
      is.na(pval)       ~ NA_real_,
      pval == 0         ~ zero_repl,
      pval < zero_repl  ~ NA_real_,
      TRUE              ~ pval
    )
  ) %>%
  dplyr::arrange(rank)
  
  # split reference vs comparison
  ref_pts <- dd %>% dplyr::filter(method_key == ref_key)
  cmp_pts <- dd %>% dplyr::filter(method_key != ref_key)

  # --- y-axis breaks & labels -------------------------------------------------
  log_breaks <- sort(unique(log_breaks))
  if (isTRUE(add_zero)) log_breaks <- sort(unique(c(log_breaks, zero_repl)))
  lower_limit <- min(log_breaks)
  label_fun <- function(breaks) {
    out <- breaks
    out[breaks > 1e-5] <- scales::label_scientific()(breaks[breaks > 1e-5])
    if (isTRUE(add_zero)) out[breaks <= zero_repl * 10] <- "0"
    out
  }

  # vertical guide per rank (== per gene)
  grid_df <- ref_order %>% dplyr::transmute(rank, ymin = lower_limit, ymax = 1)

  x_title <- bquote("Test index (ordered by decreasing " ~ p[emp_100k] ~ ")")

  # --- Legend mapping: color by method_key, labels from short/long ------------
  legend_df_all <- dd %>%
    dplyr::distinct(method_key, legend_label) %>%
    dplyr::mutate(method_key = as.character(method_key))

  if (isTRUE(show_ref_in_legend)) {
    ref_lab <- legend_df_all$legend_label[legend_df_all$method_key == ref_key][1]
    legend_keys <- c(ref_key, as.character(methods))                 # ref first
    legend_labs <- c(ref_lab, legend_df_all$legend_label[match(methods, legend_df_all$method_key)])
  } else {
    legend_keys <- as.character(methods)
    legend_labs <- legend_df_all$legend_label[match(methods, legend_df_all$method_key)]
  }
  
  # --- Plot -------------------------------------------------------------------
  p <- ggplot()
  if (isTRUE(add_grid)) {
    p <- p + geom_segment(data = grid_df,
                          aes(x = rank, xend = rank, y = ymin, yend = ymax),
                          linewidth = 0.2, color = "grey85")
  }
  p <- p + geom_point(
    data = ref_pts,
    aes(x = rank, y = pval_plot, color = ref_key),
    shape = 21, size = ref_size, stroke = 0.5,
    fill = NA,
    show.legend = show_ref_in_legend
  ) 
  
  # comparison methods (colored)
  # add one layer per method in the *given* order -> controls z-order
  for (i in seq_along(methods)) {
    mk <- methods[i]
    this_size <- base_size - 0.2 * (i - 1)
    p <- p + geom_point(
      data = cmp_pts %>% dplyr::filter(method_key == mk),
      aes(x = rank, y = pval_plot, color = method_key),
      alpha = 0.8, size = this_size, inherit.aes = FALSE
    )
  }
  
  # reference as open black circles (optionally in legend)
  p <- p + 
    scale_x_continuous(limits = x_limits, breaks = x_breaks, 
                       expand = expansion(mult = c(0, 0))) +
    scale_y_log10(breaks = log_breaks, labels = label_fun) +
    coord_cartesian(ylim = c(lower_limit, 1)) +
    labs(x = x_title, y = expression(p), color = "Method") +
    theme_minimal() +
    theme(
      axis.text   = element_text(size = 12),
      axis.title  = element_text(size = 14),
      legend.title= element_text(size = 12),
      legend.text = element_text(size = 10),
      panel.grid.minor = element_blank()
    )
  
  # Colors in legend order
  if (!is.null(method_colors)) {
    if (is.null(names(method_colors))) names(method_colors) <- legend_keys
    if (isTRUE(show_ref_in_legend) && is.na(method_colors[ref_key])) {
      method_colors[ref_key] <- "black"
    }
    col_vals <- method_colors[legend_keys]
  } else {
    if (isTRUE(show_ref_in_legend)) {
      col_vals <- setNames(c("black", scales::hue_pal()(length(legend_keys) - 1)), legend_keys)
    } else {
      col_vals <- setNames(scales::hue_pal()(length(legend_keys)), legend_keys)
    }
  }

  p <- p + scale_color_manual(values = col_vals,
                              breaks = legend_keys,
                              labels = legend_labs,
                              name   = "Method")

  if (isTRUE(show_ref_in_legend)) {
    p <- p + guides(color = guide_legend(
      override.aes = list(
        shape = c(21, rep(16, length(legend_keys) - 1)),
        fill  = c(NA,  rep(NA,  length(legend_keys) - 1))
      )
    ))
  } else {
    p <- p + guides(color = guide_legend(override.aes = list(shape = 16)))
  }

  p
}

```

```{r, fig.width=10, fig.height=4, dpi=200}
my_cols <- c(
  "empirical_1000"                 = "gray50",
  "waddR_1000"                     = "skyblue3",  
  "waddR_10000"                    = "royalblue4",  
  "permApprox_unconstrained_1000"  = "orange2",
  "permApprox_unconstrained_10000" = "chocolate4",
  "permApprox_constrained_1000"  = "#F8766D",  
  "permApprox_constrained_10000"   = "red4"
)
```

#### 1,000 vs. 10,000 permutations

```{r pvals_vs_idx_emp_wad1k_wad10k_all, fig.width=10, fig.height=4, dpi=200}
plot_pvals_ranked(
  df_pvals_global,
  methods = c("empirical_1000",
              "waddR_10000",
              "waddR_1000"
  ),
  label = "long",
  emp_to_na = TRUE,
  add_grid = FALSE,
  pval_min = NULL,
  log_breaks = c(1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5, 1e-10, 1e-15, 1e-20),
  add_zero = FALSE,
  n_top = 1000,
  method_colors = my_cols,
  show_ref_in_legend = TRUE
)
```

Since the Empirical p-values with B = 100k are constant from approx index 130 on, we plot only up to this test in the following.

```{r pvals_vs_idx_emp_wad1k_wad10k, fig.width=10, fig.height=4, dpi=200}
plot_pvals_ranked(
  df_pvals_global,
  methods = c("empirical_1000",
              "waddR_10000",
              "waddR_1000"
  ),
  label = "long",
  emp_to_na = TRUE,
  pval_min = 1e-15,
  log_breaks = c(1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5, 1e-6, 1e-7, 1e-8, 1e-9, 1e-10),
  add_zero = FALSE,
  n_top = 130,
  method_colors = my_cols,
  show_ref_in_legend = TRUE
)
```

```{r pvals_vs_idx_emp_pau1k_pau10k_all, fig.width=10, fig.height=4, dpi=200}
plot_pvals_ranked(
  df_pvals_global,
  methods = c("empirical_1000",
              "permApprox_unconstrained_10000",
              "permApprox_unconstrained_1000"),
  label = "long",
  emp_to_na = TRUE,
  add_grid = FALSE,
  log_breaks = c(1e-0, 1e-5,1e-10, 1e-25, 1e-50, 1e-75, 1e-100),
  add_zero = FALSE,
  n_top = 1000,
  method_colors = my_cols,
  show_ref_in_legend = TRUE
)
```


```{r pvals_vs_idx_emp_pau1k_pau10k, fig.width=10, fig.height=4, dpi=200}
plot_pvals_ranked(
  df_pvals_global,
  methods = c("empirical_1000",
              "permApprox_unconstrained_10000",
              "permApprox_unconstrained_1000"),
  label = "long",
  emp_to_na = TRUE,
  pval_min = 1e-15,
  log_breaks = c(1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5, 1e-6, 1e-7, 1e-8, 1e-9, 1e-10),
  add_zero = FALSE,
  n_top = 130,
  method_colors = my_cols,
  show_ref_in_legend = TRUE
)
```

```{r pvals_vs_idx_emp_pac1k_pac10k_all, fig.width=10, fig.height=4, dpi=200}
plot_pvals_ranked(
  df_pvals_global,
  methods = c("empirical_1000",
              "permApprox_constrained_10000",
              "permApprox_constrained_1000"),
  label = "long",
  emp_to_na = TRUE,
  add_grid = FALSE,
  log_breaks = c(1e-0, 1e-5,1e-10, 1e-25, 1e-50, 1e-75, 1e-100),
  add_zero = FALSE,
  n_top = 1000,
  method_colors = my_cols,
  show_ref_in_legend = TRUE
)
```

```{r pvals_vs_idx_emp_pac1k_pac10k, fig.width=10, fig.height=4, dpi=200}
plot_pvals_ranked(
  df_pvals_global,
  methods = c("empirical_1000",
              "permApprox_constrained_10000",
              "permApprox_constrained_1000"),
  label = "long",
  emp_to_na = TRUE,
  pval_min = 1e-15,
  log_breaks = c(1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5, 1e-6, 1e-7, 1e-8, 1e-9, 1e-10),
  add_zero = FALSE,
  n_top = 130,
  method_colors = my_cols,
  show_ref_in_legend = TRUE
)
```

#### waddR vs. permApprox (B = 1,000)

```{r pvals_vs_idx_wad_pau_pac_1k, fig.width=10, fig.height=4, dpi=200}
plot_pvals_ranked(
  df_pvals_global,
  methods = c("waddR_1000",
              "permApprox_unconstrained_1000",
              "permApprox_constrained_1000"),
  label = "long",
  emp_to_na = TRUE,
  pval_min = 1e-15,
  log_breaks = c(1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5, 1e-6, 1e-7, 1e-8, 1e-9, 1e-10),
  add_zero = FALSE,
  n_top = 130,
  method_colors = my_cols,
  show_ref_in_legend = TRUE
)
```

#### waddR vs. permApprox (B = 10,000)

```{r pvals_vs_idx_wad_pau_pac_10k, fig.width=10, fig.height=4, dpi=200}
plot_pvals_ranked(
  df_pvals_global,
  methods = c("waddR_10000",
              "permApprox_unconstrained_10000",
              "permApprox_constrained_10000"),
  label = "long",
  emp_to_na = TRUE,
  pval_min = 1e-15,
  log_breaks = c(1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5, 1e-6, 1e-7, 1e-8, 1e-9, 1e-10),
  add_zero = FALSE,
  n_top = 130,
  method_colors = my_cols,
  show_ref_in_legend = TRUE
)
```

### Ratio of p-values (vs. empirical with B = 100k)

```{r}
# Ratio plot: p_method / p_ref along ranked test index (ref = empirical_100000)
plot_pval_ratio_ranked <- function(df,
                                   methods = c("waddR_1000",
                                               "permApprox_unconstrained_1000",
                                               "permApprox_constrained_1000"),
                                   label        = c("short","long"),
                                   pval_min     = NULL,
                                   n_top        = NULL,
                                   n_bottom     = NULL,
                                   zero_repl    = 1e-320,   # floor for p-values before ratio
                                   ratio_floor  = 1e-300,   # clamp tiny ratios for log scale safety
                                   ref_key      = "empirical_100000",
                                   B_ref        = 100000,
                                   floor_tol    = 1e-12,
                                   drop_na_stat = c("reference","any","none"),
                                   emp_to_na    = TRUE,
                                   base_size    = 1.8,      # first method size; then -0.2 per method
                                   ratio_breaks = c(1e-6, 1e-4, 1e-2, 1, 1e2, 1e4, 1e6),
                                   method_colors = NULL) {

  label        <- match.arg(label)
  drop_na_stat <- match.arg(drop_na_stat)

  if (!is.null(n_top) && !is.null(n_bottom)) 
    stop("Specify at most one of `n_top` or `n_bottom`.")
  if (!ref_key %in% df$method_key) 
    stop("Reference method_key '", ref_key, "' not found in df$method_key.")

  methods  <- unique(setdiff(methods, ref_key))
  sel_keys <- c(ref_key, methods)

  dd <- df %>%
    dplyr::filter(method_key %in% sel_keys) %>%
    dplyr::mutate(method_key = factor(method_key, levels = sel_keys))

  # Drop NA stats early
  if (drop_na_stat == "reference") {
    bad_genes <- dd %>%
      dplyr::filter(method_key == ref_key, is.na(stat) | !is.finite(stat)) %>%
      dplyr::pull(gene) %>% unique()
    dd <- dplyr::filter(dd, !gene %in% bad_genes)
  } else if (drop_na_stat == "any") {
    dd <- dd %>% dplyr::group_by(gene) %>%
      dplyr::filter(!any(is.na(stat) | !is.finite(stat))) %>%
      dplyr::ungroup()
  }

  # keep only genes present in ALL selected series
  dd <- dd %>%
    dplyr::group_by(gene) %>%
    dplyr::filter(dplyr::n_distinct(method_key) == length(sel_keys)) %>%
    dplyr::ungroup()

  # optional p-value floor across all series
  if (!is.null(pval_min)) {
    keep_genes <- dd %>% dplyr::group_by(gene) %>%
      dplyr::summarise(n_ok = sum(!is.na(pval) & pval > pval_min), 
                       .groups = "drop") %>%
      dplyr::filter(n_ok == length(sel_keys)) %>% dplyr::pull(gene)
    dd <- dplyr::filter(dd, gene %in% keep_genes)
  }

  if (nrow(dd) == 0L) 
    return(ggplot() + annotate("text", .5, .5, 
                               label = "No points after filtering.", size = 5) + 
             theme_void())

  lab_col <- if (label == "long") "method_label_long" else "method_label"

  # --- Reference order: decreasing p_ref; ties at floor broken by increasing stat
  p_floor <- 1 / (B_ref + 1)

  ref_order <- dd %>%
    dplyr::filter(method_key == ref_key) %>%
    dplyr::transmute(
      gene,
      p_ref = pval,
      stat_ref = stat,
      p_sort = dplyr::case_when(
        is.na(p_ref) ~ -Inf,
        p_ref == 0   ~ zero_repl,
        TRUE         ~ p_ref
      ),
      is_floor = !is.na(p_ref) & (p_ref <= p_floor * (1 + floor_tol))
    ) %>%
    dplyr::arrange(dplyr::desc(p_sort),
                   dplyr::if_else(is_floor, stat_ref, Inf)) %>%
    dplyr::select(gene)

  if (!is.null(n_top))     ref_order <- ref_order %>% dplyr::slice_head(n = n_top)
  if (!is.null(n_bottom))  ref_order <- ref_order %>% dplyr::slice_tail(n = n_bottom)

  ref_order <- ref_order %>% dplyr::mutate(rank = dplyr::row_number())
  max_rank  <- nrow(ref_order)

  # x axis: start at 1, pad ±0.5 so rank 1 dot isn't cut; force ticks at 1 and max
  x_limits <- c(0.5, max_rank + 0.5)
  x_breaks <- sort(unique(c(1, scales::pretty_breaks(n = 6)(c(1, max_rank)), 
                            max_rank)))

  # join, handle emp_to_na, compute ratios
  dd <- dd %>%
    dplyr::semi_join(ref_order, by = "gene") %>%
    dplyr::left_join(ref_order, by = "gene") %>%
    dplyr::mutate(
      legend_label = .data[[lab_col]],
      # drop empirical placeholder p-values for non-empirical methods if requested
      pval = dplyr::case_when(
        isTRUE(emp_to_na) & !grepl("empirical", method_key) & ptype == "empirical" ~ NA_real_,
        TRUE ~ pval
      ),
      p_ref    = pmax(ifelse(method_key == ref_key, pval, NA_real_), zero_repl),
      p_method = pmax(ifelse(method_key != ref_key, pval, NA_real_), zero_repl)
    ) %>%
    dplyr::group_by(gene) %>%
    dplyr::mutate(p_ref_gene = max(p_ref, na.rm = TRUE)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      ratio_raw = ifelse(method_key == ref_key, NA_real_, p_method / p_ref_gene),
      ratio_plot = dplyr::case_when(
        is.na(ratio_raw)        ~ NA_real_,
        ratio_raw <= 0          ~ NA_real_,
        ratio_raw < ratio_floor ~ ratio_floor,
        TRUE                    ~ ratio_raw
      )
    ) %>%
    dplyr::arrange(rank)

  cmp_pts <- dd %>% dplyr::filter(method_key != ref_key)

  # vertical guides
  grid_df <- ref_order %>% dplyr::transmute(rank, ymin = min(ratio_breaks), 
                                            ymax = max(ratio_breaks))

  x_title <- bquote("Test index (ordered by decreasing " ~ p[emp_100k] ~ ")")

  # Legend keys/labels
  legend_df_all <- dd %>%
    dplyr::distinct(method_key, legend_label) %>%
    dplyr::mutate(method_key = as.character(method_key))
  legend_keys <- as.character(methods)
  legend_labs <- legend_df_all$legend_label[match(methods, 
                                                  legend_df_all$method_key)]

  # --- Plot
  p <- ggplot() +
    geom_segment(data = grid_df,
                 aes(x = rank, xend = rank, y = ymin, yend = ymax),
                 linewidth = 0.2, color = "grey85") +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40")

  # comparison methods (colored), one layer per method in the given order
  for (i in seq_along(methods)) {
    mk <- methods[i]
    this_size <- base_size - 0.2 * (i - 1)
    p <- p + geom_point(
      data = cmp_pts %>% dplyr::filter(method_key == mk),
      aes(x = rank, y = ratio_plot, color = method_key),
      alpha = 0.85, size = this_size, inherit.aes = FALSE
    )
  }

  p <- p +
    scale_x_continuous(limits = x_limits, breaks = x_breaks, 
                       expand = expansion(mult = c(0, 0))) +
    scale_y_log10(breaks = ratio_breaks, labels = scales::label_scientific()) +
    labs(x = x_title, y = expression(p[method]/p[emp_100k]), 
         color = "Method") +
    theme_minimal() +
    theme(
      axis.text   = element_text(size = 12),
      axis.title  = element_text(size = 14),
      legend.title= element_text(size = 12),
      legend.text = element_text(size = 10),
      panel.grid.minor = element_blank()
    )

  # Colors in legend order
  if (!is.null(method_colors)) {
    if (is.null(names(method_colors))) names(method_colors) <- legend_keys
    col_vals <- method_colors[legend_keys]
  } else {
    col_vals <- setNames(scales::hue_pal()(length(legend_keys)), legend_keys)
  }

  p <- p + scale_color_manual(values = col_vals,
                              breaks = legend_keys,
                              labels = legend_labs,
                              name   = "Method") +
    guides(color = guide_legend(override.aes = list(shape = 16)))

  p
}
```


#### 1,000 vs. 10,000 permutations

```{r ratios_vs_idx_emp_wad1k_wad10k, fig.width=10, fig.height=4, dpi=200}
plot_pval_ratio_ranked(
  df_pvals_global,
  methods = c("empirical_1000",
              "waddR_10000",
              "waddR_1000"
  ),
  label = "long",
  emp_to_na = TRUE,
  pval_min = 1e-10,
  ratio_breaks = c(1e+2, 1e+1, 1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5),
  n_top = 130,
  method_colors = my_cols
)
```

```{r ratios_vs_idx_emp_pau1k_pau10k, fig.width=10, fig.height=4, dpi=200}
plot_pval_ratio_ranked(
  df_pvals_global,
  methods = c("empirical_1000",
              "permApprox_unconstrained_10000",
              "permApprox_unconstrained_1000"),
  label = "long",
  emp_to_na = TRUE,
  pval_min = 1e-10,
  ratio_breaks = c(1e+2, 1e+1, 1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5),
  n_top = 130,
  method_colors = my_cols
)
```


```{r ratios_vs_idx_emp_pac1k_pac10k, fig.width=10, fig.height=4, dpi=200}
plot_pval_ratio_ranked(
  df_pvals_global,
  methods = c("empirical_1000",
              "permApprox_constrained_10000",
              "permApprox_constrained_1000"),
  label = "long",
  emp_to_na = TRUE,
  pval_min = 1e-10,
  ratio_breaks = c(1e+2, 1e+1, 1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5),
  n_top = 130,
  method_colors = my_cols
)
```

#### waddR vs. permApprox (B = 1,000)

```{r ratios_vs_idx_wad_pau_pac_1k, fig.width=10, fig.height=4, dpi=200}
plot_pval_ratio_ranked(
  df_pvals_global,
  methods = c("waddR_1000",
              "permApprox_unconstrained_1000",
              "permApprox_constrained_1000"),
  label = "long",
  emp_to_na = TRUE,
  pval_min = 1e-10,
  ratio_breaks = c(1e+2, 1e+1, 1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5),
  n_top = 130,
  method_colors = my_cols
)
```

#### waddR vs. permApprox (B = 10,000)

```{r ratios_vs_idx_wad_pau_pac_10k, fig.width=10, fig.height=4, dpi=200}
plot_pval_ratio_ranked(
  df_pvals_global,
  methods = c("waddR_10000",
              "permApprox_unconstrained_10000",
              "permApprox_constrained_10000"),
  label = "long",
  emp_to_na = TRUE,
  pval_min = 1e-10,
  ratio_breaks = c(1e+2, 1e+1, 1e-0, 1e-1, 1e-2, 1e-3, 1e-4, 1e-5),
  n_top = 130,
  method_colors = my_cols
)
```

## P-values vs. ordered test statistics

```{r}
plot_p_vs_stat_ordered <- function(df,
                                   methods = c("waddR_1000",
                                               "permApprox_unconstrained_1000",
                                               "permApprox_constrained_1000"),
                                   label      = c("short","long"),
                                   zero_repl  = 1e-320,
                                   add_zero   = TRUE,
                                   log_breaks = c(1e-3, 1e-25, 1e-50, 1e-100, 1e-200, 1e-300),
                                   add_path   = TRUE,   # connect points in stat order
                                   add_bins   = TRUE,   # binned median guide over x
                                   bins       = 20,
                                   add_loess  = FALSE   # optional smooth
) {
  label <- match.arg(label)
  lab_col <- if (label == "long") "method_label_long" else "method_label"

  # keep only selected methods; ensure all selected genes are in all methods
  dd <- df %>%
    dplyr::filter(method_key %in% methods) %>%
    dplyr::mutate(method_key = factor(method_key, levels = methods)) %>%
    dplyr::group_by(gene) %>%
    dplyr::filter(dplyr::n_distinct(method_key) == length(methods)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      pval_plot = dplyr::case_when(
        is.na(pval)      ~ NA_real_,
        pval == 0        ~ zero_repl,
        pval < zero_repl ~ NA_real_,
        TRUE             ~ pval
      )
    ) %>%
    dplyr::group_by(method_key) %>%
    dplyr::arrange(stat, .by_group = TRUE) %>%          # <- ORDER BY STAT (increasing)
    dplyr::mutate(ord_idx = dplyr::row_number()) %>%
    dplyr::ungroup()

  # facet labels
  labs_tbl <- dd %>% dplyr::group_by(method_key) %>%
    dplyr::summarise(lbl = dplyr::first(.data[[lab_col]]), .groups = "drop")
  lab_map <- setNames(labs_tbl$lbl, labs_tbl$method_key)

  # y-axis formatting
  log_breaks <- sort(unique(log_breaks))
  if (isTRUE(add_zero)) log_breaks <- sort(unique(c(log_breaks, zero_repl)))
  lower_limit <- min(log_breaks)
  label_fun <- function(b) {
    out <- b
    out[b > 1e-5] <- scales::label_scientific()(b[b > 1e-5])
    if (isTRUE(add_zero)) out[b <= zero_repl * 10] <- "0"
    out
  }

  # optional binned medians in x
  bin_df <- NULL
  if (isTRUE(add_bins) && nrow(dd) > 0) {
    bin_df <- dd %>%
      dplyr::group_by(method_key) %>%
      dplyr::mutate(bin = cut(stat, breaks = bins, include.lowest = TRUE)) %>%
      dplyr::group_by(method_key, bin) %>%
      dplyr::summarise(x = median(stat, na.rm = TRUE),
                       y = median(pval_plot, na.rm = TRUE),
                       .groups = "drop") %>%
      dplyr::filter(is.finite(x), is.finite(y))
  }

  p <- ggplot(dd, aes(x = stat, y = pval_plot)) +
    {
      if (isTRUE(add_path))
        geom_path(aes(group = method_key),
                  linewidth = 0.4, alpha = 0.5, color = "grey40")
    } +
    geom_point(aes(fill = ptype),
               shape = 21, size = 1.6, alpha = 0.75,
               color = "gray20", stroke = 0.2, na.rm = TRUE) +
    {
      if (isTRUE(add_bins) && nrow(bin_df) > 0)
        geom_line(data = bin_df, aes(x = x, y = y),
                  linewidth = 0.7, inherit.aes = FALSE)
    } +
    {
      if (isTRUE(add_loess))
        geom_smooth(method = "loess", se = FALSE, linewidth = 0.6,
                    span = 0.8, color = "black")
    } +
    facet_wrap(~ method_key, nrow = 1, labeller = as_labeller(lab_map), scales = "free_x") +
    scale_y_log10(breaks = log_breaks, labels = label_fun) +
    scale_x_log10() +
    coord_cartesian(ylim = c(lower_limit, 1)) +
    scale_fill_manual(values = c("GPD fit" = "#F8766D", "empirical" = "gray"),
                      name = "p-value type") +
    labs(x = expression(d[W]^2 ~ "(ordered)"),
         y = expression(p)) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14),
      strip.text = element_text(size = 14),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(color = "gray85")
    )

  p
}


```

```{r pvals_vs_odw2_wad_pau_pac_10k, fig.width=10, fig.height=5, dpi=200}
plot_p_vs_stat_ordered(
  df_pvals_global,
  methods = c("permApprox_constrained_1000",
              "permApprox_constrained_10000"),
  label = "long",
  add_path = FALSE,   # shows the stat-ordered trace
  add_bins = FALSE,
  add_loess = FALSE
)
```

```{r}
plot_p_vs_ordered_stat <- function(df,
                                   methods = c("waddR_1000",
                                               "permApprox_unconstrained_1000",
                                               "permApprox_constrained_1000"),
                                   label      = c("short","long"),
                                   zero_repl  = 1e-320,
                                   add_zero   = TRUE,
                                   log_breaks = c(1e-3, 1e-10, 1e-25, 1e-50, 1e-100, 1e-200, 1e-300),
                                   show_stat_ticks = TRUE) {

  label <- match.arg(label)
  lab_col <- if (label == "long") "method_label_long" else "method_label"

  dd <- df %>%
    dplyr::filter(method_key %in% methods) %>%
    dplyr::mutate(method_key = factor(method_key, levels = methods),
                  pval_plot  = dplyr::case_when(
                    is.na(pval)      ~ NA_real_,
                    pval == 0        ~ zero_repl,
                    pval < zero_repl ~ NA_real_,
                    TRUE             ~ pval
                  )) %>%
    dplyr::group_by(method_key) %>%
    dplyr::arrange(stat, .by_group = TRUE) %>%
    dplyr::mutate(rank_stat = dplyr::row_number(),
                  n_m = dplyr::n()) %>%
    dplyr::ungroup()

  # facet labels + stat range summaries for context
  sum_tbl <- dd %>% dplyr::group_by(method_key) %>%
    dplyr::summarise(
      lbl = dplyr::first(.data[[lab_col]]),
      stat_min = min(stat, na.rm=TRUE),
      stat_med = median(stat, na.rm=TRUE),
      stat_max = max(stat, na.rm=TRUE),
      .groups = "drop"
    )

  lab_map <- setNames(
    paste0(sum_tbl$lbl,
           if (isTRUE(show_stat_ticks))
             sprintf("\n[min=%.3g | med=%.3g | max=%.3g]",
                     sum_tbl$stat_min, sum_tbl$stat_med, sum_tbl$stat_max)
           else ""),
    sum_tbl$method_key
  )

  # y-axis labels
  log_breaks <- sort(unique(log_breaks))
  if (isTRUE(add_zero)) log_breaks <- sort(unique(c(log_breaks, zero_repl)))
  lower_limit <- min(log_breaks)
  label_fun <- function(b) {
    out <- b
    out[b > 1e-5] <- scales::label_scientific()(b[b > 1e-5])
    if (isTRUE(add_zero)) out[b <= zero_repl * 10] <- "0"
    out
  }

  # pretty x breaks per facet (map ranks to corresponding stat values in strip)
  x_breaks_df <- dd %>%
    dplyr::group_by(method_key) %>%
    dplyr::summarise(
      n = dplyr::n(),
      b = list(unique(pmax(1, round(stats::quantile(1:n, probs = c(0, .25, .5, .75, 1)))))),
      .groups = "drop"
    )

  p <- ggplot(dd, aes(x = rank_stat, y = pval_plot)) +
    geom_point(aes(fill = ptype), shape = 21, size = 1.5, alpha = 0.7,
               color = "gray20", stroke = 0.2, na.rm = TRUE) +
    scale_y_log10(breaks = log_breaks, labels = label_fun) +
    coord_cartesian(ylim = c(lower_limit, 1)) +
    facet_wrap(~ method_key, nrow = 1, labeller = as_labeller(lab_map), scales = "free_x") +
    scale_fill_manual(values = c("GPD fit" = "#F8766D", "empirical" = "gray"),
                      name = "p-value type") +
    labs(x = "Rank of test statistic (within method)", y = expression(p)) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14),
      strip.text = element_text(size = 14),
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(color = "gray85")
    )

  # Per-facet x breaks (keeps ranks readable)
  p <- p + ggplot2::facet_wrap(
    ~ method_key, nrow = 1, labeller = as_labeller(lab_map), scales = "free_x"
  )

  p
}

```

```{r}
plot_p_vs_ordered_stat(
  df_pvals_global,
  methods = c("permApprox_constrained_1000",
              "permApprox_constrained_10000"),
  label = "short"
)
```


## Visualization of genes

```{r}
## Combine waddR TS (B = 1000) with permApprox(U/C) results

res_ts <- waddr_res_ts_1k %>%
  dplyr::rename(
    p.waddR     = pval,
    shape.waddR = shape
  ) %>%
  dplyr::left_join(
    tibble::tibble(
      gene                = waddr_res_ts_1k$gene,
      p.pApp.unconstr     = papprox_ts_unconstr_1k$p_unadjusted,
      shape.pApp.unconstr = papprox_ts_unconstr_1k$fit_result$shape
    ),
    by = "gene"
  ) %>%
  dplyr::left_join(
    tibble::tibble(
      gene               = waddr_res_ts_1k$gene,
      p.pApp.constr      = papprox_ts_constr_1k$p_unadjusted,
      shape.pApp.constr  = papprox_ts_constr_1k$fit_result$shape
    ),
    by = "gene"
  )

```

### Genes with waddR p = 0

```{r}
## Genes with waddR p = 0  — TS, B = 1000
df_zero_ts <- res_ts %>%
  dplyr::filter(p.waddR == 0) %>%
  dplyr::arrange(p.pApp.constr)

genes_selected_ts <- df_zero_ts %>% dplyr::pull(gene)

# show a few rows
kable(head(df_zero_ts))

# pick up to 6 genes to visualize
genes_to_plot <- head(genes_selected_ts, 6L)

# meta info for titles (ordered by d^2_W)
pane_meta_ts <- res_ts %>%
  dplyr::filter(gene %in% genes_to_plot) %>%
  dplyr::transmute(
    gene,
    d_wass2  = `d.wass^2`,
    p_constr = p.pApp.constr
  ) %>%
  dplyr::arrange(p_constr)

# facet order by increasing dW^2
gene_order_ts <- pane_meta_ts$gene

# facet labels: "GENE\n(d = ...; p = ...)"
lab_map_ts <- setNames(
  paste0(
    pane_meta_ts$gene,
    "\n(d = ", signif(pane_meta_ts$d_wass2, 3),
    "; p = ", formatC(pane_meta_ts$p_constr, format = "e", digits = 2), ")"
  ),
  pane_meta_ts$gene
)

```

#### Expression densities 

```{r zero_pvals_density_ts, fig.width=10, fig.height=5, dpi=200}
## Expression densities (TS, non-zero counts only), log1p(count)

counts_decidua <- counts(sce_decidua)
rownames(counts_decidua) <- make_short_gene(rownames(counts_decidua))
counts_blood <- counts(sce_blood)
rownames(counts_blood) <- make_short_gene(rownames(counts_blood))


# group 1 = blood
counts_blood_sel <- counts_blood[genes_to_plot, , drop = FALSE]
expr_blood_df <- counts_blood_sel %>%
  as.matrix() %>%
  t() %>%                                   # cells as rows
  tibble::as_tibble() %>%
  tibble::add_column(cell = colnames(sce_blood), .before = 1) %>%
  tidyr::pivot_longer(-cell, names_to = "gene", values_to = "count") %>%
  dplyr::mutate(
    expression_log = log(count),
    expression = count,
    group      = "blood"
  )

# group 2 = decidua
counts_decidua_sel <- counts_decidua[genes_to_plot, , drop = FALSE]
expr_decidua_df <- counts_decidua_sel %>%
  as.matrix() %>%
  t() %>%
  tibble::as_tibble() %>%
  tibble::add_column(cell = colnames(sce_decidua), .before = 1) %>%
  tidyr::pivot_longer(-cell, names_to = "gene", values_to = "count") %>%
  dplyr::mutate(
    expression_log = log(count),
    expression = count,
    group      = "decidua"
  )

plot_df_expr_ts <- dplyr::bind_rows(expr_blood_df, expr_decidua_df) %>%
  dplyr::filter(count > 0) %>%                           # non-zero counts only
  dplyr::mutate(
    gene  = factor(gene, levels = gene_order_ts),
    group = factor(group, levels = c("blood", "decidua"))
  )

ggplot(plot_df_expr_ts, aes(x = expression, fill = group)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = palette_groups) +
  facet_wrap(~ gene, scales = "free", labeller = as_labeller(lab_map_ts)) +
  labs(
    x = "Expression value",
    y = "Density"
  ) +
  scale_x_log10() +
  theme_minimal() +
  theme(
    axis.text          = element_text(size = 12),
    axis.title         = element_text(size = 14),
    strip.text         = element_text(size = 12),
    legend.title       = element_text(size = 14),
    legend.text        = element_text(size = 12),
    panel.spacing      = unit(2, "lines"),
    legend.position    = "bottom",
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "gray80")
  )
```

#### Histograms

```{r zero_pvals_hist_ts, fig.width=10, fig.height=5, dpi=200}
## log1p(count) histograms (TS, non-zero counts only)

plot_df_expr_hist_ts <- plot_df_expr_ts   # same data as above

ggplot(plot_df_expr_hist_ts, aes(x = expression, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  scale_fill_manual(values = palette_groups) +
  scale_x_log10() +
  facet_wrap(~ gene, scales = "free", labeller = as_labeller(lab_map_ts)) +
  labs(
    x = "Expression value",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text          = element_text(size = 12),
    axis.title         = element_text(size = 14),
    strip.text         = element_text(size = 12),
    legend.title       = element_text(size = 14),
    legend.text        = element_text(size = 12),
    panel.spacing      = unit(2, "lines"),
    legend.position    = "bottom",
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "gray80")
  )
```

#### Quantile plots

```{r zero_pvals_quant_ts, fig.width=10, fig.height=5, dpi=200}
## Quantile functions (ECDF) for genes with waddR p = 0 (TS, log1p counts)

ggplot(plot_df_expr_ts, aes(x = expression, color = group)) +
  stat_ecdf(geom = "step", linewidth = 1) +
  scale_color_manual(values = palette_groups, name = NULL) +
  scale_y_log10() +
  facet_wrap(~ gene, scales = "free", labeller = as_labeller(lab_map_ts)) +
  labs(
    x = "Expression value",
    y = "Cumulative probability"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text       = element_text(size = 12),
    axis.title      = element_text(size = 14),
    legend.text     = element_text(size = 12),
    strip.text      = element_text(size = 12),
    panel.spacing   = unit(2, "lines"),
    panel.grid.minor = element_blank()
  )
```

#### Permutation histograms (B = 1000)

```{r zero_pvals_perm_dist_ts, fig.width=10, fig.height=5, dpi=200}
## Permutation histograms (TS, waddR p = 0, B = 1000)

# Ensure permutation matrix has colnames = gene names
if (is.null(colnames(perm_stats_ts_1k))) {
  colnames(perm_stats_ts_1k) <- waddr_res_ts_1k$gene
}

perm_df_ts <- perm_stats_ts_1k[, genes_to_plot, drop = FALSE]

perm_long_ts <- tibble::as_tibble(perm_df_ts) %>%
  dplyr::mutate(perm = dplyr::row_number()) %>%
  tidyr::pivot_longer(cols = -perm,
                      names_to = "gene",
                      values_to = "perm_value") %>%
  dplyr::left_join(
    waddr_res_ts_1k %>%
      dplyr::select(gene, d_wass2 = `d.wass^2`),
    by = "gene"
  ) %>%
  dplyr::mutate(gene = factor(gene, levels = gene_order_ts))

ggplot(perm_long_ts, aes(x = perm_value)) +
  geom_histogram(bins = 30, fill = "gray70", color = "white") +
  geom_vline(aes(xintercept = d_wass2), color = "red", linewidth = 0.8) +
  facet_wrap(~ gene, scales = "free", labeller = as_labeller(lab_map_ts)) +
  labs(
    title = "Permutation distributions (TS, waddR p = 0)",
    x = expression(d[W]^2),
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text          = element_text(size = 12),
    axis.title         = element_text(size = 14),
    strip.text         = element_text(size = 12),
    legend.title       = element_text(size = 14),
    legend.text        = element_text(size = 12),
    panel.spacing      = unit(2, "lines"),
    legend.position    = "bottom",
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "gray80")
  )
```


### Further significant genes (TS, B = 1000)

Now the “top non-zero” genes, ordered by constrained permApprox p-values.

```{r}
## Significant genes with waddR p != 0, ordered by permApprox(C)

df_sig_ts <- res_ts %>%
  dplyr::filter(p.waddR != 0) %>%
  dplyr::arrange(p.pApp.constr)

# Pick up to 6 genes
genes_to_plot <- df_sig_ts$gene[1:6]

pane_meta_ts <- df_sig_ts %>%
  dplyr::filter(gene %in% genes_to_plot) %>%
  dplyr::select(gene,
                d_wass2 = `d.wass^2`,
                p_constr = p.pApp.constr)

gene_order_ts <- pane_meta_ts$gene

lab_map_ts <- setNames(
  paste0(
    pane_meta_ts$gene,
    "\n(d = ", signif(pane_meta_ts$d_wass2, 3),
    "; p = ", formatC(pane_meta_ts$p_constr,
                      format = "e", digits = 2), ")"
  ),
  pane_meta_ts$gene
)
```

#### Expression densities

```{r sig_pvals_density_ts, fig.width=10, fig.height=5, dpi=200}
## Expression densities (TS, non-zero counts only), log1p(count)

# blood
counts_blood_sel <- counts_blood[genes_to_plot, , drop = FALSE]
expr_blood_df <- counts_blood_sel %>%
  as.matrix() %>%
  t() %>%
  tibble::as_tibble() %>%
  tibble::add_column(cell = colnames(sce_blood), .before = 1) %>%
  tidyr::pivot_longer(-cell, names_to = "gene", values_to = "count") %>%
  dplyr::mutate(
    abundance = count,
    group     = "blood"
  )

# decidua
counts_decidua_sel <- counts_decidua[genes_to_plot, , drop = FALSE]
expr_decidua_df <- counts_decidua_sel %>%
  as.matrix() %>%
  t() %>%
  tibble::as_tibble() %>%
  tibble::add_column(cell = colnames(sce_decidua), .before = 1) %>%
  tidyr::pivot_longer(-cell, names_to = "gene", values_to = "count") %>%
  dplyr::mutate(
    abundance = count,
    group     = "decidua"
  )

plot_df_expr_ts <- dplyr::bind_rows(expr_blood_df, expr_decidua_df) %>%
  dplyr::filter(count > 0) %>%
  dplyr::mutate(
    gene  = factor(gene, levels = gene_order_ts),
    group = factor(group, levels = c("blood", "decidua"))
  )

ggplot(plot_df_expr_ts, aes(x = abundance, fill = group)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = palette_groups) +
  scale_x_log10() +
  facet_wrap(~ gene, scales = "free", labeller = as_labeller(lab_map_ts)) +
  labs(
    x = "Expression value",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    axis.text          = element_text(size = 12),
    axis.title         = element_text(size = 14),
    strip.text         = element_text(size = 12),
    legend.title       = element_text(size = 14),
    legend.text        = element_text(size = 12),
    panel.spacing      = unit(2, "lines"),
    legend.position    = "bottom",
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "gray80")
  )
```

#### Histograms

```{r sig_pvals_hist_ts, fig.width=10, fig.height=5, dpi=200}

plot_df_hist_ts <- plot_df_expr_ts

ggplot(plot_df_hist_ts, aes(x = abundance, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  scale_fill_manual(values = palette_groups) +
  scale_x_log10() +
  facet_wrap(~ gene, scales = "free", labeller = as_labeller(lab_map_ts)) +
  labs(
    x = "Expression value",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text          = element_text(size = 12),
    axis.title         = element_text(size = 14),
    strip.text         = element_text(size = 12),
    legend.title       = element_text(size = 14),
    legend.text        = element_text(size = 12),
    panel.spacing      = unit(2, "lines"),
    legend.position    = "bottom",
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "gray80")
  )
```

#### Quantile plots

```{r sig_pvals_quant_ts, fig.width=10, fig.height=5, dpi=200}
## Quantile functions (ECDF) for genes with waddR p = 0 (TS, log1p counts)

ggplot(plot_df_expr_ts, aes(x = abundance, color = group)) +
  stat_ecdf(geom = "step", linewidth = 1) +
  scale_color_manual(values = palette_groups, name = NULL) +
  scale_y_log10() +
  facet_wrap(~ gene, scales = "free", labeller = as_labeller(lab_map_ts)) +
  labs(
    x = "Expression value",
    y = "Cumulative probability"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text       = element_text(size = 12),
    axis.title      = element_text(size = 14),
    legend.text     = element_text(size = 12),
    strip.text      = element_text(size = 12),
    panel.spacing   = unit(2, "lines"),
    panel.grid.minor = element_blank()
  )
```

#### Permutation histograms (TS, significant genes)

```{r sig_pvals_perm_dist_ts, fig.width=10, fig.height=5, dpi=200}
## Permutation histograms (TS, further significant genes)

if (is.null(colnames(perm_stats_ts_1k))) {
  colnames(perm_stats_ts_1k) <- waddr_res_ts_1k$gene
}

perm_df_ts <- perm_stats_ts_1k[, genes_to_plot, drop = FALSE]

perm_long_ts <- tibble::as_tibble(perm_df_ts) %>%
  dplyr::mutate(perm = dplyr::row_number()) %>%
  tidyr::pivot_longer(cols = -perm,
                      names_to = "gene",
                      values_to = "perm_value") %>%
  dplyr::left_join(
    waddr_res_ts_1k %>%
      dplyr::select(gene, d_wass2 = `d.wass^2`),
    by = "gene"
  ) %>%
  dplyr::mutate(gene = factor(gene, levels = gene_order_ts))

ggplot(perm_long_ts, aes(x = perm_value)) +
  geom_histogram(bins = 30, fill = "gray70", color = "white") +
  geom_vline(aes(xintercept = d_wass2), color = "red", linewidth = 0.8) +
  facet_wrap(~ gene, scales = "free", labeller = as_labeller(lab_map_ts)) +
  labs(
    title = "Permutation distributions (TS, selected significant genes)",
    x = expression(d[W]^2),
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text          = element_text(size = 12),
    axis.title         = element_text(size = 14),
    strip.text         = element_text(size = 12),
    legend.title       = element_text(size = 14),
    legend.text        = element_text(size = 12),
    panel.spacing      = unit(2, "lines"),
    legend.position    = "bottom",
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "gray80")
  )
```

### Top significant genes

```{r}
# Top 3 genes with waddR p = 0 and smallest permApprox(C) p-values (TS, B = 1000)

# Select genes: waddR p = 0, ordered by permApprox(C) p-value
df_zero_ts_top <- res_ts %>%
  #dplyr::filter(p.waddR == 0, !is.na(p.pApp.constr)) %>%
  dplyr::arrange(p.pApp.constr) %>%
  dplyr::slice_head(n = 4)

genes_to_plot <- df_zero_ts_top$gene

# meta info for facet labels (include d and p from constrained permApprox)
pane_meta_ts <- df_zero_ts_top %>%
  dplyr::transmute(
    gene,
    d_wass2  = `d.wass^2`,
    p_constr = p.pApp.constr
  )

gene_order_ts <- pane_meta_ts$gene

lab_map_ts <- setNames(
  paste0(
    pane_meta_ts$gene,
    "\n(d = ", signif(pane_meta_ts$d_wass2, 3),
    "; p = ", formatC(pane_meta_ts$p_constr, format = "e", digits = 2), ")"
  ),
  pane_meta_ts$gene
)

lab_map_empty <- setNames(rep("", length(gene_order_ts)), gene_order_ts)

# Build long data frame with non-zero expression counts (blood vs decidua)
## blood
counts_blood_sel <- counts_blood[genes_to_plot, , drop = FALSE]

expr_blood_df <- counts_blood_sel %>%
  as.matrix() %>%
  t() %>%  # cells as rows
  tibble::as_tibble() %>%
  tibble::add_column(cell = colnames(sce_blood), .before = 1) %>%
  tidyr::pivot_longer(-cell, names_to = "gene", values_to = "count") %>%
  dplyr::mutate(
    group = "blood"
  )

## decidua
counts_decidua_sel <- counts_decidua[genes_to_plot, , drop = FALSE]

expr_decidua_df <- counts_decidua_sel %>%
  as.matrix() %>%
  t() %>%
  tibble::as_tibble() %>%
  tibble::add_column(cell = colnames(sce_decidua), .before = 1) %>%
  tidyr::pivot_longer(-cell, names_to = "gene", values_to = "count") %>%
  dplyr::mutate(
    group = "decidua"
  )

## combined, non-zero only
plot_df_expr_ts <- dplyr::bind_rows(expr_blood_df, expr_decidua_df) %>%
  dplyr::filter(count > 0) %>%
  dplyr::mutate(
    gene  = factor(gene,  levels = gene_order_ts),
    group = factor(group, levels = c("blood", "decidua"))
  )

## Permutation distributions for the same top genes (TS, B = 1000)
# ensure perm_stats_ts_1k has gene column names
if (is.null(colnames(perm_stats_ts_1k))) {
  colnames(perm_stats_ts_1k) <- waddr_res_ts_1k$gene
}

perm_df_top <- perm_stats_ts_1k[, genes_to_plot, drop = FALSE]

perm_long_top <- tibble::as_tibble(perm_df_top) %>%
  dplyr::mutate(perm = dplyr::row_number()) %>%
  tidyr::pivot_longer(
    cols      = -perm,
    names_to  = "gene",
    values_to = "perm_value"
  ) %>%
  dplyr::left_join(
    waddr_res_ts_1k %>%
      dplyr::select(gene, d_wass2 = `d.wass^2`),
    by = "gene"
  ) %>%
  dplyr::mutate(
    gene = factor(gene, levels = gene_order_ts)
  )

```


```{r zero_top4_hist_ecdf_ts, fig.width=12, fig.height=8, dpi=200}
# Top row: histograms (log10 x-axis), titles with d and p
p_hist <- ggplot(plot_df_expr_ts, aes(x = count, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  scale_x_log10() +
  scale_fill_manual(values = palette_groups, name = NULL) +
  facet_wrap(~ gene, scales = "free", nrow = 1,
             labeller = as_labeller(lab_map_ts)) +
  labs(
    x = "Expression value",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text        = element_text(size = 13),
    axis.title       = element_text(size = 14),
    strip.text       = element_text(size = 13),
    legend.title     = element_text(size = 14),
    legend.text      = element_text(size = 14),
    panel.spacing    = unit(1.5, "lines"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  )

# Bottom row: ECDF / quantile plots (same genes, same labels)
p_ecdf <- ggplot(plot_df_expr_ts, aes(x = count, color = group)) +
  stat_ecdf(geom = "step", linewidth = 1) +
  scale_x_log10() +
  scale_color_manual(values = palette_groups, name = NULL) +
  facet_wrap(~ gene, scales = "free", nrow = 1,
             labeller = as_labeller(lab_map_empty)) +
  labs(
    x = "Expression value",
    y = "Cumulative probability"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text        = element_text(size = 13),
    axis.title       = element_text(size = 14),
    strip.text       = element_text(size = 12),
    legend.title     = element_text(size = 14),
    legend.text      = element_text(size = 14),
    legend.position  = "bottom",  
    panel.spacing    = unit(1.5, "lines"),
    panel.grid.minor = element_blank()
  )

# Third row: permutation distributions with observed statistic (red line)
p_perm <- ggplot(perm_long_top, aes(x = perm_value)) +
  geom_histogram(bins = 30, fill = "snow4", colour = "white") +
  #geom_vline(aes(xintercept = d_wass2), colour = "red", linewidth = 0.8) +
  facet_wrap(
    ~ gene, scales = "free", nrow = 1,
    labeller = as_labeller(lab_map_empty)  # or lab_map_ts if you want d,p again
  ) +
  labs(
    x = expression(d[W]^2),
    y = "Permutation count"
  ) +
  theme_minimal() +
  theme(
    axis.text          = element_text(size = 13),
    axis.title         = element_text(size = 14),
    strip.text         = element_text(size = 12),
    panel.spacing      = unit(1.5, "lines"),
    legend.position    = "none",
    panel.grid.minor   = element_blank(),
    panel.grid.major.y = element_line(color = "gray80")
  )

# Combine: top = histograms, bottom = ECDF
(p_hist / p_ecdf / p_perm) +
  plot_layout(heights = c(1, 1, 1), guides = "collect") &
  theme(legend.position = "bottom")
```


## LaTeX tables for TS (B = 1000)

```{r}
sci <- function(x, digits = 2) {
  ifelse(is.na(x), NA, formatC(x, format = "e", digits = digits))
}
```

### Top genes (raw p-values)

```{r table_top_effects_ts, results = "asis", message = FALSE}
n_top <- 20 

top_tbl_ts <- res_ts %>%
  dplyr::filter(!is.na(p.pApp.constr)) %>%   # keep rows with permApprox(C) p-values
  dplyr::arrange(p.pApp.constr) %>%          # sort by constrained permApprox p-value
  dplyr::slice_head(n = n_top) %>%
  dplyr::mutate(
    d_wass2_round  = round(`d.wass^2`, 3),
    p_waddR_sci    = sci(p.waddR),
    p_pApp_U_sci   = sci(p.pApp.unconstr),
    p_pApp_C_sci   = sci(p.pApp.constr),
    xi_waddR_round = round(shape.waddR, 3),
    xi_U_round     = round(shape.pApp.unconstr, 3),
    xi_C_round     = round(shape.pApp.constr, 3)
  ) %>%
  dplyr::transmute(
    Gene                          = gene,
    `$d_{W}^2$`                   = d_wass2_round,
    `$p_{\\text{waddR}}$`         = p_waddR_sci,
    `$p_{\\text{permApprox(U)}}$` = p_pApp_U_sci,
    `$p_{\\text{permApprox(C)}}$` = p_pApp_C_sci,
    `$\\hat{\\xi}_{\\text{waddR}}$`         = xi_waddR_round,
    `$\\hat{\\xi}_{\\text{(U)}}$` = xi_U_round,
    `$\\hat{\\xi}_{\\text{(C)}}$` = xi_C_round
  )

tab_latex_ts <- kable(
  top_tbl_ts,
  format   = "latex",
  booktabs = TRUE,
  escape   = FALSE,
  caption  = paste0(
    "Top ", n_top,
    " genes for the two-stage (TS) waddR approach, ",
    "sorted by constrained permApprox p-values (B = 1000)."
  ),
  label    = "tab:waddr_permApprox_TS_top_effects_sc"
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size     = 8
  )

print(tab_latex_ts)

# Save as .tex
save_kable(tab_latex_ts, path_target("permApprox_waddr_TS_top_effects_sc_table.tex"))
```

### Top genes (BH-adjusted p-values)

```{r table_top_effects_ts_BH, results = "asis", message = FALSE}
n_top <- 20 

res_ts_adj <- res_ts %>%
  dplyr::mutate(
    p_adj_waddR  = p.adjust(p.waddR,         method = "BH"),
    p_adj_pApp_U = p.adjust(p.pApp.unconstr, method = "BH"),
    p_adj_pApp_C = p.adjust(p.pApp.constr,   method = "BH")
  )

top_tbl_ts_BH <- res_ts_adj %>%
  dplyr::filter(!is.na(p_adj_pApp_C)) %>%
  dplyr::arrange(p_adj_pApp_C) %>%
  dplyr::slice_head(n = n_top) %>%
  dplyr::mutate(
    d_wass2_round    = round(`d.wass^2`, 3),
    p_adj_waddR_sci  = sci(p_adj_waddR),
    p_adj_pApp_U_sci = sci(p_adj_pApp_U),
    p_adj_pApp_C_sci = sci(p_adj_pApp_C),
    xi_waddR_round   = round(shape.waddR, 3),
    xi_U_round       = round(shape.pApp.unconstr, 3),
    xi_C_round       = round(shape.pApp.constr, 3)
  ) %>%
  dplyr::transmute(
    Gene                            = gene,
    `$d_{W}^2$`                     = d_wass2_round,
    `$p_{\\text{waddR,BH}}$`        = p_adj_waddR_sci,
    `$p_{\\text{permApprox(U),BH}}$`= p_adj_pApp_U_sci,
    `$p_{\\text{permApprox(C),BH}}$`= p_adj_pApp_C_sci,
    `$\\hat{\\xi}_{\\text{waddR}}$` = xi_waddR_round,
    `$\\hat{\\xi}_{\\text{(U)}}$` = xi_U_round,
    `$\\hat{\\xi}_{\\text{(C)}}$` = xi_C_round
  )

tab_latex_ts_BH <- kable(
  top_tbl_ts_BH,
  format   = "latex",
  booktabs = TRUE,
  escape   = FALSE,
  caption  = paste0(
    "Top ", n_top,
    " genes for the two-stage (TS) waddR approach based on BH-adjusted p-values, ",
    "sorted by BH-adjusted constrained permApprox p-values (B = 1000)."
  ),
  label    = "tab:waddr_permApprox_TS_top_effects_BH_sc"
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size     = 8
  )

print(tab_latex_ts_BH)

save_kable(
  tab_latex_ts_BH,
  path_target("permApprox_waddr_TS_top_effects_BH_sc_table.tex")
)
```

## Summary

This application study demonstrates the benefit of applying `permApprox` to improve the robustness of p-value estimation in permutation-based differential tests. Especially for extreme test statistics, `permApprox` avoids zero p-values and thus enables more accurate and reproducible results.

## References

* Vento-Tormo,R. et al. (2018) Reconstructing the human ﬁrst trimester fetal--
maternal interface using single cell transcriptomics. Nature, 563, 347–353.

## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
