---
title: "Differential abundance analysis with `dacomp` - EBF vs non-EBF"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
always_allow_html: true
params:
  name: "02_micro_diff_abund_breastf" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "bb9f8d89-5eb6-408d-acfc-729e640abec2")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(fig.width = 10, 
                      fig.height = 6, 
                      dpi = 200,
                      fig.path = paste0('figures/', params, "/app_dacomp_"),
                      echo = FALSE)

refit_permApprox <- FALSE
```

This notebook demonstrates how to perform **differential abundance analysis** of microbiome count data using the **`dacomp`** package.

We use genus-level profiles from the **PASTURE** cohort (2-month samples). After basic prevalence filtering and defining two comparison groups (here, breastfeeding categories), we visualize the data (heatmaps, densities, histograms, ECDFs) and then apply **dacomp**:

1. **Select a stable reference set of non-DA taxa**,  
2. **Test all remaining taxa for differential abundance** using a Wilcoxon-based two-sample test with permutation p-values and DS-FDR correction.

The example uses the **raw count data** for inference (as required by dacomp), while relative abundances are used only for visualization.

In this notebook, we compare two breast feeding groups: 'EBF' (exclusive breast feeding until age 2 months) vs. 'non-EBF'.

## Global options and packages

```{r packages, message=FALSE, warning=FALSE}
#remotes::install_github("stefpeschel/permApprox", ref = "develop")
library(conflicted)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(kableExtra)
library(pheatmap)
library(ggrepel)
library(patchwork)

# Differential abundance with dacomp
library(dacomp)

# p-value approximation with permApprox
library(permApprox)

conflicted::conflicts_prefer(dplyr::count, dplyr::filter, dplyr::select, 
                             stats::var, dplyr::setdiff)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```


## Load data & set parameters

```{r load_data}
bact_phylo <- readRDS(path_source("PASTURE_data", paste0("bact_2m_genus.rds")))
sampdata    <- sample_data(bact_phylo) %>% as("data.frame")
bact_phylo
```

```{r}
# Rename "24_Enterobacteriaceae(F)"
taxtab <- tax_table(bact_phylo)
Genus <- taxtab[, "Genus"]
Genus[Genus == "24_Enterobacteriaceae(F)"] <- "Enterobacteriaceae(F)"
taxtab[, "Genus"] <- Genus

tax_table(bact_phylo) <- taxtab
taxa_names(bact_phylo) <- Genus
```

## Data preprocessing

### Taxa prevalence (including zero counts)

```{r expl_prevalence_hist}
# Total number of samples
n_samples <- nsamples(bact_phylo)

# Compute number of samples where each taxon is observed (abundance > 0)
taxa_prevalence <- apply(otu_table(bact_phylo), 1, function(x) sum(x > 0))

# Build dataframe
prevalence_df <- tibble(
  taxon = names(taxa_prevalence),
  prevalence = taxa_prevalence,
  prevalence_pct = 100 * taxa_prevalence / n_samples
)

# Plot histogram (absolute counts)
ggplot(prevalence_df, aes(x = prevalence)) +
  geom_histogram(binwidth = 5, fill = "#1F78B4", color = "white", alpha = 0.8) +
  labs(
    title = "Taxon prevalence across samples",
    x = "Number of samples with abundance > 0",
    y = "Number of taxa"
  ) +
  theme_minimal()
```

### Prevalence threshold

```{r expl_prevalence_hist_filter}
bact_phylo  <- prune_taxa(taxa_sums(bact_phylo) > 0, bact_phylo)

ggplot(prevalence_df, aes(x = prevalence_pct)) +
  geom_histogram(binwidth = 2, fill = "#1F78B4", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = 1, color = "1% threshold"), 
             linetype = "dashed", linewidth = 0.8) +
  geom_vline(aes(xintercept = 5, color = "5% threshold"), 
             linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(
    name = "Prevalence thresholds",
    values = c("1% threshold" = "orange", "5% threshold" = "red")
  ) +
  labs(
    title = "Taxon prevalence across samples (including zero-abundance taxa)",
    x = "Prevalence (% of samples)",
    y = "Number of taxa"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )
```

### Taxa filtering

We keep only taxa with a non-zero count in at least 1% of the samples.

```{r}
# Minimum prevalence threshold (5% of total samples)
prev_threshold <- 0.01 * nsamples(bact_phylo)

# Compute prevalence per taxon
taxa_prevalence <- apply(otu_table(bact_phylo), 1, function(x) sum(x > 0))

# Filter taxa with prevalence >= threshold
keep_taxa <- taxa_prevalence >= prev_threshold

# Prune the phyloseq object
bact_phylo <- prune_taxa(keep_taxa, bact_phylo)
bact_phylo
```

### Define comparison groups

```{r}
cat("breast_excl_cat1:")
table(sampdata$breast_excl_cat1)

# Build new variable with categories EBF and non-EBF
sampdata <- sampdata %>%
  mutate(BF_cat = case_when(
    breast_excl_cat1 == "0" ~ "non-EBF",
    breast_excl_cat1 == ">=2" ~ "EBF",
    TRUE ~ NA_character_
  ))
sample_data(bact_phylo) <- sampdata
cat("\nNew variable:")
table(sampdata$BF_cat)
```

```{r}
## Global color palette (consistent across all plots)
col_group1 <- "#1f77b4"     # blue
col_group2 <- "darkorange2" # orange

palette_groups <- c("non-EBF" = col_group1, 
                    "EBF"     = col_group2)

# New variable
comp_var <- "BF_cat"

# Extract the vector to split on
var_vec <- sampdata[[comp_var]]
names(var_vec) <- rownames(sampdata)

comp_val1 <- label1 <- "non-EBF"
comp_val2 <- label2 <- "EBF"

# categorical split by values
group1_ids <- rownames(sampdata)[var_vec %in% comp_val1]
group2_ids <- rownames(sampdata)[var_vec %in% comp_val2]

# # make group sizes equal
# set.seed(42)
# min_length <- min(length(group1_ids), length(group2_ids))
# group2_ids <- sample(group2_ids, min_length, replace = FALSE)

# Prune phyloseq objects for each group
bact_g1 <- prune_samples(group1_ids, bact_phylo)
bact_g2 <- prune_samples(group2_ids, bact_phylo)

# Create a tibble for plotting the distribution of comp_var
dist_df <- tibble(
  sample = names(var_vec),
  value  = var_vec,
  group  = case_when(
    sample %in% group1_ids ~ label1,
    sample %in% group2_ids ~ label2,
    TRUE                   ~ "Other"
  )
) %>%
  mutate(group = factor(group, levels = c(label1, "Other", label2)))

# Numeric: histogram of values
p_dist <- ggplot(dist_df, aes(x = group, fill = group)) +
  geom_histogram(bins = 30, color = "black", alpha = .6) +
  scale_fill_manual(values = palette_groups) +
  labs(x = comp_var, y = "Count")
```


```{r expl_group_sizes, fig.width=5, fig.height=5, dpi=200}
# Categorical: bar plot
ggplot(dist_df %>% filter(group != "Other"),  # optional: drop "Other"
                 aes(x = base::as.factor(group), fill = group)) +
  geom_bar(color = "black", alpha = .6, position = "dodge") +
  scale_fill_manual(values = palette_groups) +
  labs(x = "", y = "Sample count") +
  theme_minimal() +
  theme(legend.position="none",
        axis.text  = element_text(size = 12),
        axis.title = element_text(size = 14))
```

```{r}
cat("Samples in group 1:", nsamples(bact_g1), "\n")
cat("Samples in group 2:", nsamples(bact_g2))
```

### Relative abundance transformation

```{r}
## Absolute counts
counts_g1 <- t(as(otu_table(bact_g1), "matrix"))
counts_g2 <- t(as(otu_table(bact_g2), "matrix"))

# Compute relative abundances per sample
abund_rel_g1 <- counts_g1 / rowSums(counts_g1)
abund_rel_g2 <- counts_g2 / rowSums(counts_g2)

# Combine abundance matrices for the two groups
abund_rel_combined <- rbind(abund_rel_g1, abund_rel_g2)

# Assign group labels
sample_labels <- c(rep(label1, nrow(abund_rel_g1)),
                   rep(label2, nrow(abund_rel_g2)))

```


## Data exploration

```{r}
# Top 20 overall taxa
top_taxa <- names(sort(taxa_sums(bact_phylo), decreasing = TRUE))[1:20]

mat1 <- abund_rel_g1[, top_taxa]
mat2 <- abund_rel_g2[, top_taxa]

combined <- rbind(mat1, mat2)
labels   <- c(rep(label2, nrow(mat2)), rep(label1, nrow(mat1)))

plot_df <- as.data.frame(combined) %>%
  rownames_to_column("sample") %>%
  mutate(group = labels) %>%
  pivot_longer(-c(sample, group), names_to = "taxon", values_to = "abundance") %>%
  mutate(taxon_short = stringr::str_trunc(taxon, width = 15, side = "right"))

# Ensure order of facets follows the order in `top_taxa`
top_taxa_short <- stringr::str_trunc(top_taxa[1:6], width = 15, side = "right")

# Set taxon_short as a factor with desired levels
plot_df <- plot_df %>%
  mutate(taxon_short = factor(taxon_short, levels = top_taxa_short))
```

### Library sizes

```{r expl_library_sizes, fig.width=5, fig.height=5, dpi=200}
lib_sizes_df <- tibble(
  sample   = c(rownames(counts_g1), rownames(counts_g2)),
  group    = factor(c(rep(label1, nrow(counts_g1)),
                      rep(label2, nrow(counts_g2))),
                    levels = c(label1, label2)),
  lib_size = c(rowSums(counts_g1),
               rowSums(counts_g2))
)

ggplot(lib_sizes_df, aes(x = group, y = lib_size, fill = group)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
  scale_fill_manual(values = palette_groups) +
  labs(
    #title = "Library sizes by group",
    x = NULL,
    y = "Total read count per sample"
  ) +
  theme_minimal() +
  theme(
    legend.position="none",
    axis.text  = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

```

### Prevalence vs mean relative abundance

```{r expl_prev_mean_scatter, fig.width=6, fig.height=4, dpi=200}
## Prevalence and mean relative abundance within the selected samples
rel_comb <- rbind(abund_rel_g1, abund_rel_g2)

prev_vec <- colSums(rel_comb > 0)
n_sel    <- nrow(rel_comb)

prev_mean_df <- tibble(
  taxon        = colnames(rel_comb),
  prevalence   = prev_vec / n_sel,
  mean_rel_ab  = colMeans(rel_comb)
)

ggplot(prev_mean_df,
       aes(x = prevalence, y = mean_rel_ab)) +
  geom_point(alpha = 0.6) +
  scale_y_log10() +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    #title = "Prevalence vs mean relative abundance",
    x = "Prevalence (proportion of selected samples with > 0 counts)",
    y = "Mean relative abundance (log10 scale)"
  ) +
  theme_minimal() +
  theme(
    legend.position="none",
    axis.text  = element_text(size = 12),
    axis.title = element_text(size = 12)
  )
```


### Alpha diversity

```{r}
# Restrict to the samples used in the dacomp comparison
bact_alpha <- prune_samples(c(group1_ids, group2_ids), bact_phylo)

# Standard alpha diversity estimates from phyloseq
alpha_est <- estimate_richness(
  bact_alpha,
  measures = c("Observed", "Shannon", "InvSimpson")
) %>%
  tibble::rownames_to_column("sample")

# Map sample IDs to group labels used in the dacomp script
alpha_df <- alpha_est %>%
  dplyr::mutate(
    GiniSimpson = 1 - 1 / InvSimpson
  ) %>%
  dplyr::left_join(
    tibble::tibble(
      sample = c(group1_ids, group2_ids),
      group  = c(rep(label1, length(group1_ids)),
                 rep(label2, length(group2_ids)))
    ),
    by = "sample"
  ) %>%
  dplyr::mutate(
    group = factor(group, levels = c(label1, label2))
  )

head(alpha_df)
```

```{r}
plot_alpha_metric <- function(df, metric, ylab,
                              palette = palette_groups,
                              title = NULL,
                              jitter_width = 0.1) {
  
  stopifnot(metric %in% names(df))
  
  # Wilcoxon rank-sum test between the two groups
  wt <- wilcox.test(df[[metric]] ~ df$group)
  pval <- wt$p.value
  
  y_max   <- max(df[[metric]], na.rm = TRUE)
  y_line  <- y_max * 1.05
  y_label <- y_max * 1.10
  
  ggplot(df, aes(x = group, y = .data[[metric]], fill = group)) +
    # Jittered points
    geom_jitter(width = jitter_width,
                size = 1.6,
                alpha = 0.7,
                stroke = 0,
                aes(color = group)) +
    # Boxplot (median + IQR)
    geom_boxplot(alpha = 0.6,
                 width = 0.4,
                 outlier.shape = NA) +
    # Bracket + p-value
    geom_segment(aes(x = 1, xend = 2,
                     y = y_line, yend = y_line),
                 inherit.aes = FALSE) +
    annotate("text",
             x = 1.5, y = y_label, size = 5,
             label = paste0("p = ", signif(pval, 3))) +
    scale_fill_manual(values = palette) +
    scale_color_manual(values = palette) +
    labs(
      x = NULL,
      y = ylab,
      title = title
    ) +
    coord_cartesian(ylim = c(0, y_label * 1.05)) +
    theme_minimal() +
    theme(
      #legend.position = "none",
      axis.text  = element_text(size = 14),
      axis.text.x = element_blank(),
      axis.title.y = element_text(size = 14),
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 14),
      plot.title   = element_text(size = 14, face = "bold"),
      panel.grid.minor = element_blank()
    )
}

```

#### Richness (Observed genera)

```{r expl_alpha_richness}
p_alpha_richness <- plot_alpha_metric(
  alpha_df,
  metric = "Observed",
  ylab   = "Richness (genera)"
  #title  = "Alpha diversity: richness"
)
p_alpha_richness
```

#### Shannon diversity

```{r expl_alpha_shannon}
p_alpha_shannon <- plot_alpha_metric(
  alpha_df,
  metric = "Shannon",
  ylab   = "Shannon index"
  #title  = "Alpha diversity: Shannon"
)
p_alpha_shannon

```

#### Gini–Simpson index

```{r expl_alpha_gini_simpson}
p_alpha_gini <- plot_alpha_metric(
  alpha_df,
  metric = "GiniSimpson",
  ylab   = "Gini–Simpson index"
  #title  = "Alpha diversity: evenness"
)
p_alpha_gini
```

#### Alpha diversity combined

```{r expl_alpha_diversity, fig.width=11, fig.height=6, dpi=200}
p_alpha_richness + p_alpha_shannon + p_alpha_gini +
  plot_layout(guides = 'collect')
```


### Heatmap

```{r heatmap_top20_unclust, fig.width=10, fig.height=5, dpi=200}
# Combine abundance matrices for the two groups
abund_rel_combined <- rbind(abund_rel_g1, abund_rel_g2)

# Assign group labels
sample_labels <- c(rep(label1, nrow(abund_rel_g1)),
                   rep(label2, nrow(abund_rel_g2)))

# Get top 20 most abundant taxa in the *full* dataset
top_taxa <- names(sort(taxa_sums(bact_phylo), decreasing = TRUE))[1:20]

# Select only top taxa from the combined matrix
heatmap_mat <- abund_rel_combined[, top_taxa]

# Annotation data for the columns
annotation_col <- data.frame(group = factor(sample_labels))
rownames(annotation_col) <- rownames(heatmap_mat)

annotation_colors = list(group = palette_groups)

# Heatmap (non-clustered)
pheatmap(
  t(heatmap_mat),  # transpose: taxa as rows
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  show_rownames = TRUE,
  show_colnames = FALSE,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = paste("Relative abundances (Top 20 taxa:", label1, "vs", label2, ")")
)
```

```{r heatmap_top20_clust, fig.width=10, fig.height=5, dpi=200}
# Heatmap (clustered)
pheatmap(
  t(heatmap_mat),  # transpose: taxa as rows
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,  show_rownames = TRUE,
  show_colnames = FALSE,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  main = paste("Relative abundances (Top 20 taxa:", label1, "vs", label2, ")")
)
```

### Mean composition - Phylum level

```{r expl_mean_composition_phylum, fig.width=7, fig.height=5, dpi=200}
## ----- Aggregate genus-level relative abundances to Phylum -----

tax_tab <- as.data.frame(tax_table(bact_phylo))

# Map genus taxa (rows of tax_tab) to Phylum
genus_to_phylum <- tax_tab$Phylum
names(genus_to_phylum) <- rownames(tax_tab)

# Helper to aggregate one group to phylum
aggregate_to_phylum <- function(abund_rel_mat) {
  abund_rel_mat %>%
    as.data.frame() %>%
    mutate(sample = rownames(.)) %>%
    tidyr::pivot_longer(
      cols      = -sample,
      names_to  = "taxon",
      values_to = "rel_abund"
    ) %>%
    dplyr::mutate(phylum = genus_to_phylum[taxon]) %>%
    dplyr::group_by(sample, phylum) %>%
    dplyr::summarise(rel_abund = sum(rel_abund), .groups = "drop") %>%
    tidyr::pivot_wider(
      names_from  = phylum,
      values_from = rel_abund,
      values_fill = 0
    )
}

phylum_mat_g1 <- aggregate_to_phylum(abund_rel_g1)
phylum_mat_g2 <- aggregate_to_phylum(abund_rel_g2)

phylum_g1 <- as.matrix(phylum_mat_g1[, -1, drop = FALSE])
phylum_g2 <- as.matrix(phylum_mat_g2[, -1, drop = FALSE])

## ----- Group-wise mean relative abundances at Phylum level -----

mean_phylum_g1 <- colMeans(phylum_g1)
mean_phylum_g2 <- colMeans(phylum_g2)

mean_phylum_df <- tibble::tibble(
  phylum       = colnames(phylum_g1),
  mean_rel_g1  = mean_phylum_g1,
  mean_rel_g2  = mean_phylum_g2
) %>%
  dplyr::mutate(overall_mean = (mean_rel_g1 + mean_rel_g2) / 2) %>%
  dplyr::arrange(dplyr::desc(overall_mean))

mean_phylum_long <- mean_phylum_df %>%
  dplyr::select(phylum, mean_rel_g1, mean_rel_g2) %>%
  tidyr::pivot_longer(
    cols      = dplyr::starts_with("mean_rel_"),
    names_to  = "group",
    values_to = "rel_abund"
  ) %>%
  dplyr::mutate(
    group = dplyr::recode(
      group,
      "mean_rel_g1" = label1,   # non-EBF
      "mean_rel_g2" = label2    # EBF
    )
  )

## ----- Color palette: 7 phyla, all shown -----

# Okabe–Ito 7-color palette
base_cols <- c(
  "#E69F00", # orange
  "#56B4E9", # light blue
  "#009E73", # green
  "#F0E442", # yellow
  "#0072B2", # blue
  "#D55E00", # red
  "#CC79A7"  # purple
)

phylum_levels <- unique(mean_phylum_long$phylum)
# assumes length(phylum_levels) <= 7
phylum_colors <- setNames(base_cols[seq_along(phylum_levels)], phylum_levels)

## ----- Plot: stacked bar chart at Phylum level -----

ggplot(mean_phylum_long,
       aes(x = group, y = rel_abund, fill = phylum)) +
  geom_bar(stat = "identity", position = "stack", color = "white") +
  scale_fill_manual(values = phylum_colors, name = "Phylum") +
  labs(
    title = "Mean phylum composition",
    x     = NULL,
    y     = "Relative abundance (mean across samples)"
  ) +
  theme_minimal() +
  theme(
    axis.text    = element_text(size = 14),
    axis.title   = element_text(size = 14),
    plot.title   = element_text(size = 15, face = "bold"),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 14)
  )
```

Firmucutes are increased in non-EBF, which is in line with other studies. 

### Mean composition - Genus level

```{r expl_mean_composition_genus, fig.width=7, fig.height=5, dpi=200}
## Mean relative abundances per group (using dacomp samples)

mean_rel_g1 <- colMeans(abund_rel_g1)
mean_rel_g2 <- colMeans(abund_rel_g2)

mean_rel_df <- tibble::tibble(
  taxon       = colnames(abund_rel_g1),
  mean_rel_g1 = mean_rel_g1,
  mean_rel_g2 = mean_rel_g2
) %>%
  dplyr::mutate(
    overall_mean = (mean_rel_g1 + mean_rel_g2) / 2
  ) %>%
  dplyr::arrange(dplyr::desc(overall_mean))

# Number of top genera to show explicitly
K <- 7

top_taxa <- mean_rel_df$taxon[1:K]

mean_rel_long <- mean_rel_df %>%
  dplyr::select(taxon, mean_rel_g1, mean_rel_g2) %>%
  tidyr::pivot_longer(
    cols      = starts_with("mean_rel_"),
    names_to  = "group",
    values_to = "rel_abund"
  ) %>%
  dplyr::mutate(
    group = dplyr::recode(
      group,
      "mean_rel_g1" = label1,
      "mean_rel_g2" = label2
    ),
    taxon_plot = dplyr::if_else(taxon %in% top_taxa, taxon, "Other")
  ) %>%
  dplyr::group_by(group, taxon_plot) %>%
  dplyr::summarise(
    rel_abund = sum(rel_abund),
    .groups   = "drop"
  ) %>%
  dplyr::group_by(group) %>%
  dplyr::mutate(
    rel_abund = rel_abund / sum(rel_abund)
  ) %>%
  dplyr::ungroup()

# Order taxa so that "Other" is last in the legend
ordered_taxa <- c(top_taxa, "Other")
mean_rel_long$taxon_plot <- factor(mean_rel_long$taxon_plot,
                                   levels = ordered_taxa)

# Okabe–Ito palette for taxa (you can adjust / reuse your existing colors)
cb_palette <- c(
  "#E69F00", # orange
  "#56B4E9", # light blue
  "#009E73", # green
  "#F0E442", # yellow
  "#0072B2", # blue
  "#D55E00", # red
  "#CC79A7"  # purple
)
cb_other <- "#999999"   # grey for "Other"

taxa_colors <- c(cb_palette, cb_other)
names(taxa_colors) <- ordered_taxa

ggplot(mean_rel_long,
       aes(x = group, y = rel_abund, fill = taxon_plot)) +
  geom_bar(stat = "identity", position = "stack", color = "white") +
  scale_fill_manual(values = taxa_colors, name = "Genus") +
  labs(
    title = paste("Mean genus composition (top", K, "genera)"),
    x     = NULL,
    y     = "Relative abundance (mean across samples)"
  ) +
  theme_minimal() +
  theme(
    axis.text    = element_text(size = 14),
    axis.title   = element_text(size = 14),
    plot.title   = element_text(size = 15, face = "bold"),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 14)
  )
```

## dacomp - Differential abundance analysis

### Build count matrix and phenotype vector

```{r dacomp_data_prep}
# Combine counts into one matrix X: rows = samples, columns = taxa
X_counts <- rbind(counts_g1, counts_g2)

# Phenotype / group labels corresponding to the rows of X_counts
y_groups <- factor(
  c(rep(label1, nrow(counts_g1)),
    rep(label2, nrow(counts_g2))),
  levels = c(label1, label2)
)

dim(X_counts)
table(y_groups)
```

### Select reference taxa

```{r dacomp_select_refs}
set.seed(42)

# Reference selection:
# You can tweak minimal_TA and/or median_SD_threshold.
res_refs <- dacomp.select_references(
  X = X_counts,
  minimal_TA = 50,          # ensure enough counts under reference set
  verbose   = TRUE
)

res_refs
length(res_refs$selected_references)

tax_names <- colnames(X_counts)
cat("Reference taxa: \n")
tax_names[res_refs$selected_references]

# Diagnostic plot of reference scores
dacomp.plot_reference_scores(res_refs)
```

Of the 118 taxa, 93 are selected as reference. 

```{r dacmp_prev_var_plot, fig.width=6, fig.height=4, dpi=200}
## Prevalence–variance plot for genus level (à la Sommer et al.)

# X_counts: samples x taxa (same object you use for dacomp)
X <- X_counts

n_samples <- nrow(X)

## 1) Prevalence: proportion of samples with count > 0
taxa_prev <- colSums(X > 0) / n_samples

## 2) Variance of log((x + 1)/sum(x)) across samples
#    (this matches: transform_sample_counts(ps, function(x) log((x+1)/sum(x))))
X_logrel <- log(
  sweep(X + 1, 1, rowSums(X), FUN = "/")
)

taxa_var <- apply(X_logrel, 2, var, na.rm = TRUE)

## 3) Classification into "reference" vs "test" using the same thresholds
##    as in the script you showed (for genus level):
##        variance < 2  &  prevalence > 0.9
is_ref_rule <- (taxa_var < 2) & (taxa_prev > 0.9)

prev_var_df <- tibble(
  taxon      = colnames(X),
  prevalence = taxa_prev,
  variance   = taxa_var,
  is_ref     = is_ref_rule
)

ggplot(prev_var_df,
       aes(x = prevalence, y = variance)) +
  geom_point(aes(color = is_ref), size = 1.5, alpha = 0.9) +
  geom_vline(xintercept = 0.9, linetype = "dashed") +
  geom_hline(yintercept = 2,   linetype = "dashed") +
  scale_color_manual(
    values = c(`FALSE` = "darkgrey", `TRUE` = "red"),
    labels = c(`FALSE` = "test", `TRUE` = "reference"),
    name   = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    title = "Reference selection rule at genus level",
    x     = "Prevalence (proportion of samples with count > 0)",
    y     = "Variance of log((x + 1)/sum(x))"
  ) +
  theme_bw() +
  theme(
    legend.position  = "right",
    panel.grid.minor = element_blank()
  )
```

```{r dacomp_prev_var_with_refs, fig.width=6, fig.height=4, dpi=200}
# X_counts: samples x taxa used in dacomp
X <- X_counts
n_samples <- nrow(X)

## 1) Prevalence: proportion of samples with count > 0
taxa_prev <- colSums(X > 0) / n_samples

## 2) Variance of log((x + 1)/sum(x)) across samples
X_logrel <- log(
  sweep(X + 1, 1, rowSums(X), FUN = "/")
)
taxa_var <- apply(X_logrel, 2, var, na.rm = TRUE)

## 4) dacomp-selected references (indices in columns of X)
dacomp_idx  <- res_refs$selected_references
is_dacomp_ref <- seq_along(colnames(X)) %in% dacomp_idx

## 5) Combine into a single factor for coloring
prev_var_df <- tibble(
  taxon        = colnames(X),
  prevalence   = taxa_prev,
  variance     = taxa_var,
  dacomp_ref   = is_dacomp_ref,
  ref_type     = case_when(
    dacomp_ref              ~ "reference",
    TRUE                    ~ "test"
  )
)

ggplot(prev_var_df,
       aes(x = prevalence, y = variance)) +
  geom_point(aes(color = ref_type), size = 1.6, alpha = 0.9) +
  scale_color_manual(
    values = c(
      "test"                = "darkgrey",
      "reference"     = "forestgreen"
    ),
    name = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  #geom_vline(xintercept = 0.9, linetype = "dashed") +
  #geom_hline(yintercept = 2,   linetype = "dashed") +
  labs(
    title = "Prevalence–variance and reference selection (genus level)",
    x     = "Prevalence (proportion of samples with count > 0)",
    y     = "Variance of log((x + 1)/sum(x))"
  ) +
  theme_bw() +
  theme(
    legend.position  = "right",
    panel.grid.minor = element_blank()
  )
```

```{r dacomp_prev_vs_score, fig.width=6, fig.height=4, dpi=200}
X <- X_counts
n_samples <- nrow(X)

# Prevalence (same definition as before)
taxa_prev <- colSums(X > 0) / n_samples

df_prev_score <- tibble::tibble(
  taxon      = colnames(X),
  prevalence = taxa_prev,
  score      = res_refs$scores,
  is_ref     = seq_along(colnames(X)) %in% res_refs$selected_references
)

## taxa to annotate
taxa_to_annotate <- df_prev_score %>%
  dplyr::filter(is_ref & prevalence >= 0.25) %>%
  # dplyr::arrange(dplyr::desc(score)) %>%
  # dplyr::slice(1:5) %>%
  dplyr::pull(taxon)

ggplot(df_prev_score,
       aes(x = prevalence, y = score)) +
  geom_point(aes(color = is_ref), size = 1.7, alpha = 0.9) +
  scale_color_manual(
    values = c(`FALSE` = "darkgrey", `TRUE` = "forestgreen"),
    labels = c(`FALSE` = "test", `TRUE` = "reference"),
    name   = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    title = "Prevalence vs dacomp score (genus level)",
    x     = "Prevalence (proportion of samples with count > 0)",
    y     = "dacomp median SD score"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position  = "right"
  ) +
  ggrepel::geom_text_repel(
    data = subset(df_prev_score, taxon %in% taxa_to_annotate),
    aes(label = taxon),
    size          = 3,
    box.padding   = 0.3,
    point.padding = 0.5,
    min.segment.length = 0,
    max.overlaps  = Inf,
    segment.alpha = 0.6
  )

```

### Run dacomp test

```{r dacomp_test}
# DS-FDR target
q_DSFDR <- 0.1

res_dacomp <- dacomp.test(
  X                  = X_counts,
  y                  = y_groups,
  ind_reference_taxa = res_refs,
  test               = DACOMP.TEST.NAME.WILCOXON,
  nr_perm            = 1000,
  q                  = q_DSFDR,
  verbose            = TRUE,
  return_rarefied_values = TRUE,
  Test_All           = TRUE
)

# Taxon IDs (columns of X_counts are taxa)
taxa_ids <- colnames(X_counts)

# Create a tibble with p-values and DS-FDR calls
res_dacomp_tbl <- tibble(
  taxon        = taxa_ids,
  stat_dacomp  = res_dacomp$stats_matrix[1, ],
  p_dacomp     = res_dacomp$p.values.test,
  p_dacomp_adj = res_dacomp$p.values.test.adjusted,
  is_reference = seq_along(taxa_ids) %in% res_refs$selected_references
)
```

## permApprox - robust ftr

```{r}
stats_mat  <- res_dacomp$stats_matrix
obs_stats  <- stats_mat[1, ]        # length = number of taxa
perm_stats <- stats_mat[-1, , drop = FALSE]  # nr_perm x taxa matrix
colnames(perm_stats) <- res_dacomp_tbl$taxon
dim(stats_mat)
```

### permApprox – unconstrained GPD

```{r permApprox_unconstr_ftr}
fname_pApp_unconstr <- path_target("permapprox_unconstr_ftr.rds")

if (!file.exists(fname_pApp_unconstr) || refit_permApprox) {
  
  gpd_ctrl_unconstr <- make_gpd_ctrl(
    fit_method    = "lme",
    include_obs   = FALSE,
    constraint    = "unconstrained",
    thresh_method = "rob_ftr",
    exceed0       = 0.25,
    exceed0_min   = 250,
    exceed_min    = 10,
    thresh_step   = 10
  )
  
  papprox_unconstr_ftr <- perm_approx(
    obs_stats     = obs_stats,
    perm_stats    = perm_stats,     # dim: n_perm x n_taxa
    alternative   = "two_sided",
    approx_thresh = 0.1,
    method        = "gpd",
    gpd_ctrl      = gpd_ctrl_unconstr,
    adjust_method = "BH",
    cores         = 1,
    verbose       = TRUE
  )
  
  saveRDS(papprox_unconstr_ftr, file = fname_pApp_unconstr)
} else {
  papprox_unconstr_ftr <- readRDS(fname_pApp_unconstr)
}

```

### permApprox – constrained GPD (SLLS)

```{r permApprox_constr_ftr}
## CONSTRAINED GPD FIT (support_at_obs + SLLS)
fname_pApp_constr <- path_target("permapprox_constr_ftr.rds")

if (!file.exists(fname_pApp_constr) || refit_permApprox) {
  
  sampsize <- min(nrow(counts_g1), nrow(counts_g2))  # for SLLS rule
  
  gpd_ctrl_constr <- make_gpd_ctrl(
    fit_method    = "lme",
    include_obs   = FALSE,
    constraint    = "support_at_obs",
    sample_size   = sampsize,
    thresh_method = "rob_ftr",
    exceed0       = 0.25,
    exceed0_min   = 250,
    exceed_min    = 10,
    thresh_step   = 10
  )
  
  papprox_constr_ftr <- perm_approx(
    obs_stats     = obs_stats,
    perm_stats    = perm_stats,     # dim: n_perm x n_taxa
    alternative   = "two_sided",
    approx_thresh = 0.1,
    method        = "gpd",
    gpd_ctrl      = gpd_ctrl_constr,
    adjust_method = "BH",
    cores         = 1,
    verbose       = TRUE
  )
  
  saveRDS(papprox_constr_ftr, file = fname_pApp_constr)
} else {
  papprox_constr_ftr <- readRDS(fname_pApp_constr)
}
```

### Combine dacomp and permApprox results

```{r}
## Add permApprox p-values (empirical + GPD-based)
res_tbl_ftr <- res_dacomp_tbl %>%
  mutate(
    p_dacomp_adj       = NULL,
    p_emp_pApp         = papprox_unconstr_ftr$p_empirical,
    p_perm_pApp_uncon  = papprox_unconstr_ftr$p_unadjusted,
    method_pApp_uncon  = papprox_unconstr_ftr$method_used,
    status_pApp_uncon  = papprox_unconstr_ftr$fit_result$status,
    p_perm_pApp_constr = papprox_constr_ftr$p_unadjusted,
    method_pApp_constr = papprox_constr_ftr$method_used
  )

res_tbl_adj_ftr <- res_dacomp_tbl %>%
  mutate(
    p_dacomp           = ifelse(is.infinite(p_dacomp_adj), 1, p_dacomp_adj),
    p_dacomp_adj       = NULL,
    p_emp_pApp         = papprox_unconstr_ftr$p_empirical,
    p_perm_pApp_uncon  = papprox_unconstr_ftr$p_values,
    method_pApp_uncon  = papprox_unconstr_ftr$method_used,
    p_perm_pApp_constr = papprox_constr_ftr$p_values,
    method_pApp_constr = papprox_constr_ftr$method_used
  )

## log2 fold-change
rel_counts <- X_counts / rowSums(X_counts)
group_means <- aggregate(rel_counts,
                         by = list(group = y_groups),
                         FUN = mean)
mean_g1 <- as.numeric(group_means[group_means$group == label1, -1])
mean_g2 <- as.numeric(group_means[group_means$group == label2, -1])
eps <- 1e-6

res_tbl_ftr <- res_tbl_ftr %>%
  mutate(
    mean_rel_g1 = mean_g1,
    mean_rel_g2 = mean_g2,
    log2_fc     = log2((mean_g2 + eps) / (mean_g1 + eps))
  )

res_tbl_adj_ftr <- res_tbl_adj_ftr %>%
  mutate(
    mean_rel_g1 = mean_g1,
    mean_rel_g2 = mean_g2,
    log2_fc     = log2((mean_g2 + eps) / (mean_g1 + eps))
  )

res_tbl_ftr <- res_tbl_ftr %>%
  arrange(p_perm_pApp_constr)

res_tbl_adj_ftr <- res_tbl_adj_ftr %>%
  arrange(p_perm_pApp_constr)
```

### Unadjusted results

```{r }
# Show top hits
res_tbl_ftr %>%
  slice_head(n = 20) %>%
  kable(digits = 4) %>%
  kable_styling(full_width = FALSE)
```

### Adjusted results

```{r }
# Show top hits
res_tbl_adj_ftr %>%
  slice_head(n = 20) %>%
  kable(digits = 4) %>%
  kable_styling(full_width = FALSE)
```

## permApprox - 250 exceedances

```{r}
stats_mat  <- res_dacomp$stats_matrix
obs_stats  <- stats_mat[1, ]        # length = number of taxa
perm_stats <- stats_mat[-1, , drop = FALSE]  # nr_perm x taxa matrix
colnames(perm_stats) <- res_dacomp_tbl$taxon
dim(stats_mat)
```

### permApprox – unconstrained GPD

```{r permApprox_unconstr_250}
fname_pApp_unconstr <- path_target("permapprox_unconstr_250.rds")

if (!file.exists(fname_pApp_unconstr) || refit_permApprox) {
  
  gpd_ctrl_unconstr <- make_gpd_ctrl(
    fit_method    = "lme",
    include_obs   = FALSE,
    constraint    = "unconstrained",
    thresh_method = "fix",
    exceed0       = 250,
    exceed0_min   = 250,
    exceed_min    = 10,
    thresh_step   = 10
  )
  
  papprox_unconstr <- perm_approx(
    obs_stats     = obs_stats,
    perm_stats    = perm_stats,     # dim: n_perm x n_taxa
    alternative   = "two_sided",
    approx_thresh = 0.1,
    method        = "gpd",
    gpd_ctrl      = gpd_ctrl_unconstr,
    adjust_method = "BH",
    cores         = 1,
    verbose       = TRUE
  )
  
  saveRDS(papprox_unconstr, file = fname_pApp_unconstr)
} else {
  papprox_unconstr <- readRDS(fname_pApp_unconstr)
}

```

### permApprox – constrained GPD (SLLS)

```{r permApprox_constr_250}
## CONSTRAINED GPD FIT (support_at_obs + SLLS)
fname_pApp_constr <- path_target("permapprox_constr_250.rds")

if (!file.exists(fname_pApp_constr) || refit_permApprox) {
  
  sampsize <- min(nrow(counts_g1), nrow(counts_g2))  # for SLLS rule
  
  gpd_ctrl_constr <- make_gpd_ctrl(
    fit_method    = "lme",
    include_obs   = FALSE,
    constraint    = "support_at_obs",
    sample_size   = sampsize,
    thresh_method = "fix",
    exceed0       = 250,
    exceed0_min   = 250,
    exceed_min    = 10,
    thresh_step   = 10
  )
  
  papprox_constr <- perm_approx(
    obs_stats     = obs_stats,
    perm_stats    = perm_stats,     # dim: n_perm x n_taxa
    alternative   = "two_sided",
    approx_thresh = 0.1,
    method        = "gpd",
    gpd_ctrl      = gpd_ctrl_constr,
    adjust_method = "BH",
    cores         = 1,
    verbose       = TRUE
  )
  
  saveRDS(papprox_constr, file = fname_pApp_constr)
} else {
  papprox_constr <- readRDS(fname_pApp_constr)
}
```

### Combine dacomp and permApprox results

```{r}
## Add permApprox p-values (empirical + GPD-based)
res_tbl <- res_dacomp_tbl %>%
  mutate(
    p_dacomp_adj       = NULL,
    p_emp_pApp         = papprox_unconstr$p_empirical,
    p_perm_pApp_uncon  = papprox_unconstr$p_unadjusted,
    method_pApp_uncon  = papprox_unconstr$method_used,
    status_pApp_uncon  = papprox_unconstr$fit_result$status,
    gpd_shape_uncon    = papprox_unconstr$fit_result$shape,
    p_perm_pApp_constr = papprox_constr$p_unadjusted,
    method_pApp_constr = papprox_constr$method_used,
    gpd_shape_constr   = papprox_constr$fit_result$shape
  )

res_tbl_adj <- res_dacomp_tbl %>%
  mutate(
    p_dacomp           = ifelse(is.infinite(p_dacomp_adj), 1, p_dacomp_adj),
    p_dacomp_adj       = NULL,
    p_emp_pApp         = papprox_unconstr$p_empirical,
    p_perm_pApp_uncon  = papprox_unconstr$p_values,
    method_pApp_uncon  = papprox_unconstr$method_used,
    gpd_shape_uncon    = papprox_unconstr$fit_result$shape,
    p_perm_pApp_constr = papprox_constr$p_values,
    method_pApp_constr = papprox_constr$method_used,
    gpd_shape_constr   = papprox_constr$fit_result$shape
  )

## log2 fold-change
rel_counts <- X_counts / rowSums(X_counts)
group_means <- aggregate(rel_counts,
                         by = list(group = y_groups),
                         FUN = mean)
mean_g1 <- as.numeric(group_means[group_means$group == label1, -1])
mean_g2 <- as.numeric(group_means[group_means$group == label2, -1])
eps <- 1e-6

res_tbl <- res_tbl %>%
  mutate(
    mean_rel_g1 = mean_g1,
    mean_rel_g2 = mean_g2,
    log2_fc     = log2((mean_g2 + eps) / (mean_g1 + eps))
  )

res_tbl_adj <- res_tbl_adj %>%
  mutate(
    mean_rel_g1 = mean_g1,
    mean_rel_g2 = mean_g2,
    log2_fc     = log2((mean_g2 + eps) / (mean_g1 + eps))
  )

res_tbl <- res_tbl %>%
  arrange(p_perm_pApp_constr)

res_tbl_adj <- res_tbl_adj %>%
  arrange(p_perm_pApp_constr)
```

### Unadjusted results

```{r }
# Show top hits
res_tbl %>%
  slice_head(n = 20) %>%
  kable(digits = 4) %>%
  kable_styling(full_width = FALSE)
```

### Adjusted results

```{r }
# Show top hits
res_tbl_adj %>%
  slice_head(n = 20) %>%
  kable(digits = 4) %>%
  kable_styling(full_width = FALSE)
```


## Plots

### P-values vs test statistics

```{r}
## Build long df with 3 methods: dacomp / permApprox U / permApprox C
prepare_plot_three_df_dacomp <- function(res_df,
                                         stat_col  = "stat_dacomp",
                                         p_dac_col = "p_dacomp",
                                         p_u_col   = "p_perm_pApp_uncon",
                                         p_c_col   = "p_perm_pApp_constr") {
  
  need <- c("taxon", stat_col, p_dac_col, p_u_col, p_c_col,
            "method_pApp_uncon", "method_pApp_constr")
  stopifnot(all(need %in% names(res_df)))
  
  dd_dac <- res_df %>%
    transmute(
      taxon,
      method_key        = "dacomp",
      method_label      = "dacomp",
      stat              = .data[[stat_col]],
      pval              = .data[[p_dac_col]],
      ptype             = "empirical"
    )
  
  dd_u <- res_df %>%
    transmute(
      taxon,
      method_key        = "permApprox_unconstrained",
      method_label      = "dacomp + permApprox(U)",
      stat              = .data[[stat_col]],
      pval              = .data[[p_u_col]],
      ptype             = if_else(method_pApp_uncon == "empirical",
                                  "empirical", "GPD fit")
    )
  
  dd_c <- res_df %>%
    transmute(
      taxon,
      method_key        = "permApprox_constrained",
      method_label      = "dacomp + permApprox(C)",
      stat              = .data[[stat_col]],
      pval              = .data[[p_c_col]],
      ptype             = if_else(method_pApp_constr == "empirical",
                                  "empirical", "GPD fit")
    )
  
  bind_rows(dd_dac, dd_u, dd_c)
}
```

```{r}
plot_pvals_three_methods_dacomp <- function(df_long,
                                            alpha_level = 0.05,
                                            zero_repl   = 1e-100,
                                            show_zero_label = TRUE,
                                            log_breaks  = c(1, 1e-2, 1e-5, 1e-10),
                                            annotate_taxa = NULL,
                                            annotate_size = 3,
                                            annotate_methods = NULL,
                                            box_padding = 0.2,
                                            force = 1.4,
                                            methods = c("dacomp",
                                                        "permApprox_unconstrained",
                                                        "permApprox_constrained")) {
  
  if (is.null(annotate_methods)) {
    annotate_methods <- methods
  }
  
  dd <- df_long %>%
    filter(method_key %in% methods) %>%
    mutate(method_key = factor(method_key, levels = methods))
  
  if (nrow(dd) == 0L) {
    return(ggplot() +
             annotate("text", x = 0.5, y = 0.5,
                      label = "No points.", size = 5) +
             theme_void())
  }
  
  dd <- dd %>%
    mutate(
      pval_plot = case_when(
        is.na(pval)      ~ NA_real_,
        pval == 0        ~ zero_repl,
        pval < zero_repl ~ NA_real_,
        TRUE             ~ pval
      )
    )
  
  base_labs <- dd %>%
    group_by(method_key) %>%
    summarise(lbl = dplyr::first(method_label), .groups = "drop")
  
  zero_tbl <- dd %>%
    group_by(method_key) %>%
    summarise(
      n_total = sum(!is.na(pval)),
      n_zero  = sum(pval == 0, na.rm = TRUE),
      .groups = "drop"
    )
  
  lab_map <- setNames(
    paste0(
      base_labs$lbl#,
      # "\n(zeros: ",
      # zero_tbl$n_zero[match(base_labs$method_key, zero_tbl$method_key)],
      # " / ",
      # zero_tbl$n_total[match(base_labs$method_key, zero_tbl$method_key)],
      # ")"
    ),
    as.character(base_labs$method_key)
  )
  
  log_breaks <- sort(unique(log_breaks))
  log_breaks <- sort(unique(c(log_breaks, zero_repl)))
  lower_limit <- min(log_breaks)
  
  label_fun <- function(breaks) {
    out <- breaks
    
    # scientific labels for "normal" values
    out[breaks > 1e-5] <- scales::label_scientific()(breaks[breaks > 1e-5])
    
    # zero-replacement labeling is optional
    if (show_zero_label) {
      out[breaks <= zero_repl * 10] <- "0"
    } else {
      # label with the scientific value instead of "0"
      out[breaks <= zero_repl * 10] <- scales::label_scientific()(breaks[breaks <= zero_repl * 10])
    }
    out
  }
  
  p <- ggplot(dd, aes(x = stat, y = pval_plot)) +
    geom_point(aes(fill = ptype),
               shape = 21, size = 2.5, alpha = 0.7,
               color = "gray10", stroke = 0.2) +
    geom_hline(yintercept = alpha_level, linetype = "dashed") +
    facet_wrap(~ method_key, nrow = 1, labeller = as_labeller(lab_map)) +
    scale_y_log10(breaks = log_breaks, labels = label_fun) +
    coord_cartesian(ylim = c(lower_limit, 1)) +
    scale_fill_manual(values = c("GPD fit" = "#F8766D", "empirical" = "gray50"),
                      name = "p-value type") +
    labs(
      x = "dacomp test statistic",
      y = expression(hat(p))
    ) +
    theme_minimal() +
    theme(
      axis.text       = element_text(size = 15),
      axis.title      = element_text(size = 16),
      strip.text      = element_text(size = 15),
      legend.title    = element_text(size = 16),
      legend.text     = element_text(size = 16),
      legend.position    = "bottom",
      panel.spacing      = unit(2, "lines"),
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_line(color = "gray80")
    )
  
  ## ---- Optional annotation of selected genera (ONLY permApprox panel) --
  if (!is.null(annotate_taxa)) {
    ann_df <- dd %>%
      filter(
        method_key %in% annotate_methods,# "permApprox_constrained", 
        taxon %in% annotate_taxa,
        !is.na(pval_plot)
      )
    
    if (nrow(ann_df) > 0) {
      p <- p +
        ggrepel::geom_text_repel(
          data              = ann_df,
          aes(label = taxon),
          size              = annotate_size,
          min.segment.length = 0,
          max.overlaps      = Inf,
          box.padding = box_padding,
          force = force,
          segment.alpha     = 0.5
        )
    }
  }
  
  p
}
```

#### Threshold method: rob_ftr

```{r pvals_vs_tstats_ftr, fig.width=10, fig.height=5, dpi=200}
## Three-panel p-value plot
df_three <- prepare_plot_three_df_dacomp(res_tbl_ftr)

plot_pvals_three_methods_dacomp(
  df_three,
  zero_repl  = 1e-40,
  log_breaks = c(1e-3, 1e-10, 1e-20, 1e-30, 1e-40)
)
```

For Lactobacillus, no theshold could be defined so that the empirical p-values is used.

#### 250 exceedances

```{r pvals_vs_tstats_three, fig.width=10, fig.height=5, dpi=200}
## Three-panel p-value plot
df_three <- prepare_plot_three_df_dacomp(res_tbl)

taxa_to_annotate <- df_three %>% 
  filter(method_key == "permApprox_constrained") %>%
  arrange(pval) %>% slice(1:8) %>% pull(taxon)

plot_pvals_three_methods_dacomp(
  df_three,
  zero_repl  = 1e-22,
  log_breaks = c(1e-3, 1e-5, 1e-10, 1e-15, 1e-20),
  annotate_size = 3.5,
  force = 2,
  annotate_taxa = taxa_to_annotate,
  annotate_methods = c("permApprox_constrained", "permApprox_unconstrained")
)
```

```{r pvals_vs_tstats_two, fig.width=8, fig.height=5, dpi=200}
## Three-panel p-value plot
df_three <- prepare_plot_three_df_dacomp(res_tbl)

taxa_to_annotate <- df_three %>% 
  filter(method_key == "permApprox_constrained") %>%
  arrange(pval) %>% slice(1:8) %>% pull(taxon)

plot_pvals_three_methods_dacomp(
  df_three,
  alpha_level = 0.01,
  show_zero_label = FALSE,
  zero_repl  = 1e-20,
  log_breaks = c(1e-3, 1e-5, 1e-10, 1e-15, 1e-20),
  annotate_size = 3.5,
  annotate_taxa = taxa_to_annotate,
  box_padding = 0.5,
  annotate_methods = c("permApprox_constrained"),
  methods =c("dacomp", "permApprox_constrained")
)
```

#### 250 exceedances; adjusted p-values

```{r pvals_vs_tstats_adj_three, fig.width=10, fig.height=5, dpi=200}
## Three-panel p-value plot
df_three <- prepare_plot_three_df_dacomp(res_tbl_adj)

taxa_to_annotate <- df_three %>% 
  filter(method_key == "permApprox_constrained") %>%
  arrange(pval) %>% slice(1:8) %>% pull(taxon)

plot_pvals_three_methods_dacomp(
  df_three,
  alpha_level = 0.01,
  zero_repl  = 1e-22,
  log_breaks = c(1e-3, 1e-5, 1e-10, 1e-15, 1e-20),
  annotate_size = 3.5,
  annotate_taxa = taxa_to_annotate,
  box_padding = 0.5,
  annotate_methods = c("permApprox_constrained", "permApprox_unconstrained")
)
```


```{r pvals_vs_tstats_adj_two, fig.width=8, fig.height=5, dpi=200}
## Three-panel p-value plot
df_three <- prepare_plot_three_df_dacomp(res_tbl_adj)

taxa_to_annotate <- df_three %>% 
  filter(method_key == "permApprox_constrained") %>%
  arrange(pval) %>% slice(1:8) %>% pull(taxon)

plot_pvals_three_methods_dacomp(
  df_three,
  alpha_level = 0.01,
  zero_repl  = 1e-20,
  show_zero_label = FALSE,
  log_breaks = c(1e-3, 1e-5, 1e-10, 1e-15, 1e-20),
  annotate_size = 3.5,
  annotate_taxa = taxa_to_annotate,
  box_padding = 0.5,
  annotate_methods = c("permApprox_constrained"),
  methods =c("dacomp", "permApprox_constrained")
)
```

### Volcano plots

```{r}
plot_volcano_three_fc <- function(
    res_df,
    fc_col        = "log2_fc",
    p_dac_col     = "p_dacomp",
    p_u_col       = "p_perm_pApp_uncon",
    p_c_col       = "p_perm_pApp_constr",
    taxon_col     = "taxon",
    alpha_thresh  = 0.01,
    lfc_thresh    = 1,
    annotate_taxa = NULL,        # vector of taxa names to label
    annotate_methods = NULL,     # which methods' panels to annotate
    box_padding   = 0.3,
    point_padding = 0.7,
    force         = 1.4,
    n_label       = 10,          # (optional, if you want to use inside)
    title         = NULL
) {

  ## ------------------------------------------------------------------ ##
  ## 1) Build long data frame: 2 (or 3) methods, same log2FC, different p
  ## ------------------------------------------------------------------ ##
  df_long <- dplyr::bind_rows(
    res_df %>%
      dplyr::transmute(
        method  = "dacomp",
        taxon   = .data[[taxon_col]],
        log2_fc = .data[[fc_col]],
        p       = .data[[p_dac_col]]
      ),
    res_df %>%
      dplyr::transmute(
        method  = "dacomp + permApprox",
        taxon   = .data[[taxon_col]],
        log2_fc = .data[[fc_col]],
        p       = .data[[p_c_col]]
      )
  ) %>%
    dplyr::mutate(
      neglogp_raw = dplyr::if_else(p > 0, -log10(p), NA_real_),
      signif_p    = p <= alpha_thresh & !is.na(p),
      big_lfc     = abs(log2_fc) >= lfc_thresh,
      sig_both    = signif_p,
      direction   = dplyr::case_when(
        sig_both & log2_fc >  0 ~ paste0("Higher in '", label2, "'"),
        sig_both & log2_fc <  0 ~ paste0("Higher in '", label1, "'"),
        TRUE                    ~ "No difference"
      )
    )

  # Fix factor order for facets
  df_long <- df_long %>%
    dplyr::mutate(
      method = factor(
        method,
        levels = c("dacomp",
                   "dacomp + permApprox")  # add "permApprox (unconstr)" if used
      )
    )

  ## ------------------------------------------------------------------ ##
  ## 2) Handle zero p-values → show them at a single “Inf” height
  ## ------------------------------------------------------------------ ##
  has_zero <- any(df_long$p == 0, na.rm = TRUE)
  finite_vals <- df_long$neglogp_raw[is.finite(df_long$neglogp_raw)]
  max_finite  <- if (length(finite_vals) > 0) max(finite_vals) else 0

  if (has_zero) {
    y_top <- max_finite + 1  # above all finite values

    df_long <- df_long %>%
      dplyr::mutate(
        neglogp_plot = dplyr::if_else(p == 0, y_top, neglogp_raw)
      )

    base_breaks <- pretty(c(0, max_finite))
    base_breaks <- base_breaks[base_breaks < y_top]

    breaks_y <- c(base_breaks, y_top)
    labels_y <- ifelse(breaks_y == y_top, "Inf", as.character(breaks_y))

    y_limits <- c(0, y_top)
  } else {
    df_long <- df_long %>%
      dplyr::mutate(neglogp_plot = neglogp_raw)
    breaks_y <- pretty(c(0, max_finite))
    labels_y <- as.character(breaks_y)
    y_limits <- c(0, max_finite)
  }

  ## ------------------------------------------------------------------ ##
  ## 3) Colors for direction
  ## ------------------------------------------------------------------ ##
  dir_colors <- setNames(
    c(col_group1, col_group2, "grey70"),
    c(
      paste0("Higher in '", label1, "'"),
      paste0("Higher in '", label2, "'"),
      "No difference"
    )
  )

  ## ------------------------------------------------------------------ ##
  ## 4) Base volcano plot
  ## ------------------------------------------------------------------ ##
  p <- ggplot(df_long, aes(x = log2_fc, y = neglogp_plot)) +
    geom_point(aes(color = direction), alpha = 0.8, size = 3) +
    facet_wrap(~ method, nrow = 1) +
    geom_hline(
      yintercept = -log10(alpha_thresh),
      linetype   = "dashed",
      color      = "gray50"
    ) +
    geom_vline(
      xintercept = c(-lfc_thresh, lfc_thresh),
      linetype   = "dashed",
      color      = "gray50"
    ) +
    scale_color_manual(values = dir_colors, name = NULL) +
    scale_y_continuous(
      breaks = breaks_y,
      labels = labels_y,
      limits = y_limits
    ) +
    labs(
      title = title,
      x     = "log2 fold-change",
      y     = expression(-log[10](hat(p)))
    ) +
    theme_minimal() +
    theme(
      axis.text       = element_text(size = 14),
      axis.title      = element_text(size = 14),
      strip.text      = element_text(size = 14),
      legend.title    = element_text(size = 14),
      legend.text     = element_text(size = 14),
      legend.position = "bottom",
      panel.spacing   = unit(2, "lines")
    )

  ## ------------------------------------------------------------------ ##
  ## 5) Annotation: choose which methods + which taxa to label
  ## ------------------------------------------------------------------ ##
  if (!is.null(annotate_taxa)) {
    
    # default: annotate all methods in the plot
    if (is.null(annotate_methods)) {
      annotate_methods <- levels(df_long$method)
    }
    
    df_annot <- df_long %>%
      dplyr::filter(
        method %in% annotate_methods,
        taxon %in% annotate_taxa,
        !is.na(neglogp_plot)
      )
    
    if (nrow(df_annot) > 0) {
      p <- p +
        ggrepel::geom_text_repel(
          data          = df_annot,
          aes(label = taxon),
          size          = 3,
          box.padding   = box_padding,
          point.padding = point_padding,
          min.segment.length = unit(0, "lines"),
          segment.alpha = 0.4,
          max.overlaps  = Inf,
          force         = force
        )
    }
  }

  p
}

```


```{r}
res_tbl_adj %>% filter(log2_fc > 1, p_perm_pApp_constr < 0.01)
res_tbl_adj %>% filter(log2_fc < 1, p_perm_pApp_constr < 0.01) %>% arrange(p_perm_pApp_constr)
```

```{r}
res_tbl_adj %>% filter(log2_fc > 1, p_dacomp < 0.01)
res_tbl_adj %>% filter(log2_fc < 1, p_dacomp < 0.01) %>% arrange(p_perm_pApp_constr)
```

#### Unadjusted

```{r volcano_unadjusted, fig.width=10, fig.height=5, dpi=300}
res_tbl %>% 
  #filter(!is_reference) %>%
  plot_volcano_three_fc(
    alpha_thresh = 0.01,
    lfc_thresh   = 0,
    box_padding  = 0.9,
    point_padding = 0.5,
    annotate_taxa = FALSE
  )
```

#### Adjusted

```{r volcano_adjusted, fig.width=10, fig.height=5, dpi=300}
set.seed(1234)

taxa_to_annotate <- res_tbl_adj %>%
  filter(p_perm_pApp_constr < 0.01) %>% 
  pull(taxon)

res_tbl_adj %>% 
  #filter(!is_reference) %>%
  plot_volcano_three_fc(
    alpha_thresh = 0.01,
    lfc_thresh   = 0,
    box_padding  = 0.5,
    point_padding = 0.3,
    force        = 2,
    annotate_taxa = c(taxa_to_annotate, "Bifidobacterium"),
    annotate_methods = c("dacomp + permApprox")
  )
```

```{r volcano_adjusted_allsig, fig.width=10, fig.height=5, dpi=300}
set.seed(1234)

taxa_to_annotate <- res_tbl_adj %>%
  filter(p_perm_pApp_constr < 0.01) %>% 
  pull(taxon)

res_tbl_adj %>% 
  #filter(!is_reference) %>%
  plot_volcano_three_fc(
    alpha_thresh = 0.01,
    lfc_thresh   = 0,
    box_padding  = 0.5,
    point_padding = 0.3,
    force        = 2,
    annotate_taxa = c(taxa_to_annotate),
    annotate_methods = c("dacomp + permApprox")
  )
```


### Relative abundances of significant taxa

```{r abund_sig_permApprox, fig.width=10, fig.height=6, dpi=200}
## Abundance plots for taxa significant by permApprox (BH-adjusted)
alpha_sig <- 0.01
top_n <- 8  # number of taxa to plot

## Select 8 taxa with the smallest adjusted permApprox p-values
top_taxa_df <- res_tbl_adj %>%
  filter(!is.na(p_perm_pApp_constr)) %>%    # just to be safe
  arrange(p_perm_pApp_constr) %>%
  slice_head(n = top_n)

sig_taxa <- top_taxa_df$taxon
sig_taxa
```

```{r counts_rel_abund_top8, fig.width=12, fig.height=6, dpi=200}
## Sample-level meta data
sample_groups <- tibble(
  sample = rownames(X_counts),
  group  = y_groups
)

## Long format for selected taxa (ALL abundances, including zeros)
abund_sig_df <- rel_counts[, sig_taxa, drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_groups, by = "sample") %>%
  pivot_longer(
    cols      = all_of(sig_taxa),
    names_to  = "taxon",
    values_to = "rel_abund"
  )

## Facet order: increasing adjusted permApprox p-value
taxon_order <- top_taxa_df %>%
  arrange(p_perm_pApp_constr) %>%
  pull(taxon)

abund_sig_df <- abund_sig_df %>%
  mutate(
    taxon = factor(taxon, levels = taxon_order),
    group = factor(group, levels = c(label1, label2))
  )

## Data for boxplots: non-zero abundances only
abund_nonzero_df <- abund_sig_df %>%
  filter(rel_abund > 0)

## Small offset for log-scale
eps_abund <- 1e-6

ggplot() +
  ## JITTER: all values, zeros included
  geom_jitter(
    data = abund_sig_df,
    aes(x = group, y = rel_abund + eps_abund, color = group),
    width = 0.15, alpha = 0.5, size = 1.2
  ) +
  ## BOXPLOTS: only non-zero values
  geom_boxplot(
    data = abund_nonzero_df,
    aes(x = group, y = rel_abund + eps_abund, fill = group),
    outlier.shape = NA,
    alpha = 0.8
  ) +
  scale_y_log10(labels = scales::label_scientific()) +
  scale_fill_manual(values = palette_groups) +
  scale_color_manual(values = palette_groups) +
  facet_wrap(~ taxon, scales = "free_y", ncol = 4) +
  labs(
    title = paste0(
      "Relative abundances of top ", length(sig_taxa),
      " taxa ordered by increasing BH-adjusted permApprox p-value"
    ),
    x = NULL,
    y = "Relative abundance"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 14),
    axis.text.x  = element_blank(),
    axis.text.y  = element_text(size = 12),
    axis.title   = element_text(size = 14),
    strip.text   = element_text(size = 12)
  )
```

### dacomp-rarefied taxa counts

```{r}
alpha_sig <- 0.01
top_n     <- 8

top_taxa_df <- res_tbl_adj %>%
  filter(!is.na(p_perm_pApp_constr),
         p_perm_pApp_constr < alpha_sig) %>%
  arrange(p_perm_pApp_constr) %>%
  slice_head(n = top_n)

sig_taxa <- top_taxa_df$taxon
sig_taxa
```


```{r}
## dacomp-normalized (rarefied) counts
dacomp_counts <- res_dacomp$rarefied_counts
dimnames(dacomp_counts) <- dimnames(X_counts)

## Sample-level meta data
sample_groups <- tibble(
  sample = rownames(dacomp_counts),
  group  = y_groups
)

## Long format for selected taxa (all counts, zeros included)
dacomp_long <- dacomp_counts[, sig_taxa, drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  left_join(sample_groups, by = "sample") %>%
  pivot_longer(
    cols      = all_of(sig_taxa),
    names_to  = "taxon",
    values_to = "norm_count"
  )

## Facet order: increasing adjusted permApprox p-value
taxon_order <- top_taxa_df %>%
  arrange(p_perm_pApp_constr) %>%
  pull(taxon)

dacomp_long <- dacomp_long %>%
  mutate(
    taxon = factor(taxon, levels = taxon_order),
    group = factor(group, levels = c(label1, label2))
  )

## Boxplots: non-zero counts only
dacomp_nonzero <- dacomp_long %>%
  filter(norm_count > 0)

## Offset for log10 scale (counts are integers; 0.5 works nicely)
eps_count <- 0.5
```


```{r counts_dacomp_norm_top8, fig.width=12, fig.height=6, dpi=200}
ggplot() +
  # Jitter: all counts (including zeros)
  geom_jitter(
    data = dacomp_long,
    aes(x = group, y = norm_count + eps_count, color = group),
    width = 0.15, alpha = 0.8, size = 1.2
  ) +
  # Boxplots: non-zero counts only
  geom_boxplot(
    data = dacomp_nonzero,
    aes(x = group, y = norm_count + eps_count, fill = group),
    outlier.shape = NA,
    alpha = 0.6
  ) +
    scale_y_log10(
    limits = c(eps_count, max(dacomp_long$norm_count + eps_count) * 1.1),
    breaks = c(eps_count, 1, 10, 100),
    labels = c("0", "1", "10", "100")
  ) +
  scale_fill_manual(values = palette_groups) +
  scale_color_manual(values = palette_groups) +
  facet_wrap(~ taxon, scales = "free_y", ncol = 4) +
  labs(
    title = paste0(
      "dacomp-rarefied counts of top ", length(sig_taxa),
      " taxa ordered by increasing BH-adjusted permApprox p-value"
    ),
    x = NULL,
    y = "dacomp-rarefied counts"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 14),
    axis.text.x  = element_blank(),
    axis.text.y  = element_text(size = 12),
    axis.title   = element_text(size = 14),
    strip.text   = element_text(size = 12)
  )
```
```{r counts_violin_dacomp_norm_top8, fig.width=12, fig.height=6, dpi=200}
library(ggforce)

ggplot() +
  # # Jitter: all counts (including zeros)
  # geom_jitter(
  #   data = dacomp_long,
  #   aes(x = group, y = norm_count + eps_count, color = group),
  #   width = 0.15, alpha = 0.5, size = 1.2
  # ) +
  # Violins: full distribution
  geom_violin(
    data = dacomp_long,
    aes(x = group, y = norm_count + eps_count, fill = group),
    trim = FALSE,
    alpha = 0.6,
    color = NA
  ) +
  # Density-shaped dots
  geom_sina(
    data = dacomp_long,
    aes(x = group, y = norm_count + eps_count, color = "black"),
    color = "black",
    maxwidth = 0.7,          # width of the “violin” of dots
    alpha   = 0.7,
    size    = 0.5
  ) +
  # # Overlaid boxplot to visualize medians (optional, looks very good)
  # geom_boxplot(
  #   data = dacomp_long,
  #   aes(x = group, y = norm_count + eps_count),
  #   width = 0.2,
  #   outlier.shape = NA,
  #   alpha = 0.5,
  #   color = "black",
  #   fill = "white"
  # ) +
  scale_y_log10(
    limits = c(eps_count, max(dacomp_long$norm_count + eps_count) * 1.1),
    breaks = c(eps_count, 1, 10, 100),
    labels = c("0", "1", "10", "100")
  ) +
  scale_fill_manual(values = palette_groups) +
  scale_color_manual(values = palette_groups) +
  facet_wrap(~ taxon, scales = "free_y", ncol = 4) +
  labs(
    # title = paste0(
    #   "dacomp-rarefied counts of top ", length(sig_taxa),
    #   " taxa ordered by increasing BH-adjusted permApprox p-value"
    # ),
    x = NULL,
    y = "dacomp-rarefied counts"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 14),
    axis.text.x  = element_blank(),
    axis.text.y  = element_text(size = 12),
    axis.title   = element_text(size = 14),
    strip.text   = element_text(size = 12)
  )

```


### Tables

#### Unadjusted p-values

```{r table_top_effects, message = FALSE}
# Number of top effects to display
n_top <- 15

sci <- function(x) {
  ifelse(is.na(x), NA, formatC(x, format = "e", digits = 2))
}

top_effects_tbl <- res_tbl %>%
  # drop reference taxa if desired
  #filter(!is_reference) %>%
  arrange(p_perm_pApp_constr) %>%
  slice_head(n = n_top) %>%
  mutate(
    p_dacomp_sci           = sci(p_dacomp),
    p_permApprox_uncon_sci = sci(p_perm_pApp_uncon),
    p_permApprox_constr_sci= sci(p_perm_pApp_constr)
  ) %>%
  transmute(
    taxon,
    stat_dacomp,
    log2_fc,
    `p_dacomp`              = p_dacomp_sci,
    `p_permApprox(U)`       = p_permApprox_uncon_sci,
    `p_permApprox`       = p_permApprox_constr_sci
  )

top_effects_tbl %>%
  kable(
    escape = TRUE,
    caption = paste0(
      "Top ", n_top, " genera sorted by permApprox p-values."),
    ) %>%
  kable_styling(full_width = FALSE)

```

```{r table_top_effects_latex, results = "asis", message = FALSE}
# Number of taxa to show
n_top <- 20 

# Helper: scientific notation as character
sci <- function(x, digits = 2) {
  ifelse(is.na(x), NA, formatC(x, format = "e", digits = digits))
}

top_tbl <- res_tbl %>%
  #filter(!is_reference) %>%               # exclude dacomp reference taxa
  arrange(p_perm_pApp_constr) %>%         # sort by permApprox
  slice_head(n = n_top) %>%
  mutate(
    stat_dacomp       = round(stat_dacomp, 3),
    log2_fc           = round(log2_fc, 2),
    p_dacomp_sci      = sci(p_dacomp),
    p_pApp_U_sci      = sci(p_perm_pApp_uncon),
    p_pApp_C_sci      = sci(p_perm_pApp_constr),
    gpd_shape_U       = round(gpd_shape_uncon, 3),
    gpd_shape_C       = round(gpd_shape_constr, 3)
  ) %>%
  transmute(
    Genus                         = taxon,
    `dacomp stat`                 = stat_dacomp,
    `log2 FC`                     = log2_fc,
    `$p_{\\text{dacomp}}$`        = p_dacomp_sci,
    `$p_{\\text{permApprox(U)}}$` = p_pApp_U_sci,
    `$p_{\\text{permApprox(C)}}$` = p_pApp_C_sci,
    `$\\hat{\\xi}_{(U)}$`         = gpd_shape_U,
    `$\\hat{\\xi}_{(C)}$`         = gpd_shape_C
  )

# Create LaTeX-ready table
tab_latex <- kable(
  top_tbl,
  format   = "latex",
  booktabs = TRUE,
  escape   = FALSE,
  caption  = paste0(
    "p-values of the top ", n_top,
    " genera sorted by permApprox p-values. "
  ),
  label    = "tab:permApprox_top_effects"
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size     = 8
  )

print(tab_latex)

# Optional: export as a standalone .tex file
save_kable(tab_latex, path_target("permApprox_top_effects_table.tex"))
```

#### Adjusted p-values

```{r table_top_effects_adj, message = FALSE}
# Number of top effects to display
n_top <- 15

sci <- function(x) {
  ifelse(is.na(x), NA, formatC(x, format = "e", digits = 2))
}

top_effects_tbl <- res_tbl_adj %>%
  # drop reference taxa if desired
  #filter(!is_reference) %>%
  arrange(p_perm_pApp_constr) %>%
  slice_head(n = n_top) %>%
  mutate(
    p_dacomp_sci           = sci(p_dacomp),
    p_permApprox_uncon_sci = sci(p_perm_pApp_uncon),
    p_permApprox_constr_sci= sci(p_perm_pApp_constr)
  ) %>%
  transmute(
    taxon,
    stat_dacomp,
    log2_fc,
    `p_dacomp`            = p_dacomp_sci,
    `p_permApprox(U)`     = p_permApprox_uncon_sci,
    `p_permApprox`        = p_permApprox_constr_sci
  )

top_effects_tbl %>%
  kable(
    escape = TRUE,
    caption = paste0(
      "Top ", n_top, " genera sorted by permApprox p-values."),
    ) %>%
  kable_styling(full_width = FALSE)

```

```{r table_top_effects_adj_latex, results = "asis", message = FALSE}
# Number of taxa to show
n_top <- 20 

# Helper: scientific notation as character
sci <- function(x, digits = 2) {
  ifelse(is.na(x), NA, formatC(x, format = "e", digits = digits))
}

top_tbl <- res_tbl_adj %>%
  #filter(!is_reference) %>%               # exclude dacomp reference taxa
  arrange(p_perm_pApp_constr) %>%         # sort by permApprox
  slice_head(n = n_top) %>%
  mutate(
    stat_dacomp       = round(stat_dacomp, 3),
    log2_fc           = round(log2_fc, 2),
    p_dacomp_sci      = sci(p_dacomp),
    p_pApp_U_sci      = sci(p_perm_pApp_uncon),
    p_pApp_C_sci      = sci(p_perm_pApp_constr),
    gpd_shape_U       = round(gpd_shape_uncon, 3),
    gpd_shape_C       = round(gpd_shape_constr, 3)
  ) %>%
  transmute(
    Genus                        = taxon,
    `dacomp stat`                = stat_dacomp,
    `log2 FC`                    = log2_fc,
    `$p_{\\text{dacomp}}$`        = p_dacomp_sci,
    `$p_{\\text{permApprox(U)}}$` = p_pApp_U_sci,
    `$p_{\\text{permApprox(C)}}$` = p_pApp_C_sci,
    `$\\hat{\\xi}_{(U)}$`         = gpd_shape_U,
    `$\\hat{\\xi}_{(C)}$`         = gpd_shape_C
  )

# Create LaTeX-ready table
tab_latex <- kable(
  top_tbl,
  format   = "latex",
  booktabs = TRUE,
  escape   = FALSE,
  caption  = paste0(
    "p-values of the top ", n_top,
    " genera sorted by BH-adjusted permApprox p-values."
  ),
  label    = "tab:permApprox_top_effects"
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size     = 8
  )

print(tab_latex)

# Optional: export as a standalone .tex file
save_kable(tab_latex, path_target("permApprox_top_effects_adj_table.tex"))
```

