---
title: "Differential abundance analysis with `dacomp` - EBF vs non-EBF"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
always_allow_html: true
params:
  name: "02_micro_diff_abund" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "bb9f8d89-5eb6-408d-acfc-729e640abec2")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 10, 
  fig.height = 6, 
  dpi = 200,
  fig.path = paste0('figures/', params, "/app_dacomp_"),
  echo = FALSE)

refit_permApprox <- FALSE
```

This notebook demonstrates how to perform **differential abundance analysis** of microbiome count data using the **`dacomp`** package.

We use genus-level profiles from the **PASTURE** cohort (2-month samples). After basic prevalence filtering and defining two comparison groups (here, breastfeeding categories), we visualize the data (heatmaps, densities, histograms, ECDFs) and then apply **dacomp**:

1. **Select a stable reference set of non-DA taxa**,  
2. **Test all remaining taxa for differential abundance** using a Wilcoxon-based two-sample test with permutation p-values and DS-FDR correction.

The example uses the **raw count data** for inference (as required by dacomp), while relative abundances are used only for visualization.

In this notebook, we compare two breast feeding groups: 'EBF' (exclusive breast feeding until age 2 months) vs. 'non-EBF'.

## Global options and packages

```{r packages, message=FALSE, warning=FALSE}
library(conflicted)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(kableExtra)
library(pheatmap)
library(ggrepel)
library(patchwork)

# Differential abundance with dacomp
library(dacomp)

# p-value approximation with permApprox
#remotes::install_github("stefpeschel/permApprox")
library(permApprox)

conflicted::conflicts_prefer(dplyr::count, dplyr::filter, dplyr::select, 
                             stats::var, dplyr::setdiff, dplyr::slice, 
                             base::unname, base::union, base::intersect)
```

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```


## Load data & set parameters

```{r load_data}
bact_phylo <- readRDS(path_source("PASTURE_data", paste0("bact_2m_genus.rds")))
sampdata    <- sample_data(bact_phylo) %>% as("data.frame")
bact_phylo
```

```{r}
# Rename "24_Enterobacteriaceae(F)"
taxtab <- tax_table(bact_phylo)
Genus <- taxtab[, "Genus"]
Genus[Genus == "24_Enterobacteriaceae(F)"] <- "Enterobacteriaceae(F)"
taxtab[, "Genus"] <- Genus

tax_table(bact_phylo) <- taxtab
taxa_names(bact_phylo) <- Genus
```

## Data preprocessing

### Taxa prevalence (including zero counts)

```{r expl_prevalence_hist}
# Total number of samples
n_samples <- nsamples(bact_phylo)

# Compute number of samples where each taxon is observed (abundance > 0)
taxa_prevalence <- apply(otu_table(bact_phylo), 1, function(x) sum(x > 0))

# Build dataframe
prevalence_df <- tibble(
  taxon = names(taxa_prevalence),
  prevalence = taxa_prevalence,
  prevalence_pct = 100 * taxa_prevalence / n_samples
)

# Plot histogram (absolute counts)
ggplot(prevalence_df, aes(x = prevalence)) +
  geom_histogram(binwidth = 5, fill = "#1F78B4", color = "white", alpha = 0.8) +
  labs(
    title = "Taxon prevalence across samples",
    x = "Number of samples with abundance > 0",
    y = "Number of taxa"
  ) +
  theme_minimal()
```

### Prevalence threshold

```{r expl_prevalence_hist_filter}
bact_phylo  <- prune_taxa(taxa_sums(bact_phylo) > 0, bact_phylo)

ggplot(prevalence_df, aes(x = prevalence_pct)) +
  geom_histogram(binwidth = 2, fill = "#1F78B4", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = 1, color = "1% threshold"), 
             linetype = "dashed", linewidth = 0.8) +
  geom_vline(aes(xintercept = 5, color = "5% threshold"), 
             linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(
    name = "Prevalence thresholds",
    values = c("1% threshold" = "orange", "5% threshold" = "red")
  ) +
  labs(
    title = "Taxon prevalence across samples (including zero-abundance taxa)",
    x = "Prevalence (% of samples)",
    y = "Number of taxa"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )
```

### Taxa filtering

We keep only taxa with a non-zero count in at least 1% of the samples.

```{r}
# Minimum prevalence threshold (5% of total samples)
prev_threshold <- 0.01 * nsamples(bact_phylo)

# Compute prevalence per taxon
taxa_prevalence <- apply(otu_table(bact_phylo), 1, function(x) sum(x > 0))

# Filter taxa with prevalence >= threshold
keep_taxa <- taxa_prevalence >= prev_threshold

# Prune the phyloseq object
bact_phylo <- prune_taxa(keep_taxa, bact_phylo)
bact_phylo
```

### Define comparison groups

```{r}
cat("breast_excl_cat1:")
table(sampdata$breast_excl_cat1)

# Build new variable with categories EBF and non-EBF
sampdata <- sampdata %>%
  mutate(BF_cat = case_when(
    breast_excl_cat1 == "0" ~ "non-EBF",
    breast_excl_cat1 == ">=2" ~ "EBF",
    TRUE ~ NA_character_
  ))
sample_data(bact_phylo) <- sampdata
cat("\nNew variable:")
table(sampdata$BF_cat)
```

```{r}
## Global color palette (consistent across all plots)
col_group1 <- "#1f77b4"     # blue
col_group2 <- "darkorange2" # orange

palette_groups <- c("non-EBF" = col_group1, 
                    "EBF"     = col_group2)

# New variable
comp_var <- "BF_cat"

# Extract the vector to split on
var_vec <- sampdata[[comp_var]]
names(var_vec) <- rownames(sampdata)

comp_val1 <- label1 <- "non-EBF"
comp_val2 <- label2 <- "EBF"

# categorical split by values
group1_ids <- rownames(sampdata)[var_vec %in% comp_val1]
group2_ids <- rownames(sampdata)[var_vec %in% comp_val2]

# # make group sizes equal
# set.seed(42)
# min_length <- min(length(group1_ids), length(group2_ids))
# group2_ids <- sample(group2_ids, min_length, replace = FALSE)

# Prune phyloseq objects for each group
bact_g1 <- prune_samples(group1_ids, bact_phylo)
bact_g2 <- prune_samples(group2_ids, bact_phylo)

# Create a tibble for plotting the distribution of comp_var
dist_df <- tibble(
  sample = names(var_vec),
  value  = var_vec,
  group  = case_when(
    sample %in% group1_ids ~ label1,
    sample %in% group2_ids ~ label2,
    TRUE                   ~ "Other"
  )
) %>%
  mutate(group = factor(group, levels = c(label1, "Other", label2)))

# Numeric: histogram of values
p_dist <- ggplot(dist_df, aes(x = group, fill = group)) +
  geom_histogram(bins = 30, color = "black", alpha = .6) +
  scale_fill_manual(values = palette_groups) +
  labs(x = comp_var, y = "Count")
```


```{r expl_group_sizes, fig.width=5, fig.height=5, dpi=200}
# Categorical: bar plot
ggplot(dist_df %>% filter(group != "Other"),  # optional: drop "Other"
                 aes(x = base::as.factor(group), fill = group)) +
  geom_bar(color = "black", alpha = .6, position = "dodge") +
  scale_fill_manual(values = palette_groups) +
  labs(x = "", y = "Sample count") +
  theme_minimal() +
  theme(legend.position="none",
        axis.text  = element_text(size = 12),
        axis.title = element_text(size = 14))
```

```{r}
cat("Samples in group 1:", nsamples(bact_g1), "\n")
cat("Samples in group 2:", nsamples(bact_g2))
```

### Relative abundance transformation

```{r}
## Absolute counts
counts_g1 <- t(as(otu_table(bact_g1), "matrix"))
counts_g2 <- t(as(otu_table(bact_g2), "matrix"))

# Compute relative abundances per sample
abund_rel_g1 <- counts_g1 / rowSums(counts_g1)
abund_rel_g2 <- counts_g2 / rowSums(counts_g2)

# Combine abundance matrices for the two groups
abund_rel_combined <- rbind(abund_rel_g1, abund_rel_g2)

# Assign group labels
sample_labels <- c(rep(label1, nrow(abund_rel_g1)),
                   rep(label2, nrow(abund_rel_g2)))

```


## Data exploration

```{r}
# Top 20 overall taxa
top_taxa <- names(sort(taxa_sums(bact_phylo), decreasing = TRUE))[1:20]

mat1 <- abund_rel_g1[, top_taxa]
mat2 <- abund_rel_g2[, top_taxa]

combined <- rbind(mat1, mat2)
labels   <- c(rep(label2, nrow(mat2)), rep(label1, nrow(mat1)))

plot_df <- as.data.frame(combined) %>%
  rownames_to_column("sample") %>%
  mutate(group = labels) %>%
  pivot_longer(-c(sample, group), names_to = "taxon", values_to = "abundance") %>%
  mutate(taxon_short = stringr::str_trunc(taxon, width = 15, side = "right"))

# Ensure order of facets follows the order in `top_taxa`
top_taxa_short <- stringr::str_trunc(top_taxa[1:6], width = 15, side = "right")

# Set taxon_short as a factor with desired levels
plot_df <- plot_df %>%
  mutate(taxon_short = factor(taxon_short, levels = top_taxa_short))
```

### Library sizes

```{r expl_library_sizes, fig.width=5, fig.height=5, dpi=200}
lib_sizes_df <- tibble(
  sample   = c(rownames(counts_g1), rownames(counts_g2)),
  group    = factor(c(rep(label1, nrow(counts_g1)),
                      rep(label2, nrow(counts_g2))),
                    levels = c(label1, label2)),
  lib_size = c(rowSums(counts_g1),
               rowSums(counts_g2))
)

ggplot(lib_sizes_df, aes(x = group, y = lib_size, fill = group)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1) +
  scale_fill_manual(values = palette_groups) +
  labs(
    #title = "Library sizes by group",
    x = NULL,
    y = "Total read count per sample"
  ) +
  theme_minimal() +
  theme(
    legend.position="none",
    axis.text  = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

```

### Prevalence vs mean relative abundance

```{r expl_prev_mean_scatter, fig.width=6, fig.height=4, dpi=200}
## Prevalence and mean relative abundance within the selected samples
rel_comb <- rbind(abund_rel_g1, abund_rel_g2)

prev_vec <- colSums(rel_comb > 0)
n_sel    <- nrow(rel_comb)

prev_mean_df <- tibble(
  taxon        = colnames(rel_comb),
  prevalence   = prev_vec / n_sel,
  mean_rel_ab  = colMeans(rel_comb)
)

ggplot(prev_mean_df,
       aes(x = prevalence, y = mean_rel_ab)) +
  geom_point(alpha = 0.6) +
  scale_y_log10() +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    #title = "Prevalence vs mean relative abundance",
    x = "Prevalence (proportion of selected samples with > 0 counts)",
    y = "Mean relative abundance (log10 scale)"
  ) +
  theme_minimal() +
  theme(
    legend.position="none",
    axis.text  = element_text(size = 12),
    axis.title = element_text(size = 12)
  )
```


### Alpha diversity

```{r}
# Restrict to the samples used in the dacomp comparison
bact_alpha <- prune_samples(c(group1_ids, group2_ids), bact_phylo)

# Standard alpha diversity estimates from phyloseq
alpha_est <- estimate_richness(
  bact_alpha,
  measures = c("Observed", "Shannon", "InvSimpson")
) %>%
  tibble::rownames_to_column("sample")

# Map sample IDs to group labels used in the dacomp script
alpha_df <- alpha_est %>%
  dplyr::mutate(
    GiniSimpson = 1 - 1 / InvSimpson
  ) %>%
  dplyr::left_join(
    tibble::tibble(
      sample = c(group1_ids, group2_ids),
      group  = c(rep(label1, length(group1_ids)),
                 rep(label2, length(group2_ids)))
    ),
    by = "sample"
  ) %>%
  dplyr::mutate(
    group = factor(group, levels = c(label1, label2))
  )

head(alpha_df)
```

```{r}
plot_alpha_metric <- function(df, metric, ylab,
                              palette = palette_groups,
                              title = NULL,
                              jitter_width = 0.1) {
  
  stopifnot(metric %in% names(df))
  
  # Wilcoxon rank-sum test between the two groups
  wt <- wilcox.test(df[[metric]] ~ df$group)
  pval <- wt$p.value
  
  y_max   <- max(df[[metric]], na.rm = TRUE)
  y_line  <- y_max * 1.05
  y_label <- y_max * 1.10
  
  ggplot(df, aes(x = group, y = .data[[metric]], fill = group)) +
    # Jittered points
    geom_jitter(width = jitter_width,
                size = 1.6,
                alpha = 0.7,
                stroke = 0,
                aes(color = group)) +
    # Boxplot (median + IQR)
    geom_boxplot(alpha = 0.6,
                 width = 0.4,
                 outlier.shape = NA) +
    # Bracket + p-value
    geom_segment(aes(x = 1, xend = 2,
                     y = y_line, yend = y_line),
                 inherit.aes = FALSE) +
    annotate("text",
             x = 1.5, y = y_label, size = 5,
             label = paste0("p = ", signif(pval, 3))) +
    scale_fill_manual(values = palette) +
    scale_color_manual(values = palette) +
    labs(
      x = NULL,
      y = ylab,
      title = title
    ) +
    coord_cartesian(ylim = c(0, y_label * 1.05)) +
    theme_minimal() +
    theme(
      #legend.position = "none",
      axis.text  = element_text(size = 14),
      axis.text.x = element_blank(),
      axis.title.y = element_text(size = 14),
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 14),
      plot.title   = element_text(size = 14, face = "bold"),
      panel.grid.minor = element_blank()
    )
}

```

#### Richness (Observed genera)

```{r expl_alpha_richness}
p_alpha_richness <- plot_alpha_metric(
  alpha_df,
  metric = "Observed",
  ylab   = "Richness (genera)"
  #title  = "Alpha diversity: richness"
)
p_alpha_richness
```

#### Shannon diversity

```{r expl_alpha_shannon}
p_alpha_shannon <- plot_alpha_metric(
  alpha_df,
  metric = "Shannon",
  ylab   = "Shannon index"
  #title  = "Alpha diversity: Shannon"
)
p_alpha_shannon

```

#### Gini–Simpson index

```{r expl_alpha_gini_simpson}
p_alpha_gini <- plot_alpha_metric(
  alpha_df,
  metric = "GiniSimpson",
  ylab   = "Gini–Simpson index"
  #title  = "Alpha diversity: evenness"
)
p_alpha_gini
```

#### Alpha diversity combined

```{r expl_alpha_diversity, fig.width=11, fig.height=6, dpi=200}
p_alpha_richness + p_alpha_shannon + p_alpha_gini +
  plot_layout(guides = 'collect')
```


### Heatmap

```{r heatmap_top20_unclust, fig.width=10, fig.height=5, dpi=200}
# Combine abundance matrices for the two groups
abund_rel_combined <- rbind(abund_rel_g1, abund_rel_g2)

# Assign group labels
sample_labels <- c(rep(label1, nrow(abund_rel_g1)),
                   rep(label2, nrow(abund_rel_g2)))

# Get top 20 most abundant taxa in the *full* dataset
top_taxa <- names(sort(taxa_sums(bact_phylo), decreasing = TRUE))[1:20]

# Select only top taxa from the combined matrix
heatmap_mat <- abund_rel_combined[, top_taxa]

# Annotation data for the columns
annotation_col <- data.frame(group = factor(sample_labels))
rownames(annotation_col) <- rownames(heatmap_mat)

annotation_colors = list(group = palette_groups)

# Heatmap (non-clustered)
pheatmap(
  t(heatmap_mat),  # transpose: taxa as rows
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  show_rownames = TRUE,
  show_colnames = FALSE,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  main = paste("Relative abundances (Top 20 taxa:", label1, "vs", label2, ")")
)
```

```{r heatmap_top20_clust, fig.width=10, fig.height=5, dpi=200}
# Heatmap (clustered)
pheatmap(
  t(heatmap_mat),  # transpose: taxa as rows
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,  show_rownames = TRUE,
  show_colnames = FALSE,
  cluster_rows = FALSE,
  cluster_cols = TRUE,
  main = paste("Relative abundances (Top 20 taxa:", label1, "vs", label2, ")")
)
```

### Mean composition - Phylum level

```{r expl_mean_composition_phylum, fig.width=7, fig.height=5, dpi=200}
## ----- Aggregate genus-level relative abundances to Phylum -----

tax_tab <- as.data.frame(tax_table(bact_phylo))

# Map genus taxa (rows of tax_tab) to Phylum
genus_to_phylum <- tax_tab$Phylum
names(genus_to_phylum) <- rownames(tax_tab)

# Helper to aggregate one group to phylum
aggregate_to_phylum <- function(abund_rel_mat) {
  abund_rel_mat %>%
    as.data.frame() %>%
    mutate(sample = rownames(.)) %>%
    tidyr::pivot_longer(
      cols      = -sample,
      names_to  = "taxon",
      values_to = "rel_abund"
    ) %>%
    dplyr::mutate(phylum = genus_to_phylum[taxon]) %>%
    dplyr::group_by(sample, phylum) %>%
    dplyr::summarise(rel_abund = sum(rel_abund), .groups = "drop") %>%
    tidyr::pivot_wider(
      names_from  = phylum,
      values_from = rel_abund,
      values_fill = 0
    )
}

phylum_mat_g1 <- aggregate_to_phylum(abund_rel_g1)
phylum_mat_g2 <- aggregate_to_phylum(abund_rel_g2)

phylum_g1 <- as.matrix(phylum_mat_g1[, -1, drop = FALSE])
phylum_g2 <- as.matrix(phylum_mat_g2[, -1, drop = FALSE])

## ----- Group-wise mean relative abundances at Phylum level -----

mean_phylum_g1 <- colMeans(phylum_g1)
mean_phylum_g2 <- colMeans(phylum_g2)

mean_phylum_df <- tibble::tibble(
  phylum       = colnames(phylum_g1),
  mean_rel_g1  = mean_phylum_g1,
  mean_rel_g2  = mean_phylum_g2
) %>%
  dplyr::mutate(overall_mean = (mean_rel_g1 + mean_rel_g2) / 2) %>%
  dplyr::arrange(dplyr::desc(overall_mean))

mean_phylum_long <- mean_phylum_df %>%
  dplyr::select(phylum, mean_rel_g1, mean_rel_g2) %>%
  tidyr::pivot_longer(
    cols      = dplyr::starts_with("mean_rel_"),
    names_to  = "group",
    values_to = "rel_abund"
  ) %>%
  dplyr::mutate(
    group = dplyr::recode(
      group,
      "mean_rel_g1" = label1,   # non-EBF
      "mean_rel_g2" = label2    # EBF
    )
  )

## ----- Color palette: 7 phyla, all shown -----

# Okabe–Ito 7-color palette
base_cols <- c(
  "#E69F00", # orange
  "#56B4E9", # light blue
  "#009E73", # green
  "#F0E442", # yellow
  "#0072B2", # blue
  "#D55E00", # red
  "#CC79A7"  # purple
)

phylum_levels <- unique(mean_phylum_long$phylum)
# assumes length(phylum_levels) <= 7
phylum_colors <- setNames(base_cols[seq_along(phylum_levels)], phylum_levels)

## ----- Plot: stacked bar chart at Phylum level -----

ggplot(mean_phylum_long,
       aes(x = group, y = rel_abund, fill = phylum)) +
  geom_bar(stat = "identity", position = "stack", color = "white") +
  scale_fill_manual(values = phylum_colors, name = "Phylum") +
  labs(
    title = "Mean phylum composition",
    x     = NULL,
    y     = "Relative abundance (mean across samples)"
  ) +
  theme_minimal() +
  theme(
    axis.text    = element_text(size = 14),
    axis.title   = element_text(size = 14),
    plot.title   = element_text(size = 15, face = "bold"),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 14)
  )
```

Firmucutes are increased in non-EBF, which is in line with other studies. 

### Mean composition - Genus level

```{r expl_mean_composition_genus, fig.width=7, fig.height=5, dpi=200}
## Mean relative abundances per group (using dacomp samples)

mean_rel_g1 <- colMeans(abund_rel_g1)
mean_rel_g2 <- colMeans(abund_rel_g2)

mean_rel_df <- tibble::tibble(
  taxon       = colnames(abund_rel_g1),
  mean_rel_g1 = mean_rel_g1,
  mean_rel_g2 = mean_rel_g2
) %>%
  dplyr::mutate(
    overall_mean = (mean_rel_g1 + mean_rel_g2) / 2
  ) %>%
  dplyr::arrange(dplyr::desc(overall_mean))

# Number of top genera to show explicitly
K <- 7

top_taxa <- mean_rel_df$taxon[1:K]

mean_rel_long <- mean_rel_df %>%
  dplyr::select(taxon, mean_rel_g1, mean_rel_g2) %>%
  tidyr::pivot_longer(
    cols      = starts_with("mean_rel_"),
    names_to  = "group",
    values_to = "rel_abund"
  ) %>%
  dplyr::mutate(
    group = dplyr::recode(
      group,
      "mean_rel_g1" = label1,
      "mean_rel_g2" = label2
    ),
    taxon_plot = dplyr::if_else(taxon %in% top_taxa, taxon, "Other")
  ) %>%
  dplyr::group_by(group, taxon_plot) %>%
  dplyr::summarise(
    rel_abund = sum(rel_abund),
    .groups   = "drop"
  ) %>%
  dplyr::group_by(group) %>%
  dplyr::mutate(
    rel_abund = rel_abund / sum(rel_abund)
  ) %>%
  dplyr::ungroup()

# Order taxa so that "Other" is last in the legend
ordered_taxa <- c(top_taxa, "Other")
mean_rel_long$taxon_plot <- factor(mean_rel_long$taxon_plot,
                                   levels = ordered_taxa)

# Okabe–Ito palette for taxa (you can adjust / reuse your existing colors)
cb_palette <- c(
  "#E69F00", # orange
  "#56B4E9", # light blue
  "#009E73", # green
  "#F0E442", # yellow
  "#0072B2", # blue
  "#D55E00", # red
  "#CC79A7"  # purple
)
cb_other <- "#999999"   # grey for "Other"

taxa_colors <- c(cb_palette, cb_other)
names(taxa_colors) <- ordered_taxa

ggplot(mean_rel_long,
       aes(x = group, y = rel_abund, fill = taxon_plot)) +
  geom_bar(stat = "identity", position = "stack", color = "white") +
  scale_fill_manual(values = taxa_colors, name = "Genus") +
  labs(
    title = paste("Mean genus composition (top", K, "genera)"),
    x     = NULL,
    y     = "Relative abundance (mean across samples)"
  ) +
  theme_minimal() +
  theme(
    axis.text    = element_text(size = 14),
    axis.title   = element_text(size = 14),
    plot.title   = element_text(size = 15, face = "bold"),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 14)
  )
```

## dacomp - Differential abundance analysis

### Build count matrix and phenotype vector

```{r dacomp_data_prep}
# Combine counts into one matrix X: rows = samples, columns = taxa
X_counts <- rbind(counts_g1, counts_g2)

# Phenotype / group labels corresponding to the rows of X_counts
y_groups <- factor(
  c(rep(label1, nrow(counts_g1)),
    rep(label2, nrow(counts_g2))),
  levels = c(label1, label2)
)

dim(X_counts)
table(y_groups)
```

### Select reference taxa

```{r dacomp_select_refs}
set.seed(42)

# Reference selection:
# You can tweak minimal_TA and/or median_SD_threshold.
res_refs <- dacomp.select_references(
  X = X_counts,
  minimal_TA = 50,          # ensure enough counts under reference set
  verbose   = TRUE
)

res_refs
length(res_refs$selected_references)

tax_names <- colnames(X_counts)
cat("Reference taxa: \n")
tax_names[res_refs$selected_references]

# Diagnostic plot of reference scores
dacomp.plot_reference_scores(res_refs)
```

Of the 118 taxa, 93 are selected as reference. 

```{r dacmp_prev_var_plot, fig.width=6, fig.height=4, dpi=200}
## Prevalence–variance plot for genus level (à la Sommer et al.)

# X_counts: samples x taxa (same object you use for dacomp)
X <- X_counts

n_samples <- nrow(X)

## 1) Prevalence: proportion of samples with count > 0
taxa_prev <- colSums(X > 0) / n_samples

## 2) Variance of log((x + 1)/sum(x)) across samples
#    (this matches: transform_sample_counts(ps, function(x) log((x+1)/sum(x))))
X_logrel <- log(
  sweep(X + 1, 1, rowSums(X), FUN = "/")
)

taxa_var <- apply(X_logrel, 2, var, na.rm = TRUE)

## 3) Classification into "reference" vs "test" using the same thresholds
##    as in the script you showed (for genus level):
##        variance < 2  &  prevalence > 0.9
is_ref_rule <- (taxa_var < 2) & (taxa_prev > 0.9)

prev_var_df <- tibble(
  taxon      = colnames(X),
  prevalence = taxa_prev,
  variance   = taxa_var,
  is_ref     = is_ref_rule
)

ggplot(prev_var_df,
       aes(x = prevalence, y = variance)) +
  geom_point(aes(color = is_ref), size = 1.5, alpha = 0.9) +
  geom_vline(xintercept = 0.9, linetype = "dashed") +
  geom_hline(yintercept = 2,   linetype = "dashed") +
  scale_color_manual(
    values = c(`FALSE` = "darkgrey", `TRUE` = "red"),
    labels = c(`FALSE` = "test", `TRUE` = "reference"),
    name   = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    title = "Reference selection rule at genus level",
    x     = "Prevalence (proportion of samples with count > 0)",
    y     = "Variance of log((x + 1)/sum(x))"
  ) +
  theme_bw() +
  theme(
    legend.position  = "right",
    panel.grid.minor = element_blank()
  )
```

```{r dacomp_prev_var_with_refs, fig.width=6, fig.height=4, dpi=200}
# X_counts: samples x taxa used in dacomp
X <- X_counts
n_samples <- nrow(X)

## 1) Prevalence: proportion of samples with count > 0
taxa_prev <- colSums(X > 0) / n_samples

## 2) Variance of log((x + 1)/sum(x)) across samples
X_logrel <- log(
  sweep(X + 1, 1, rowSums(X), FUN = "/")
)
taxa_var <- apply(X_logrel, 2, var, na.rm = TRUE)

## 4) dacomp-selected references (indices in columns of X)
dacomp_idx  <- res_refs$selected_references
is_dacomp_ref <- seq_along(colnames(X)) %in% dacomp_idx

## 5) Combine into a single factor for coloring
prev_var_df <- tibble(
  taxon        = colnames(X),
  prevalence   = taxa_prev,
  variance     = taxa_var,
  dacomp_ref   = is_dacomp_ref,
  ref_type     = case_when(
    dacomp_ref              ~ "reference",
    TRUE                    ~ "test"
  )
)

ggplot(prev_var_df,
       aes(x = prevalence, y = variance)) +
  geom_point(aes(color = ref_type), size = 1.6, alpha = 0.9) +
  scale_color_manual(
    values = c(
      "test"                = "darkgrey",
      "reference"     = "forestgreen"
    ),
    name = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  #geom_vline(xintercept = 0.9, linetype = "dashed") +
  #geom_hline(yintercept = 2,   linetype = "dashed") +
  labs(
    title = "Prevalence–variance and reference selection (genus level)",
    x     = "Prevalence (proportion of samples with count > 0)",
    y     = "Variance of log((x + 1)/sum(x))"
  ) +
  theme_bw() +
  theme(
    legend.position  = "right",
    panel.grid.minor = element_blank()
  )
```

```{r dacomp_prev_vs_score, fig.width=6, fig.height=4, dpi=200}
X <- X_counts
n_samples <- nrow(X)

# Prevalence (same definition as before)
taxa_prev <- colSums(X > 0) / n_samples

df_prev_score <- tibble::tibble(
  taxon      = colnames(X),
  prevalence = taxa_prev,
  score      = res_refs$scores,
  is_ref     = seq_along(colnames(X)) %in% res_refs$selected_references
)

## taxa to annotate
taxa_to_annotate <- df_prev_score %>%
  dplyr::filter(is_ref & prevalence >= 0.25) %>%
  # dplyr::arrange(dplyr::desc(score)) %>%
  # dplyr::slice(1:5) %>%
  dplyr::pull(taxon)

ggplot(df_prev_score,
       aes(x = prevalence, y = score)) +
  geom_point(aes(color = is_ref), size = 1.7, alpha = 0.9) +
  scale_color_manual(
    values = c(`FALSE` = "darkgrey", `TRUE` = "forestgreen"),
    labels = c(`FALSE` = "test", `TRUE` = "reference"),
    name   = NULL
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  labs(
    title = "Prevalence vs dacomp score (genus level)",
    x     = "Prevalence (proportion of samples with count > 0)",
    y     = "dacomp median SD score"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position  = "right"
  ) +
  ggrepel::geom_text_repel(
    data = subset(df_prev_score, taxon %in% taxa_to_annotate),
    aes(label = taxon),
    size          = 3,
    box.padding   = 0.3,
    point.padding = 0.5,
    min.segment.length = 0,
    max.overlaps  = Inf,
    segment.alpha = 0.6
  )

```

### Run dacomp test (runtime comparison)

```{r run_dacomp_test, eval=FALSE}
# DS-FDR target
q_DSFDR <- 0.1

run_dacomp <- function(n_perm) {
  dacomp.test(
    X                  = X_counts,
    y                  = y_groups,
    ind_reference_taxa = res_refs,
    test               = DACOMP.TEST.NAME.WILCOXON,
    nr_perm            = n_perm,
    q                  = q_DSFDR,
    verbose            = FALSE,
    return_rarefied_values = TRUE,
    Test_All           = TRUE
  )
}

# ----------------------------
# Settings
# ----------------------------
perm_grid <- c(1e3, 1e4)#, 1e5, 1e6, 1e7)

seed_run <- 42          # define once; used for RNG + filenames
refit_dacomp <- FALSE   # TRUE = force recompute

# helper for consistent file tags like "1e+03"
fmt_nperm <- function(n) format(n, scientific = TRUE)

# ----------------------------
# Run + cache one file per n_perm
# Each file contains ONLY the dacomp.result.object for that n_perm
# ----------------------------

for (n_perm in perm_grid) {

  set.seed(seed_run)  # ensures reproducible dacomp run for this n_perm/seed

  nperm_tag <- fmt_nperm(n_perm)

  # File for a single dacomp result (one n_perm)
  fname_res <- path_target(
    paste0("dacomp_res_nperm=", nperm_tag, "_seed=", seed_run, ".rds")
  )

  # File for runtime (one n_perm)
  fname_rt <- path_target(
    paste0("dacomp_runtime_nperm=", nperm_tag, "_seed=", seed_run, ".rds")
  )

  if (!file.exists(fname_res) || isTRUE(refit_dacomp)) {
    cat("Running dacomp:", "n_perm =", nperm_tag, "| seed =", seed_run, "\n")

    t <- system.time({
      res_dacomp <- run_dacomp(n_perm)
    })

    # IMPORTANT: save ONLY the dacomp.result.object (same as dacomp_res_list[[...]])
    saveRDS(res_dacomp, file = fname_res)

    # store runtime row (single-file)
    rt_row <- tibble::tibble(
      nr_perm = n_perm,
      elapsed = unname(t["elapsed"]),
      seed    = seed_run
    )
    saveRDS(rt_row, file = fname_rt)
  } else {
    cat("Skipping (cached):", "n_perm =", nperm_tag, "| seed =", seed_run, "\n")
  }
}
```

```{r load_dacomp_results}
fmt_nperm <- function(n) format(n, scientific = TRUE)

load_dacomp_seeded <- function(
    seed,
    perm_grid = c(1e3, 1e4, 1e5, 1e6, 1e7),
    strict = TRUE  # TRUE: error if any missing; FALSE: load what exists
) {

  dacomp_res_list <- list()

  for (n_perm in perm_grid) {

    fname_res <- path_target(
      paste0("dacomp_res_nperm=", fmt_nperm(n_perm), "_seed=", seed, ".rds")
    )

    if (!file.exists(fname_res)) {
      msg <- paste0("Missing dacomp file: ", basename(fname_res))
      if (isTRUE(strict)) stop(msg) else { message(msg); next }
    }

    # keep your old naming style for downstream compatibility:
    # 1000/10000 as plain, >=1e5 as scientific (like your screenshot)
    key <- if (n_perm < 1e5) as.character(as.integer(n_perm)) else fmt_nperm(n_perm)

    dacomp_res_list[[key]] <- readRDS(fname_res)
  }

  ## runtime table: prefer reading per-n_perm runtime files if present;
  ## otherwise, compute elapsed = NA (you can fill later)
  runtime_rows <- purrr::map_dfr(perm_grid, function(n_perm) {

    fname_rt <- path_target(
      paste0("dacomp_runtime_nperm=", fmt_nperm(n_perm), "_seed=", seed, ".rds")
    )

    if (file.exists(fname_rt)) {
      readRDS(fname_rt)
    } else {
      tibble::tibble(nr_perm = n_perm, elapsed = NA_real_, seed = seed)
    }
  }) %>%
    dplyr::arrange(nr_perm)

  list(
    dacomp_res_list = dacomp_res_list,
    runtime_tbl     = runtime_rows
  )
}

seed_run <- 42

loaded <- load_dacomp_seeded(
  seed = seed_run,
  perm_grid = c(1e3, 1e4, 1e5, 1e6, 1e7),
  strict = FALSE  # loads whatever exists
)

dacomp_res_list <- loaded$dacomp_res_list
runtime_tbl     <- loaded$runtime_tbl

str(dacomp_res_list, 1)
runtime_tbl
```

### Extrapolate runtime

```{r extrapolate_time}
fit_lin <- lm(elapsed ~ nr_perm, data = runtime_tbl)
summary(fit_lin)

new_perm <- tibble::tibble(
  nr_perm = c(1e6, 1e7, 1e8)
)

# Observed
obs_tbl <- runtime_tbl %>%
  dplyr::mutate(
    source       = "observed",
    elapsed_min  = round(elapsed / 60, 3),
    elapsed_h    = round(elapsed_min / 60, 3)
  )

# Predicted
pred_tbl <- new_perm %>%
  dplyr::mutate(
    elapsed      = as.numeric(predict(fit_lin, newdata = new_perm)),
    elapsed_min  = round(elapsed / 60, 3),
    elapsed_h    = round(elapsed_min / 60, 3),
    source       = "extrapolated"
  )

# Combine
runtime_full_tbl <- 
  dplyr::bind_rows(obs_tbl) %>%
  dplyr::arrange(nr_perm) %>%
  dplyr::rename(n_perm = nr_perm, elapsed_sec = elapsed) %>%
  select(n_perm, source, elapsed_sec, elapsed_min, elapsed_h)

runtime_full_tbl
```


```{r runtime, fig.width=7, fig.height=3, dpi=300}
x_grid <- sort(unique(runtime_full_tbl$n_perm))

minute_grid <- c(1, 10, 60, 120, 300)
second_grid <- minute_grid * 60

ggplot(runtime_full_tbl, aes(x = n_perm, y = elapsed_sec)) +

  ## horizontal grid lines for minutes
  geom_hline(
    yintercept = second_grid,
    color = "grey85",
    linewidth = 0.3
  ) +

  ## main data
  geom_line(
    color = "black",
    linewidth = 0.6
  ) +
  geom_point(
    color = "black",
    size = 2,
    stroke = 0.8
  ) +

  ## x axis
  scale_x_log10(
    breaks = x_grid,
    # labels = scales::label_number(
    #   scale_cut = scales::cut_short_scale()
    # )
    labels = scales::label_scientific(digits = 0)
  ) +

  ## y axis with minutes
  scale_y_log10(
    name = "Runtime (seconds)",
    sec.axis = sec_axis(
      ~ . / 60,
      name   = "Runtime (minutes)",
      breaks = minute_grid,
      labels = scales::label_number()
    )
  ) +

  labs(
    x = "Number of permutations",
    y = "Runtime (seconds)",
    linetype = NULL,
    shape    = NULL,
    fill     = NULL
  ) +
  
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
  )

```

## permApprox

We recompute the dacomp p-values with permApprox for 1e03 and 1e04 permutations. 

For 1e04 (10000) permutations, we consider two cases: exceed0 = 250 and exceed0 = 0.25.

```{r}
perm_grid_pApp <- c(1e3, 1e4)

sampsize <- min(nrow(counts_g1), nrow(counts_g2))  # for SLLS rule

# Helpers for the control functions
make_ctrl_slls <- function(exceed0, thresh_method) {
  make_gpd_ctrl(
    fit_method    = "lme",
    include_obs   = FALSE,
    constraint    = "support_at_obs",
    sample_size   = sampsize,
    thresh_method = thresh_method,
    exceed0       = exceed0,
    exceed0_min   = 250,
    exceed_min    = 10,
    thresh_step   = 10
  )
}

make_ctrl_unconstrained <- function(exceed0 = 250, thresh_method) {
  make_gpd_ctrl(
    fit_method    = "lme",
    include_obs   = FALSE,
    constraint    = "unconstrained",
    thresh_method = thresh_method,
    exceed0       = exceed0,
    exceed0_min   = exceed0,
    exceed_min    = 10,
    thresh_step   = 10
  )
}

# Helper to run permApprox
run_permApprox_from_dacomp <- function(res_dacomp,
                                       taxa_ids,
                                       constraint = c("slls", "unconstrained"),
                                       exceed0,
                                       approx_thresh = 0.1) {

  constraint <- match.arg(constraint)

  stats_mat  <- res_dacomp$stats_matrix
  obs_stats  <- stats_mat[1, ]
  perm_stats <- stats_mat[-1, , drop = FALSE]

  names(obs_stats)     <- taxa_ids
  colnames(perm_stats) <- taxa_ids
  
  thresh_method <- case_when(
    exceed0 == 250 ~ "fix",
    exceed0 == 0.25 ~ "rob_ftr"
  )
  
  gpd_ctrl <- switch(
    constraint,
    slls          = make_ctrl_slls(exceed0, thresh_method),
    unconstrained = make_ctrl_unconstrained(exceed0, thresh_method)
  )

  perm_approx(
    obs_stats     = obs_stats,
    perm_stats    = perm_stats,
    alternative   = "two_sided",
    approx_thresh = approx_thresh,
    method        = "gpd",
    gpd_ctrl      = gpd_ctrl,
    adjust_method = "BH",
    cores         = 1,
    verbose       = TRUE
  )
}


refit_permApprox <- FALSE
pApp_res_list <- list()

for (n_perm in perm_grid_pApp) {

  res_dacomp <- dacomp_res_list[[as.character(n_perm)]]
  if (is.null(res_dacomp)) stop("Missing dacomp result for n_perm = ", n_perm)

  taxa_ids <- colnames(X_counts)

  ## define settings per n_perm
  settings <- list(
    list(constraint = "unconstrained", exceed0 = 250),
    list(constraint = "slls",           exceed0 = 250)
  )

  if (n_perm == 1e4) {
    settings <- append(
      settings,
      list(list(constraint = "slls", exceed0 = 0.25))
    )
  }

  for (s in settings) {

    key <- paste0(
      "nperm=", format(n_perm, scientific = TRUE),
      "_constraint=", s$constraint,
      "_exceed0=", s$exceed0
    )

    fname <- path_target(paste0("permApprox_", key, ".rds"))

    if (!file.exists(fname) || isTRUE(refit_permApprox)) {
      cat("permApprox fit:", key, "\n")

      t <- system.time({
        pApp <- run_permApprox_from_dacomp(
          res_dacomp  = res_dacomp,
          taxa_ids    = taxa_ids,
          constraint  = s$constraint,
          exceed0     = s$exceed0
        )
      })

      saveRDS(
        list(
          papprox    = pApp,
          elapsed_s  = unname(t["elapsed"]),
          n_perm     = n_perm,
          constraint = s$constraint,
          exceed0    = s$exceed0
        ),
        file = fname
      )
    }

    pApp_res_list[[key]] <- readRDS(fname)
  }
}
```

## Compute log2 fold-changes

```{r}
## Compute log2 fold-change once (shared across all panels)

rel_counts <- X_counts / rowSums(X_counts)

group_means <- aggregate(rel_counts,
                         by = list(group = y_groups),
                         FUN = mean)

mean_g1 <- as.numeric(group_means[group_means$group == label1, -1])
mean_g2 <- as.numeric(group_means[group_means$group == label2, -1])

eps <- 1e-6
log2_fc_vec <- log2((mean_g2 + eps) / (mean_g1 + eps))
names(log2_fc_vec) <- colnames(X_counts)
```

## Build one combined tibble

```{r}
## ----------------------------
## Build ONE unified long tibble (updated for new permApprox settings):
## - permApprox now has: constraint ∈ {unconstrained, slls} + exceed0
## - setting string should reflect BOTH (constraint + exceed0)
## ----------------------------

taxa_ids <- colnames(X_counts)

## Helper: robust n_perm parser for dacomp_res_list keys ("1000", "1e+06", ...)
parse_nperm <- function(x) {
  out <- suppressWarnings(as.numeric(x))
  if (!is.na(out)) return(out)
  # try to parse scientific-style keys like "1e+06"
  suppressWarnings(as.numeric(gsub("\\+0", "+", x)))
}

## Helper: convert dacomp result -> tibble
dacomp_to_tbl <- function(res_dacomp, taxa_ids, res_refs) {
  tibble::tibble(
    taxon        = taxa_ids,
    stat_dacomp  = as.numeric(res_dacomp$stats_matrix[1, ]),
    p_dacomp_raw = as.numeric(res_dacomp$p.values.test),
    p_dacomp_adj = as.numeric(res_dacomp$p.values.test.adjusted),
    is_reference = seq_along(taxa_ids) %in% res_refs$selected_references
  ) %>%
    dplyr::mutate(
      # your previous handling: Inf -> 1 for plotting/thresholding
      p_dacomp_adj = dplyr::if_else(is.infinite(p_dacomp_adj), 1, p_dacomp_adj)
    )
}

## ----------------------------
## dacomp long results (all n_perm)
## ----------------------------
dacomp_long <- purrr::imap_dfr(dacomp_res_list, function(res_dacomp, n_perm_chr) {

  n_perm <- parse_nperm(n_perm_chr)

  base_tbl <- dacomp_to_tbl(res_dacomp, taxa_ids, res_refs)

  base_tbl %>%
    dplyr::transmute(
      taxon,
      log2_fc      = unname(log2_fc_vec[taxon]),
      is_reference,
      panel_method = "dacomp",
      n_perm       = n_perm,
      constraint   = NA_character_,
      exceed0      = NA_real_,
      setting      = NA_character_,
      stat         = stat_dacomp,
      p_raw        = p_dacomp_raw,
      p_adj        = p_dacomp_adj
    )
})

## ----------------------------
## permApprox long results (selected n_perm + constraint + exceed0)
## ----------------------------
permApprox_long <- purrr::imap_dfr(pApp_res_list, function(x, key) {

  pApp        <- x$papprox
  n_perm      <- x$n_perm
  constraint  <- x$constraint
  ex0         <- x$exceed0

  ## nice human-readable label
  setting_chr <- dplyr::case_when(
    constraint == "slls"          ~ paste0("SLLS, exceed0=", ex0),
    constraint == "unconstrained" ~ paste0("unconstrained, exceed0=", ex0),
    TRUE                          ~ paste0(constraint, ", exceed0=", ex0)
  )

  tibble::tibble(
    taxon        = taxa_ids,
    log2_fc      = unname(log2_fc_vec[taxa_ids]),
    is_reference = seq_along(taxa_ids) %in% res_refs$selected_references,
    panel_method = "permApprox",
    n_perm       = n_perm,
    constraint   = constraint,
    exceed0      = ex0,
    setting      = setting_chr,
    stat         = NA_real_,
    p_raw        = as.numeric(pApp$p_unadjusted),
    p_adj        = as.numeric(pApp$p_values)
  )
})

## ----------------------------
## unified
## ----------------------------
res_all_long <- dplyr::bind_rows(dacomp_long, permApprox_long) %>%
  dplyr::mutate(
    n_perm_label = scales::label_scientific(digits = 0)(n_perm),
    panel = dplyr::case_when(
      panel_method == "dacomp" ~ paste0("dacomp (", n_perm_label, ")"),
      panel_method == "permApprox" ~ paste0("permApprox (", n_perm_label, ", ", 
                                            setting, ")"),
      TRUE ~ "other"
    ),
    panel = factor(panel)
  )

saveRDS(res_all_long, file = path_target("results_all_long.rds"))

## quick check
res_all_long %>%
  dplyr::count(panel_method, n_perm, constraint, exceed0, setting)
```

## Significant taxa

```{r}
## Load the unified long table
res_all_long <- readRDS(path_target("results_all_long.rds"))
alpha_sig <- 0.01
```

### Adjusted

```{r}
fmt_p <- function(x, digits = 2) {
  ifelse(is.na(x), NA_character_, formatC(x, format = "e", digits = digits))
}

p_wide_adj <- res_all_long %>%
  dplyr::select(taxon, panel, p_adj) %>%
  tidyr::pivot_wider(names_from = panel, values_from = p_adj) %>%
  dplyr::arrange(`permApprox (1e+03, SLLS, exceed0=250)`)

p_wide_adj_print <- p_wide_adj %>%
  dplyr::mutate(
    dplyr::across(
      -taxon,
      ~ fmt_p(.x, digits = 2)
    )
  )

p_wide_adj_print %>%
  knitr::kable(
    align = "l",
    caption = "Adjusted p-values across methods and permutation settings"
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover")
  )
```

```{r}
sig_wide_adj <- res_all_long %>%
  dplyr::mutate(sig = !is.na(p_adj) & (p_adj <= alpha_sig)) %>%
  dplyr::select(taxon, panel, sig) %>%
  tidyr::pivot_wider(
    names_from  = panel,
    values_from = sig,
    values_fill = FALSE
  ) %>%
  dplyr::arrange(taxon)

sig_overview <- sig_wide_adj %>%
  dplyr::mutate(
    n_sig   = rowSums(dplyr::across(-taxon)),
    sig_any = n_sig > 0,
    sig_all = n_sig == (ncol(.) - 1)
  ) %>%
  dplyr::arrange(dplyr::desc(n_sig), taxon) %>%
  select(taxon, n_sig, sig_any, sig_all, everything())

sig_overview %>%
  dplyr::slice_head(n = 30) %>%
  kableExtra::kable() %>%
  kableExtra::kable_styling(full_width = FALSE)
```

### Unadjusted

```{r}
p_wide_raw <- res_all_long %>%
  dplyr::select(taxon, panel, p_raw) %>%
  tidyr::pivot_wider(
    names_from  = panel,
    values_from = p_raw
  ) %>%
  dplyr::arrange(`permApprox (1e+03, SLLS, exceed0=250)`)

fmt_p <- function(x, digits = 2) {
  ifelse(is.na(x), NA_character_, formatC(x, format = "e", digits = digits))
}

p_wide_raw_print <- p_wide_raw %>%
  dplyr::mutate(
    dplyr::across(
      -taxon,
      ~ fmt_p(.x, digits = 2)
    )
  )

p_wide_raw_print %>%
  knitr::kable(
    align   = "l",
    caption = "Unadjusted p-values across methods and permutation settings"
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover")
  )

```

```{r}
sig_wide_raw <- res_all_long %>%
  dplyr::mutate(sig = !is.na(p_raw) & (p_raw <= alpha_sig)) %>%
  dplyr::select(taxon, panel, sig) %>%
  tidyr::pivot_wider(
    names_from  = panel,
    values_from = sig,
    values_fill = FALSE
  )

sig_overview <- sig_wide_raw %>%
  dplyr::mutate(
    n_sig   = rowSums(dplyr::across(-taxon)),
    sig_any = n_sig > 0,
    sig_all = n_sig == (ncol(.) - 1)
  ) %>%
  dplyr::arrange(dplyr::desc(n_sig), taxon) %>%
  select(taxon, n_sig, sig_any, sig_all, everything())

sig_overview %>%
  dplyr::slice_head(n = 30) %>%
  kableExtra::kable() %>%
  kableExtra::kable_styling(full_width = FALSE)
```


## Taxa with zero p-values

```{r}
uncon_cols <- grep(
  "permApprox.*unconstrained",
  colnames(p_wide_adj),
  value = TRUE
)

p_zero_uncon <- p_wide_adj %>%
  dplyr::filter(
    dplyr::if_any(
      dplyr::all_of(uncon_cols),
      ~ !is.na(.) & . == 0
    )
  )

p_zero_uncon %>%
  dplyr::mutate(across(-taxon, ~ fmt_p(.x))) %>%
  knitr::kable() %>%
  kableExtra::kable_styling(full_width = FALSE)
```


## Plots

### Volcano plots

#### Plot function

```{r}
plot_volcano_long <- function(
    res_all_long,
    panels = NULL,                  # exact panel strings; if NULL, uses filters below
    panel_methods = NULL,           # e.g. c("dacomp","permApprox")
    n_perm_dacomp = NULL,           # permutations used for dacomp
    n_perm_permApprox = NULL,       # permutations used for permApprox
    constraints = NULL,             # e.g. c("slls","unconstrained") (permApprox only)
    exceed0_vals = NULL,            # e.g. c(250, 0.25) (permApprox only)
    p_type = c("adj", "raw"),
    alpha_thresh = 0.01,
    lfc_thresh = 1,
    title = NULL,
    annotate_taxa = "significant",  # FALSE/NULL, "significant", or character vector
    # facet label control:
    panel_label_style = c("auto", "none"),  # auto = title + subtitle per panel; none = use panel as-is
    panel_label_map = NULL, # named character vector: names = original panel, values = "Title\nSubtitle"
    box_padding = 0.3,
    point_padding = 0.7,
    force = 1.4,
    label_col_some = "red",
    y_breaks = NULL
) {
  p_type <- match.arg(p_type)
  panel_label_style <- match.arg(panel_label_style)

  # basic checks
  req_cols <- c("taxon","log2_fc","panel","panel_method","n_perm","p_raw","p_adj",
                "constraint","exceed0","setting")
  stopifnot(all(req_cols %in% names(res_all_long)))

  df <- res_all_long
  p_col <- if (p_type == "adj") "p_adj" else "p_raw"

  # ----------------------------
  # Subset to panels of interest
  # ----------------------------
  if (!is.null(panels)) {
    df <- df %>% dplyr::filter(panel %in% panels)
  } else {
    if (!is.null(panel_methods)) df <- df %>% dplyr::filter(panel_method %in% panel_methods)
    
    if (!is.null(n_perm_dacomp)) {
      df <- df %>%
        dplyr::filter(
          panel_method != "dacomp" | n_perm %in% n_perm_dacomp
        )
    }
    
    if (!is.null(n_perm_permApprox)) {
      df <- df %>%
        dplyr::filter(
          panel_method != "permApprox" | n_perm %in% n_perm_permApprox
        )
    }
    
    # Apply constraint/exceed0 filters only to permApprox rows; keep dacomp rows (NA) unaffected.
    if (!is.null(constraints)) {
      df <- df %>%
        dplyr::filter(panel_method != "permApprox" | constraint %in% constraints)
    }
    if (!is.null(exceed0_vals)) {
      df <- df %>%
        dplyr::filter(panel_method != "permApprox" | exceed0 %in% exceed0_vals)
    }
  }

  if (nrow(df) == 0) stop("No rows left after filtering. Check your panels/filters.")

  # keep only existing panel levels (prevents empty facets)
  df$panel <- droplevels(factor(df$panel))

  # ----------------------------
  # Build panel labels: rewrite titles + add subtitles
  # ----------------------------
  if (panel_label_style == "auto") {

    df <- df %>%
      dplyr::mutate(
        n_perm_pretty = scales::label_scientific(digits = 0)(n_perm),

        panel_title = dplyr::case_when(
          panel_method == "dacomp" ~ "dacomp",
          panel_method == "permApprox" & constraint == "slls" ~ "permApprox (SLLS)",
          panel_method == "permApprox" & constraint == "unconstrained" ~ "permApprox (unconstrained)",
          panel_method == "permApprox" ~ paste0("permApprox (", constraint, ")"),
          TRUE ~ as.character(panel_method)
        ),

        panel_subtitle = dplyr::case_when(
          panel_method == "dacomp" ~ paste0(n_perm_pretty, " permutations"),
          panel_method == "permApprox" ~ paste0(
            n_perm_pretty, " permutations, exceed0 = ", exceed0
          ),
          TRUE ~ ""
        ),

        panel_lab = paste(panel_title, panel_subtitle, sep = "\n")
      )

  } else {
    df <- df %>% dplyr::mutate(panel_lab = as.character(panel))
  }

  # Manual overrides (if provided): values should already contain "\n" for subtitle
  if (!is.null(panel_label_map)) {
    stopifnot(is.character(panel_label_map), !is.null(names(panel_label_map)))
    df <- df %>%
      dplyr::mutate(
        panel_lab = dplyr::if_else(
          as.character(panel) %in% names(panel_label_map),
          unname(panel_label_map[as.character(panel)]),
          panel_lab
        )
      )
  }

  df$panel_lab <- droplevels(factor(df$panel_lab, levels = unique(df$panel_lab)))

  # ----------------------------
  # Significance flags across selected panels
  # ----------------------------
  df <- df %>%
    dplyr::mutate(
      p = .data[[p_col]],
      sig_this = !is.na(p) & (p <= alpha_thresh),
      neglogp_raw = dplyr::if_else(!is.na(p) & p > 0, -log10(p), NA_real_)
    )

  sig_by_taxon <- df %>%
    dplyr::group_by(taxon) %>%
    dplyr::summarise(
      n_panels = dplyr::n_distinct(panel_lab),
      n_sig    = sum(sig_this, na.rm = TRUE),
      sig_all  = (n_sig == n_panels),
      .groups  = "drop"
    )

  df <- df %>%
    dplyr::left_join(sig_by_taxon, by = "taxon") %>%
    dplyr::mutate(
      label_class = dplyr::case_when(
        sig_this & sig_all  ~ "all",
        sig_this & !sig_all ~ "some",
        TRUE                ~ "none"
      ),
      direction = dplyr::case_when(
        sig_this & log2_fc > 0 ~ paste0("Higher in '", label2, "'"),
        sig_this & log2_fc < 0 ~ paste0("Higher in '", label1, "'"),
        TRUE                   ~ "No difference"
      )
    )

  # ----------------------------
  # Handle zero p-values -> show at one "Inf" height PER PANEL
  # ----------------------------
  df <- df %>%
    dplyr::group_by(panel_lab) %>%
    dplyr::mutate(
      max_finite = {
        fv <- neglogp_raw[is.finite(neglogp_raw)]
        if (length(fv)) max(fv) else 0
      },
      y_top = max_finite + 1,
      has_zero = any(p == 0, na.rm = TRUE),
      neglogp_plot = dplyr::case_when(
        has_zero & !is.na(p) & p == 0 ~ y_top,
        TRUE                          ~ neglogp_raw
      )
    ) %>%
    dplyr::ungroup()

  has_zero_any <- any(df$p == 0, na.rm = TRUE)
  finite_vals  <- df$neglogp_raw[is.finite(df$neglogp_raw)]
  max_finite_g <- if (length(finite_vals)) max(finite_vals) else 0
  y_top_g      <- if (has_zero_any) max_finite_g + 1 else max_finite_g
  
  # ----------------------------
  # Global y scale
  # ----------------------------
  if (!is.null(y_breaks)) {
    
    # user-specified breaks (on -log10 scale)
    breaks_y <- sort(unique(y_breaks))
    
    # include Inf-top if needed
    if (has_zero_any) {
      breaks_y <- c(breaks_y[breaks_y < y_top_g], y_top_g)
      labels_y <- c(
        as.character(breaks_y[-length(breaks_y)]),
        "Inf"
      )
      y_limits <- c(0, y_top_g)
    } else {
      labels_y <- as.character(breaks_y)
      y_limits <- range(breaks_y)
    }
    
  } else {
    
    # automatic behavior
    if (has_zero_any) {
      base_breaks <- pretty(c(0, max_finite_g))
      base_breaks <- base_breaks[base_breaks < y_top_g]
      breaks_y <- c(base_breaks, y_top_g)
      labels_y <- ifelse(breaks_y == y_top_g, "Inf", as.character(breaks_y))
      y_limits <- c(0, y_top_g)
    } else {
      breaks_y <- pretty(c(0, max_finite_g))
      labels_y <- as.character(breaks_y)
      y_limits <- c(0, max_finite_g)
    }
  }

  # ----------------------------
  # Colors for direction (relies on your globals: col_group1/2 + label1/2)
  # ----------------------------
  dir_colors <- setNames(
    c(col_group1, col_group2, "grey70"),
    c(
      paste0("Higher in '", label1, "'"),
      paste0("Higher in '", label2, "'"),
      "No difference"
    )
  )

  # ----------------------------
  # Base plot (facet labels are now multi-line title+subtitle)
  # ----------------------------
  p <- ggplot(df, aes(x = log2_fc, y = neglogp_plot)) +
    geom_point(aes(color = direction), alpha = 0.8, size = 3) +
    # facet_wrap( ~ panel_lab,
    #             nrow = 1,
    #             labeller = labeller(panel_lab = label_parsed)
    # ) +
    facet_wrap(~ panel_lab, nrow = 1) +
    geom_hline(
      yintercept = -log10(alpha_thresh),
      linetype   = "dashed",
      color      = "gray50"
    ) +
    geom_vline(
      xintercept = c(-lfc_thresh, lfc_thresh),
      linetype   = "dashed",
      color      = "gray50"
    ) +
    scale_color_manual(values = dir_colors, name = NULL) +
    scale_y_continuous(breaks = breaks_y, labels = labels_y, limits = y_limits) +
    labs(
      title = title,
      x     = "log2 fold-change",
      y     = expression(-log[10](p))
    ) +
    theme_minimal() +
    theme(
      axis.text       = element_text(size = 14),
      axis.title      = element_text(size = 14),
      strip.text      = element_text(size = 13),
      legend.text     = element_text(size = 14),
      legend.position = "bottom",
      panel.spacing   = grid::unit(2, "lines")
    )

  # ----------------------------
  # Annotation
  #   - black: significant in ALL selected panels
  #   - red:   significant only in SOME selected panels
  # ----------------------------
  do_annotate <- !(isFALSE(annotate_taxa) || is.null(annotate_taxa) ||
                     (is.character(annotate_taxa) && length(annotate_taxa) == 0))

  if (isTRUE(do_annotate)) {

    df_annot <- df %>% dplyr::filter(!is.na(neglogp_plot))

    if (is.character(annotate_taxa) && length(annotate_taxa) == 1 &&
        identical(annotate_taxa, "significant")) {

      df_annot <- df_annot %>% dplyr::filter(sig_this)

    } else if (is.character(annotate_taxa)) {

      df_annot <- df_annot %>% dplyr::filter(taxon %in% annotate_taxa)

    } else {
      stop("annotate_taxa must be FALSE/NULL, 'significant', or a character vector of taxa.")
    }

    if (nrow(df_annot) > 0) {

      df_annot <- df_annot %>%
        dplyr::mutate(label_col = dplyr::case_when(
          label_class == "all"  ~ "black",
          label_class == "some" ~ label_col_some,
          TRUE                  ~ "black"
        ))

      p <- p +
        ggrepel::geom_text_repel(
          data = df_annot,
          aes(label = taxon),
          color         = df_annot$label_col,
          size          = 3,
          box.padding   = box_padding,
          point.padding = point_padding,
          min.segment.length = grid::unit(0, "lines"),
          segment.alpha = 0.4,
          max.overlaps  = Inf,
          force         = force,
          show.legend   = FALSE
        )
    }
  }

  p
}
```

#### Raw p-values - All significant taxa

##### permApprox unconstrained

```{r volcano_raw_all_sig_1e3_unconstr, fig.width=12, fig.height=5, dpi=300}
set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+03, unconstrained, exceed0=250)" = "dacomp + permApprox(U)\n1000 permutations"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e3,
  constraints   = c("unconstrained"),
  exceed0_vals  = c(250),
  p_type        = "raw",
  annotate_taxa = "significant",
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

```{r volcano_raw_selected_1e3_unconstr, fig.width=12, fig.height=5, dpi=300}
taxa_to_annotate <- p_wide_raw %>%
  filter(`permApprox (1e+03, unconstrained, exceed0=250)` < 1e-04) %>%
  arrange(`permApprox (1e+03, unconstrained, exceed0=250)`) %>%
  slice_head(n = 8) %>%
  pull(taxon)

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+03, unconstrained, exceed0=250)" = "dacomp + permApprox(U)\n1000 permutations"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e3,
  constraints   = c("unconstrained"),
  exceed0_vals  = c(250),
  p_type        = "raw",
  annotate_taxa = taxa_to_annotate,
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

##### permApprox with 1000 permutations

```{r volcano_raw_all_sig_1e3, fig.width=12, fig.height=5, dpi=300}
set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+03, SLLS, exceed0=250)" = "dacomp + permApprox\n1000 permutations"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e3,
  constraints   = c("slls"),
  exceed0_vals  = c(250),
  p_type        = "raw",
  annotate_taxa = "significant",
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```


```{r volcano_raw_selected_1e3, fig.width=12, fig.height=5, dpi=300}
taxa_to_annotate <- p_wide_raw %>%
  filter(`permApprox (1e+03, SLLS, exceed0=250)` < 1e-04) %>%
  arrange(`permApprox (1e+03, SLLS, exceed0=250)`) %>%
  slice_head(n = 8) %>%
  pull(taxon)

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+03, SLLS, exceed0=250)" = "dacomp + permApprox\n1000 permutations"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e3,
  constraints   = c("slls"),
  exceed0_vals  = c(250),
  p_type        = "raw",
  annotate_taxa = taxa_to_annotate,
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

##### permApprox with 10,000 permutations and k0=250

```{r volcano_raw_all_sig_1e4_250, fig.width=12, fig.height=5, dpi=300}
set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+04, SLLS, exceed0=250)" = "dacomp + permApprox\n1e+04 permutations; k0=250"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e4,
  constraints   = c("slls"),
  exceed0_vals  = c(250),
  p_type        = "raw",
  annotate_taxa = "significant",
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

```{r volcano_raw_selected_1e4_250, fig.width=12, fig.height=5, dpi=300}
taxa_to_annotate <- p_wide_raw %>%
  filter(`permApprox (1e+04, SLLS, exceed0=250)` < 1e-04) %>%
  arrange(`permApprox (1e+04, SLLS, exceed0=250)`) %>%
  slice_head(n = 8) %>%
  pull(taxon)

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+04, SLLS, exceed0=250)" = "dacomp + permApprox\n1e+04 permutations; k0=250"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e4,
  constraints   = c("slls"),
  exceed0_vals  = c(250),
  p_type        = "raw",
  annotate_taxa = taxa_to_annotate,
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

##### permApprox with 10,000 permutations and k0=0.25B

```{r volcano_raw_all_sig_1e4_0.25, fig.width=12, fig.height=5, dpi=300}
set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+04, SLLS, exceed0=0.25)" = "dacomp + permApprox\n1e+04 permutations; k0=0.25B"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e4,
  constraints   = c("slls"),
  exceed0_vals  = c(0.25),
  p_type        = "raw",
  annotate_taxa = "significant",
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

```{r volcano_raw_selected_1e4_0.25, fig.width=12, fig.height=5, dpi=300}
taxa_to_annotate <- p_wide_raw %>%
  filter(`permApprox (1e+04, SLLS, exceed0=0.25)` < 1e-04) %>%
  arrange(`permApprox (1e+04, SLLS, exceed0=0.25)`) %>%
  slice_head(n = 8) %>%
  pull(taxon)

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+04, SLLS, exceed0=0.25)" = "dacomp + permApprox\n1e+04 permutations; k0=0.25B"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e4,
  constraints   = c("slls"),
  exceed0_vals  = c(0.25),
  p_type        = "raw",
  annotate_taxa = taxa_to_annotate,
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```



#### Adjusted p-values - All significant taxa

##### permApprox unconstrained

```{r volcano_adj_all_sig_1e3_unconstr, fig.width=12, fig.height=5, dpi=300}
set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+03, unconstrained, exceed0=250)" = "dacomp + permApprox(U)\n1000 permutations"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e3,
  constraints   = c("unconstrained"),
  exceed0_vals  = c(250),
  p_type        = "adj",
  annotate_taxa = "significant",
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

```{r volcano_adj_selected_1e3_unconstr, fig.width=12, fig.height=5, dpi=300}
taxa_to_annotate <- p_wide_adj %>%
  filter(`permApprox (1e+03, unconstrained, exceed0=250)` < 1e-04) %>%
  arrange(`permApprox (1e+03, unconstrained, exceed0=250)`) %>%
  slice_head(n = 8) %>%
  pull(taxon)

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+03, unconstrained, exceed0=250)" = "dacomp + permApprox(U)\n1000 permutations"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e3,
  constraints   = c("unconstrained"),
  exceed0_vals  = c(250),
  p_type        = "adj",
  annotate_taxa = taxa_to_annotate,
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

##### permApprox with 1000 permutations

```{r volcano_adj_all_sig_1e3, fig.width=12, fig.height=5, dpi=300}
set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+03, SLLS, exceed0=250)" = "dacomp + permApprox\n1000 permutations"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e3,
  constraints   = c("slls"),
  exceed0_vals  = c(250),
  p_type        = "adj",
  annotate_taxa = "significant",
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```


```{r volcano_adj_selected_1e3, fig.width=12, fig.height=5, dpi=300}
taxa_to_annotate <- p_wide_adj %>%
  filter(`permApprox (1e+03, SLLS, exceed0=250)` < 1e-04) %>%
  arrange(`permApprox (1e+03, SLLS, exceed0=250)`) %>%
  slice_head(n = 8) %>%
  pull(taxon)

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+03, SLLS, exceed0=250)" = "dacomp + permApprox\n1000 permutations"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e3,
  constraints   = c("slls"),
  exceed0_vals  = c(250),
  p_type        = "adj",
  annotate_taxa = taxa_to_annotate,
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

##### permApprox with 10,000 permutations and k0=250

```{r volcano_adj_all_sig_1e4_250, fig.width=12, fig.height=5, dpi=300}
set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+04, SLLS, exceed0=250)" = "dacomp + permApprox\n1e+04 permutations; k0=250"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e4,
  constraints   = c("slls"),
  exceed0_vals  = c(250),
  p_type        = "adj",
  annotate_taxa = "significant",
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

```{r volcano_adj_selected_1e4_250, fig.width=12, fig.height=5, dpi=300}
taxa_to_annotate <- p_wide_adj %>%
  filter(`permApprox (1e+04, SLLS, exceed0=250)` < 1e-04) %>%
  arrange(`permApprox (1e+04, SLLS, exceed0=250)`) %>%
  slice_head(n = 8) %>%
  pull(taxon)

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+04, SLLS, exceed0=250)" = "dacomp + permApprox\n1e+04 permutations; k0=250"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e4,
  constraints   = c("slls"),
  exceed0_vals  = c(250),
  p_type        = "adj",
  annotate_taxa = taxa_to_annotate,
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

##### permApprox with 10,000 permutations and k0=0.25B

```{r volcano_adj_all_sig_1e4_0.25, fig.width=12, fig.height=5, dpi=300}
set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+04, SLLS, exceed0=0.25)" = "dacomp + permApprox\n1e+04 permutations; k0=0.25B"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e4,
  constraints   = c("slls"),
  exceed0_vals  = c(0.25),
  p_type        = "adj",
  annotate_taxa = "significant",
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```

```{r volcano_adj_selected_1e4_0.25, fig.width=12, fig.height=5, dpi=300}
taxa_to_annotate <- p_wide_adj %>%
  filter(`permApprox (1e+04, SLLS, exceed0=0.25)` < 1e-04) %>%
  arrange(`permApprox (1e+04, SLLS, exceed0=0.25)`) %>%
  slice_head(n = 8) %>%
  pull(taxon)

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations",
  "permApprox (1e+04, SLLS, exceed0=0.25)" = "dacomp + permApprox\n1e+04 permutations; k0=0.25B"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e6, 1e7),
  n_perm_permApprox = 1e4,
  constraints   = c("slls"),
  exceed0_vals  = c(0.25),
  p_type        = "adj",
  annotate_taxa = taxa_to_annotate,
  title         = NULL,
  lfc_thresh   = 0,
  box_padding  = 0.4,
  point_padding = 0.3,
  force        = 2,
  panel_label_map = panel_label_map
)
```


#### Four panel plots (raw p-values)

##### permApprox with 1000 permutations

```{r volcano_raw_four_panel_1e3, fig.width=12, fig.height=5.5, dpi=300}
sel_threshold <- 1e-05

top_permApprox <- p_wide_raw %>%
  filter(`permApprox (1e+03, SLLS, exceed0=250)` < sel_threshold) %>%
  arrange(`permApprox (1e+03, SLLS, exceed0=250)`) %>%
  pull(taxon)

top_dacomp1e6 <- p_wide_raw %>%
  filter(`dacomp (1e+06)` < sel_threshold) %>%
  arrange(`dacomp (1e+06)`) %>%
  pull(taxon)

top_dacomp1e7 <- p_wide_raw %>%
  filter(`dacomp (1e+07)` < sel_threshold) %>%
  arrange(`dacomp (1e+07)`) %>%
  pull(taxon)

taxa_to_annotate_1e3 <- union(top_permApprox, union(top_dacomp1e6, top_dacomp1e7))

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+03)" = "dacomp\n1e+03 permutations\nRuntime: 0.03min",
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations\nRuntime: 27.72min",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations\nRuntime: 283.07min",
  "permApprox (1e+03, SLLS, exceed0=250)" = 
    "dacomp + permApprox\n1e+03 permutations\nRuntime: 0.06min"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e3, 1e6, 1e7),
  n_perm_permApprox = 1e3,
  constraints   = c("slls"),
  exceed0_vals  = c(250),
  p_type        = "raw",
  annotate_taxa = taxa_to_annotate_1e3,
  title         = NULL,
  lfc_thresh    = 0,
  box_padding   = 0.4,
  point_padding = 0.3,
  force         = 2,
  panel_label_map = panel_label_map,
  y_breaks      = c(0, 3, 6, 10, 15, 20)
)
```


##### permApprox with 10,000 permutations and k0=250

```{r volcano_raw_four_panel_1e4_250, fig.width=12, fig.height=5, dpi=300}
sel_threshold <- 1e-05

top_permApprox <- p_wide_raw %>%
  filter(`permApprox (1e+04, SLLS, exceed0=250)` < sel_threshold) %>%
  arrange(`permApprox (1e+04, SLLS, exceed0=250)`) %>%
  pull(taxon)

top_dacomp1e6 <- p_wide_raw %>%
  filter(`dacomp (1e+06)` < sel_threshold) %>%
  arrange(`dacomp (1e+06)`) %>%
  pull(taxon)

top_dacomp1e7 <- p_wide_raw %>%
  filter(`dacomp (1e+07)` < sel_threshold) %>%
  arrange(`dacomp (1e+07)`) %>%
  pull(taxon)

taxa_to_annotate <- union(top_permApprox, union(top_dacomp1e6, top_dacomp1e7))

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+04)" = "dacomp\n1e+04 permutations\nRuntime: 0.32min",
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations\nRuntime: 27.72min",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations\nRuntime: 283.07min",
  "permApprox (1e+04, SLLS, exceed0=250)" = 
    "dacomp + permApprox\n1e+04 permutations\nRuntime: 0.34min"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e4, 1e6, 1e7),
  n_perm_permApprox = 1e4,
  constraints   = c("slls"),
  exceed0_vals  = c(250),
  p_type        = "raw",
  annotate_taxa = taxa_to_annotate,
  title         = NULL,
  lfc_thresh    = 0,
  box_padding   = 0.4,
  point_padding = 0.3,
  force         = 2,
  panel_label_map = panel_label_map,
  y_breaks      = c(0, 2, 4, 6, 9, 12, 15, 18)
)
```

##### permApprox with 10,000 permutations and k0=0.25B

```{r volcano_raw_four_panel_1e4_025B, fig.width=12, fig.height=5, dpi=300}
sel_threshold <- 1e-05

top_permApprox <- p_wide_raw %>%
  filter(`permApprox (1e+04, SLLS, exceed0=0.25)` < sel_threshold) %>%
  arrange(`permApprox (1e+04, SLLS, exceed0=0.25)`) %>%
  pull(taxon)

top_dacomp1e6 <- p_wide_raw %>%
  filter(`dacomp (1e+06)` < sel_threshold) %>%
  arrange(`dacomp (1e+06)`) %>%
  pull(taxon)

top_dacomp1e7 <- p_wide_raw %>%
  filter(`dacomp (1e+07)` < sel_threshold) %>%
  arrange(`dacomp (1e+07)`) %>%
  pull(taxon)

taxa_to_annotate <- union(top_permApprox, union(top_dacomp1e6, top_dacomp1e7))

set.seed(42)

panel_label_map <- c(
  "dacomp (1e+04)" = "dacomp\n1e+04 permutations\nRuntime: 0.32min",
  "dacomp (1e+06)" = "dacomp\n1e+06 permutations\nRuntime: 27.72min",
  "dacomp (1e+07)" = "dacomp\n1e+07 permutations\nRuntime: 283.07min",
  "permApprox (1e+04, SLLS, exceed0=0.25)" = 
    "dacomp + permApprox\n1e+04 permutations\nRuntime: 0.43min"
)

plot_volcano_long(
  res_all_long,
  panel_methods = c("dacomp", "permApprox"),
  n_perm_dacomp = c(1e4, 1e6, 1e7),
  n_perm_permApprox = 1e4,
  constraints   = c("slls"),
  exceed0_vals  = c(0.25),
  p_type        = "raw",
  annotate_taxa = taxa_to_annotate,
  title         = NULL,
  lfc_thresh    = 0,
  box_padding   = 0.4,
  point_padding = 0.3,
  force         = 2,
  panel_label_map = panel_label_map,
  y_breaks      = c(0, 2, 4, 6, 9)
)
```

### Boxplots and Histograms

We plot boxplots for the dacomp-rarefied counts and histograms of the permutation distributions.

```{r}
## -----------------------------------------------------------------------------
## Helper: Compute the taxa sets from res_all_long
## -----------------------------------------------------------------------------

sig_sets_from_long <- function(res_all_long,
                               alpha = 0.01,
                               p_type = c("adj", "raw"),
                               dacomp_panels,
                               permApprox_panel) {
  p_type <- match.arg(p_type)
  p_col <- if (p_type == "adj") "p_adj" else "p_raw"

  # keep only the panels we care about for the set definitions
  df <- res_all_long %>%
    dplyr::filter(panel %in% c(dacomp_panels, permApprox_panel)) %>%
    dplyr::mutate(sig = !is.na(.data[[p_col]]) & (.data[[p_col]] <= alpha)) %>%
    dplyr::select(taxon, panel, sig) %>%
    tidyr::pivot_wider(names_from = panel, values_from = sig, values_fill = FALSE)

  stopifnot(permApprox_panel %in% colnames(df))
  stopifnot(all(dacomp_panels %in% colnames(df)))

  df <- df %>%
    dplyr::mutate(
      sig_permApprox = .data[[permApprox_panel]],
      n_sig_dacomp    = rowSums(dplyr::across(dplyr::all_of(dacomp_panels))),
      sig_any_dacomp  = n_sig_dacomp > 0,
      sig_all_dacomp  = n_sig_dacomp == length(dacomp_panels),

      sig_all = sig_permApprox & sig_all_dacomp,
      sig_permApprox_only = sig_permApprox & !sig_any_dacomp,
      sig_dacomp_only     = sig_any_dacomp & !sig_permApprox
    )

  list(
    wide = df,
    sig_all = df %>% dplyr::filter(sig_all) %>% dplyr::pull(taxon),
    sig_permApprox_only = df %>% dplyr::filter(sig_permApprox_only) %>% dplyr::pull(taxon),
    sig_dacomp_only = df %>% dplyr::filter(sig_dacomp_only) %>% dplyr::pull(taxon)
  )
}
```


```{r}
## -----------------------------------------------------------------------------
## Plot function for boxplots
## -----------------------------------------------------------------------------

plot_rarefied_rel_abund <- function(
  taxa,
  res_dacomp,
  y_groups,
  palette_groups,
  res_all_long,
  order_panel = NULL,     # e.g. order_panel
  p_type = c("adj", "raw"),
  top_n = 8,
  eps_abund = 1e-6,
  ncol = 4,
  title = NULL
) {
  p_type <- match.arg(p_type)
  p_col <- if (p_type == "adj") "p_adj" else "p_raw"

  taxa <- unique(as.character(taxa))
  taxa <- taxa[!is.na(taxa) & nzchar(taxa)]

  if (length(taxa) == 0) {
    return(ggplot() + theme_void() +
             annotate("text", x = 0, y = 0, label = "No taxa to plot", size = 5))
  }

  # dacomp rarefied counts
  dacomp_counts <- res_dacomp$rarefied_counts
  dimnames(dacomp_counts) <- dimnames(X_counts)  # uses your existing X_counts naming

  # keep only taxa present in dacomp_counts
  taxa <- intersect(taxa, colnames(dacomp_counts))
  if (length(taxa) == 0) {
    return(ggplot() + theme_void() +
             annotate("text", x = 0, y = 0, label = "Taxa not found in rarefied_counts", size = 5))
  }

  # optionally select top_n taxa
  if (!is.null(top_n) && length(taxa) > top_n) {

    if (!is.null(order_panel)) {
      ord_tbl <- res_all_long %>%
        dplyr::filter(panel == order_panel, taxon %in% taxa) %>%
        dplyr::select(taxon, p = .data[[p_col]]) %>%
        dplyr::arrange(p)

      taxa <- ord_tbl$taxon[seq_len(min(top_n, nrow(ord_tbl)))]
    } else {
      taxa <- taxa[seq_len(top_n)]
    }
  }

  # relative abundances from rarefied counts
  rel_abund <- sweep(dacomp_counts[, taxa, drop = FALSE], 1, rowSums(dacomp_counts), FUN = "/")

  sample_groups <- tibble::tibble(
    sample = rownames(dacomp_counts),
    group  = y_groups
  )

  abund_df <- rel_abund %>%
    as.data.frame() %>%
    tibble::rownames_to_column("sample") %>%
    dplyr::left_join(sample_groups, by = "sample") %>%
    tidyr::pivot_longer(
      cols      = dplyr::all_of(taxa),
      names_to  = "taxon",
      values_to = "rel_abund"
    ) %>%
    dplyr::mutate(
      group = factor(group, levels = c(label1, label2))
    )

  # order facets by p-values in order_panel (if provided)
  if (!is.null(order_panel)) {
    tax_order <- res_all_long %>%
      dplyr::filter(panel == order_panel, taxon %in% taxa) %>%
      dplyr::select(taxon, p = .data[[p_col]]) %>%
      dplyr::arrange(p) %>%
      dplyr::pull(taxon)

    abund_df <- abund_df %>%
      dplyr::mutate(taxon = factor(taxon, levels = tax_order))
  } else {
    abund_df <- abund_df %>%
      dplyr::mutate(taxon = factor(taxon, levels = taxa))
  }

  abund_nonzero <- abund_df %>% dplyr::filter(rel_abund > 0)

  ggplot() +
    geom_jitter(
      data = abund_df,
      aes(x = group, y = rel_abund + eps_abund, color = group),
      width = 0.15, alpha = 0.5, size = 1.2
    ) +
    geom_boxplot(
      data = abund_nonzero,
      aes(x = group, y = rel_abund + eps_abund, fill = group),
      outlier.shape = NA,
      alpha = 0.8
    ) +
    scale_y_log10(labels = scales::label_scientific(digits = 1)) +
    scale_fill_manual(values = palette_groups) +
    scale_color_manual(values = palette_groups) +
    facet_wrap(~ taxon, scales = "free_y", ncol = ncol) +
    labs(
      title = title,
      x = NULL,
      y = "dacomp-rarefied counts"
    ) +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      legend.title = element_text(size = 14),
      legend.text  = element_text(size = 14),
      axis.text.x  = element_blank(),
      axis.text.y  = element_text(size = 12),
      axis.title   = element_text(size = 14),
      strip.text   = element_text(size = 12)
    )
}

## -----------------------------------------------------------------------------
## Plot function for histograms
## -----------------------------------------------------------------------------

plot_perm_histograms <- function(
  taxa,
  res_dacomp,
  taxa_ids,
  res_all_long = NULL,
  order_panel = NULL,        # e.g. order_panel
  p_type = c("adj", "raw"),  # how to rank taxa if order_panel is used
  top_n = NULL,              # e.g. 8 to mirror your boxplots
  bins = 40,
  ncol = 4,
  title = NULL
) {
  p_type <- match.arg(p_type)
  p_col <- if (p_type == "adj") "p_adj" else "p_raw"

  taxa <- unique(as.character(taxa))
  taxa <- taxa[!is.na(taxa) & nzchar(taxa)]

  if (length(taxa) == 0) {
    return(
      ggplot() +
        theme_void() +
        annotate("text", x = 0, y = 0, label = "No taxa to plot", size = 5)
    )
  }

  # --- pull permutation stats from dacomp ---
  stats_mat  <- res_dacomp$stats_matrix
  obs_stats  <- stats_mat[1, ]
  perm_stats <- stats_mat[-1, , drop = FALSE]

  # align names (important!)
  names(obs_stats)     <- taxa_ids
  colnames(perm_stats) <- taxa_ids

  # keep only taxa that exist in this dacomp run
  taxa <- intersect(taxa, taxa_ids)
  if (length(taxa) == 0) {
    return(
      ggplot() +
        theme_void() +
        annotate("text", x = 0, y = 0, label = "Taxa not found in stats_matrix", size = 5)
    )
  }

  # --- optionally order + take top_n (same idea as your boxplots) ---
  if (!is.null(top_n) && length(taxa) > top_n) {

    if (!is.null(order_panel) && !is.null(res_all_long)) {
      ord_tbl <- res_all_long %>%
        dplyr::filter(panel == order_panel, taxon %in% taxa) %>%
        dplyr::select(taxon, p = .data[[p_col]]) %>%
        dplyr::arrange(p)

      taxa <- ord_tbl$taxon[seq_len(min(top_n, nrow(ord_tbl)))]
    } else {
      taxa <- taxa[seq_len(top_n)]
    }
  } else if (!is.null(order_panel) && !is.null(res_all_long)) {

    tax_order <- res_all_long %>%
      dplyr::filter(panel == order_panel, taxon %in% taxa) %>%
      dplyr::select(taxon, p = .data[[p_col]]) %>%
      dplyr::arrange(p) %>%
      dplyr::pull(taxon)

    # keep only those that were found
    taxa <- intersect(tax_order, taxa)
  }

  # subset permutation matrix
  perm_sub <- perm_stats[, taxa, drop = FALSE]
  obs_sub  <- obs_stats[taxa]

  # long data for ggplot
  perm_long <- as.data.frame(perm_sub) %>%
    tibble::as_tibble() %>%
    tidyr::pivot_longer(
      cols      = dplyr::everything(),
      names_to  = "taxon",
      values_to = "stat_perm"
    ) %>%
    dplyr::mutate(taxon = factor(taxon, levels = taxa))

  obs_df <- tibble::tibble(
    taxon    = factor(taxa, levels = taxa),
    stat_obs = as.numeric(obs_sub)
  )

  facet_labs <- sprintf("%s\n(dacomp stat = %.2f)", taxa, obs_sub[taxa])
  names(facet_labs) <- taxa

  ggplot(perm_long, aes(x = stat_perm)) +
    geom_histogram(bins = bins, color = "white", alpha = 0.9) +
    geom_vline(
      data = obs_df,
      aes(xintercept = stat_obs),
      color = "black",
      linewidth = 0.6
    ) +
    facet_wrap(
      ~ taxon,
      scales = "free",
      ncol = ncol,
      labeller = labeller(taxon = facet_labs)
    ) +
    labs(
      title = title,
      x = "Permutation test statistic",
      y = "Frequency"
    ) +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(size = 12),
      axis.text  = element_text(size = 11),
      axis.title = element_text(size = 13)
    )
}

```

#### 1000 permApprox permutations

```{r}
## -----------------------------------------------------------------------------
## Choose meaningful settings
## -----------------------------------------------------------------------------

alpha_sig <- 0.01

dacomp_panels_sig <- c(
  "dacomp (1e+06)",
  "dacomp (1e+07)"
)

order_panel <- "permApprox (1e+03, SLLS, exceed0=250)"  # for sig + ordering

## -----------------------------------------------------------------------------
## Pick which dacomp run to use for rarefied counts
## -----------------------------------------------------------------------------

n_perm_rarefy <- 1e3
res_dacomp_rarefy <- dacomp_res_list[[as.character(n_perm_rarefy)]]
stopifnot(!is.null(res_dacomp_rarefy))

res_dacomp_hist <- dacomp_res_list[[as.character(n_perm_rarefy)]]
stopifnot(!is.null(res_dacomp_hist))

taxa_ids <- colnames(X_counts)

sig_sets <- sig_sets_from_long(
  res_all_long = res_all_long,
  alpha = alpha_sig,
  p_type = "adj",
  dacomp_panels = dacomp_panels_sig,
  permApprox_panel = order_panel
)

sig_all             <- sig_sets$sig_all
sig_permApprox_only <- sig_sets$sig_permApprox_only
sig_dacomp_only     <- sig_sets$sig_dacomp_only
```

##### Significant in ALL settings

```{r boxplots_1e3_sig_all, fig.width=12, fig.height=6, dpi=200}
# 1) significant in ALL (permApprox main + all selected dacomp panels)
plot_rarefied_rel_abund(
  taxa = sig_all,
  res_dacomp = res_dacomp_rarefy,
  y_groups = y_groups,
  palette_groups = palette_groups,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 4
)
```


```{r perm_histograms_1e3_sig_all, fig.width=12, fig.height=6, dpi=200}
plot_perm_histograms(
  taxa = sig_all,
  res_dacomp = res_dacomp_hist,
  taxa_ids = taxa_ids,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 3
)
```

##### Significant in permApprox only

```{r boxplots_1e3_sig_permApprox, fig.width=4, fig.height=4, dpi=200}
# 2) significant in permApprox main only
plot_rarefied_rel_abund(
  taxa = sig_permApprox_only,
  res_dacomp = res_dacomp_rarefy,
  y_groups = y_groups,
  palette_groups = palette_groups,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 4
)
```

```{r perm_histograms_1e3_sig_permApprox, fig.width=5, fig.height=3, dpi=200}
plot_perm_histograms(
  taxa = sig_permApprox_only,
  res_dacomp = res_dacomp_hist,
  taxa_ids = taxa_ids,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 3
)
```


##### Significant in dacomp panels, but not permApprox

```{r boxplots_1e3_sig_dacomp, fig.width=12, fig.height=4, dpi=200}
# 3) significant in any selected dacomp panel but not permApprox main
plot_rarefied_rel_abund(
  taxa = sig_dacomp_only,
  res_dacomp = res_dacomp_rarefy,
  y_groups = y_groups,
  palette_groups = palette_groups,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 4
)
```

```{r perm_histograms_1e3_sig_dacomp, fig.width=9, fig.height=4, dpi=200}
plot_perm_histograms(
  taxa = sig_dacomp_only,
  res_dacomp = res_dacomp_hist,
  taxa_ids = taxa_ids,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 2
)
```

##### Taxa from four-panel plot for 1e03 permutations

```{r boxplots_1e3_four_panel_dacomp, fig.width=12, fig.height=6, dpi=200}
# 3) significant in any selected dacomp panel but not permApprox main
plot_rarefied_rel_abund(
  taxa = taxa_to_annotate_1e3,
  res_dacomp = res_dacomp_rarefy,
  y_groups = y_groups,
  palette_groups = palette_groups,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 9,
  ncol = 3
)
```

```{r perm_histograms_1e3_four_panel_dacomp, fig.width=12, fig.height=6, dpi=200}
plot_perm_histograms(
  taxa = taxa_to_annotate_1e3,
  res_dacomp = res_dacomp_hist,
  taxa_ids = taxa_ids,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 9,
  ncol = 3
)
```


#### 10000 permApprox permutations

```{r}
## -----------------------------------------------------------------------------
## Choose meaningful settings
## -----------------------------------------------------------------------------

alpha_sig <- 0.01

dacomp_panels_sig <- c(
  "dacomp (1e+06)",
  "dacomp (1e+07)"
)

order_panel <- "permApprox (1e+04, SLLS, exceed0=0.25)"  # for sig + ordering

## -----------------------------------------------------------------------------
## Pick which dacomp run to use for rarefied counts
## -----------------------------------------------------------------------------

n_perm_rarefy <- 1e4
res_dacomp_rarefy <- dacomp_res_list[[as.character(n_perm_rarefy)]]
stopifnot(!is.null(res_dacomp_rarefy))
```

##### Significant in ALL settings

```{r boxplots_1e4_sig_all, fig.width=12, fig.height=6, dpi=200}
# 1) significant in ALL (permApprox main + all selected dacomp panels)
plot_rarefied_rel_abund(
  taxa = sig_all,
  res_dacomp = res_dacomp_rarefy,
  y_groups = y_groups,
  palette_groups = palette_groups,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 4
  #title = "Relative abundances (dacomp rarefied) — significant in ALL selected panels"
)
```

```{r perm_histograms_1e4_sig_all, fig.width=12, fig.height=6, dpi=200}
plot_perm_histograms(
  taxa = sig_all,
  res_dacomp = res_dacomp_hist,
  taxa_ids = taxa_ids,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 3
)
```

##### Significant in permApprox only

```{r boxplots_1e4_sig_permApprox, fig.width=12, fig.height=6, dpi=200}
# 2) significant in permApprox main only
plot_rarefied_rel_abund(
  taxa = sig_permApprox_only,
  res_dacomp = res_dacomp_rarefy,
  y_groups = y_groups,
  palette_groups = palette_groups,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 4
  #title = "Relative abundances (dacomp rarefied) — permApprox-only hits"
)
```

```{r perm_histograms_1e4_sig_permApprox, fig.width=5, fig.height=3, dpi=200}
plot_perm_histograms(
  taxa = sig_permApprox_only,
  res_dacomp = res_dacomp_hist,
  taxa_ids = taxa_ids,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 3
)
```

##### Significant in dacomp panels, but not permApprox

```{r boxplots_1e4_sig_dacomp, fig.width=12, fig.height=6, dpi=200}
# 3) significant in any selected dacomp panel but not permApprox main
plot_rarefied_rel_abund(
  taxa = sig_dacomp_only,
  res_dacomp = res_dacomp_rarefy,
  y_groups = y_groups,
  palette_groups = palette_groups,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 4
  #title = "Relative abundances (dacomp rarefied) — dacomp-only hits"
)
```

```{r perm_histograms_1e4_sig_dacomp, fig.width=9, fig.height=4, dpi=200}
plot_perm_histograms(
  taxa = sig_dacomp_only,
  res_dacomp = res_dacomp_hist,
  taxa_ids = taxa_ids,
  res_all_long = res_all_long,
  order_panel = order_panel,
  p_type = "adj",
  top_n = 8,
  ncol = 2
)
```

## Tables

```{r}
n_top <- 20

# Panel used to SORT the taxa (“main” permApprox panel)
panel_rank <- "permApprox (1e+03, SLLS, exceed0=250)"

# Panels to be used as columns in the table
panels_show <- c(
  "dacomp (1e+03)",
  "dacomp (1e+06)",
  "dacomp (1e+07)",
  "permApprox (1e+03, unconstrained, exceed0=250)",
  "permApprox (1e+03, SLLS, exceed0=250)",
  "permApprox (1e+04, unconstrained, exceed0=250)",
  "permApprox (1e+04, SLLS, exceed0=250)",
  "permApprox (1e+04, SLLS, exceed0=0.25)"
)

stopifnot(panel_rank %in% unique(res_all_long$panel))
stopifnot(all(panels_show %in% unique(res_all_long$panel)))
```

```{r}
# Helpers
sci <- function(x, digits = 2) {
  ifelse(is.na(x), NA_character_, formatC(x, format = "e", digits = digits))
}

build_top_table_from_long <- function(
  res_all_long,
  panel_rank,
  panels_show,
  p_type = c("raw", "adj"),
  n_top = 20,
  exclude_reference = FALSE,
  digits_p = 2
) {
  p_type <- match.arg(p_type)
  p_col  <- if (p_type == "raw") "p_raw" else "p_adj"

  df <- res_all_long %>%
    dplyr::filter(panel %in% union(panel_rank, panels_show)) %>%
    dplyr::select(taxon, log2_fc, is_reference, panel, p = .data[[p_col]])

  if (exclude_reference) {
    df <- df %>% dplyr::filter(!is_reference)
  }

  ## --- determine top taxa by ranking panel -----------------------
  taxa_top <- df %>%
    dplyr::filter(panel == panel_rank) %>%
    dplyr::arrange(p) %>%
    dplyr::slice_head(n = n_top) %>%
    dplyr::pull(taxon)

  ## --- wide p-value table ----------------------------------------
  p_wide <- df %>%
    dplyr::filter(taxon %in% taxa_top, panel %in% panels_show) %>%
    dplyr::select(taxon, panel, p) %>%
    tidyr::pivot_wider(names_from = panel, values_from = p)

  ## --- metadata (one row per taxon) ------------------------------
  meta <- res_all_long %>%
    dplyr::filter(taxon %in% taxa_top) %>%
    dplyr::distinct(taxon, log2_fc, is_reference)

  ## --- combine ---------------------------------------------------
  out <- meta %>%
    dplyr::left_join(p_wide, by = "taxon") %>%
    dplyr::mutate(
      taxon = factor(taxon, levels = taxa_top)
    ) %>%
    dplyr::arrange(taxon) %>%
    dplyr::mutate(
      taxon   = as.character(taxon),
      log2_fc = if ("log2_fc" %in% names(.)) round(log2_fc, 2) else NULL
    )

  ## --- printable version ----------------------------------------
  out_print <- out %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(panels_show),
        ~ ifelse(is.na(.x), NA_character_,
                 formatC(.x, format = "e", digits = digits_p))
      )
    )

  list(
    data_numeric = out,
    data_print   = out_print
  )
}
```

### Unadjusted p-values table

```{r}
tab_raw <- build_top_table_from_long(
  res_all_long = res_all_long,
  panel_rank   = panel_rank,
  panels_show  = panels_show,
  p_type       = "raw",
  n_top        = n_top,
  exclude_reference = FALSE
)

tab_raw$data_print %>%
  dplyr::select(taxon, log2_fc, dplyr::all_of(panels_show)) %>%
  knitr::kable(
    align   = "l",
    caption = paste0(
      "Top ", n_top,
      " genera sorted by ", panel_rank,
      " (unadjusted p-values)."
    )
  ) %>%
  kableExtra::kable_styling(full_width = FALSE)

```


```{r}
tab_raw_latex <- tab_raw$data_print %>%
  dplyr::select(Genus = taxon, `log2 FC` = log2_fc, dplyr::all_of(panels_show))

tab_raw_latex_out <- knitr::kable(
  tab_raw_latex,
  format   = "latex",
  booktabs = TRUE,
  escape   = TRUE,   # keep panel names literal; set FALSE only if you insert math yourself
  caption  = paste0(
    "Top ", n_top,
    " genera sorted by ", panel_rank,
    " (unadjusted p-values)."
  ),
  label    = "tab:top_effects_raw"
) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size     = 8
  )

print(tab_raw_latex_out)
kableExtra::save_kable(tab_raw_latex_out, path_target("top_effects_raw_table.tex"))
```

### Adjusted p-values table

```{r}
tab_adj <- build_top_table_from_long(
  res_all_long = res_all_long,
  panel_rank   = panel_rank,
  panels_show  = panels_show,
  p_type       = "adj",
  n_top        = n_top,
  exclude_reference = FALSE
)

tab_adj$data_print %>%
  dplyr::select(taxon, log2_fc, dplyr::all_of(panels_show)) %>%
  knitr::kable(
    align   = "l",
    caption = paste0(
      "Top ", n_top,
      " genera sorted by ", panel_rank,
      " (BH-adjusted p-values)."
    )
  ) %>%
  kableExtra::kable_styling(full_width = FALSE)

```

```{r}
tab_adj_latex <- tab_adj$data_print %>%
  dplyr::select(Genus = taxon, `log2 FC` = log2_fc, dplyr::all_of(panels_show))

tab_adj_latex_out <- knitr::kable(
  tab_adj_latex,
  format   = "latex",
  booktabs = TRUE,
  escape   = TRUE,
  caption  = paste0(
    "Top ", n_top,
    " genera sorted by ", panel_rank,
    " (BH-adjusted p-values)."
  ),
  label    = "tab:top_effects_adj"
) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size     = 8
  )

print(tab_adj_latex_out)
kableExtra::save_kable(tab_adj_latex_out, path_target("top_effects_adj_table.tex"))

```




