---
title: "Runtime comparison: dacomp vs dacomp + permApprox"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output:
  github_document: default
  html_document:
    toc: true
    toc_float: true
always_allow_html: true
params:
  name: "03_micro_diff_abund_runtime" # change if you rename file
---

```{r here, message=FALSE}
here::i_am(paste0(params$name, ".Rmd"), uuid = "fa8ffc8b-3c10-4a20-ac8d-56f5f6784817")
```


```{r global-options, include=FALSE}
knitr::opts_chunk$set(
  dev = c("png", "cairo_pdf"),
  fig.ext = c("png", "pdf"),
  fig.width = 10, 
  fig.height = 6, 
  dpi = 200,
  fig.path = paste0('figures/', params, "/runtime_"),
  echo = FALSE)

## set TRUE to recompute runtimes (slow), FALSE to just read from disk
refit_runtime <- TRUE
```

This notebook benchmarks runtime for:

1. **dacomp only** (permutation testing + DS-FDR),
2. **dacomp + permApprox** (adds tail approximation on top of dacomp’s permutation statistics),

   * permApprox constrained GPD with **SLLS** constraint and **exceed0 = 250**
   * permApprox constrained GPD with **SLLS** constraint and **exceed0 = 0.25** (run only for 1e3 and 1e4)

For the combined methods, we report **total runtime = dacomp runtime + permApprox runtime**, using the same permutation statistics matrix returned by dacomp.

## CPU info

```{r, eval = FALSE}
cpuinfo <- readLines("/proc/cpuinfo")

model_name <- sub("^model name\\s*:\\s*", "", cpuinfo[grep("^model name\\s*:", cpuinfo)[1]])
mhz_vals   <- as.numeric(sub("^cpu MHz\\s*:\\s*", "", cpuinfo[grep("^cpu MHz\\s*:", cpuinfo)]))

list(
  model_name = model_name,
  mhz_median = median(mhz_vals, na.rm = TRUE),
  mhz_mean   = mean(mhz_vals, na.rm = TRUE),
  n_cores_reported = length(mhz_vals)
)
```


## Packages

```{r packages}
library(conflicted)
library(phyloseq)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(ggplot2)
library(kableExtra)

library(dacomp)
library(permApprox)

conflicted::conflicts_prefer(
  dplyr::count, dplyr::filter, dplyr::select, dplyr::slice,
  stats::var
)
```

## Directories & data

```{r directories}
# create or *empty* the target directory, used to write this file's data: 
#projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

```{r load_data}
bact_phylo <- readRDS(path_source("PASTURE_data", "bact_2m_genus.rds"))
sampdata   <- sample_data(bact_phylo) %>% as("data.frame")

## Rename "24_Enterobacteriaceae(F)"
taxtab <- tax_table(bact_phylo)
Genus <- taxtab[, "Genus"]
Genus[Genus == "24_Enterobacteriaceae(F)"] <- "Enterobacteriaceae(F)"
taxtab[, "Genus"] <- Genus
tax_table(bact_phylo) <- taxtab
taxa_names(bact_phylo) <- Genus
```

## Minimal preprocessing 

```{r preprocessing}
## drop taxa with total 0
bact_phylo <- prune_taxa(taxa_sums(bact_phylo) > 0, bact_phylo)

## 1% prevalence filter
prev_threshold <- 0.01 * nsamples(bact_phylo)
taxa_prev <- apply(otu_table(bact_phylo), 1, function(x) sum(x > 0))
keep_taxa <- taxa_prev >= prev_threshold
bact_phylo <- prune_taxa(keep_taxa, bact_phylo)

## define groups: EBF vs non-EBF
sampdata <- sample_data(bact_phylo) %>% as("data.frame") %>%
  mutate(
    BF_cat = case_when(
      breast_excl_cat1 == "0"   ~ "non-EBF",
      breast_excl_cat1 == ">=2" ~ "EBF",
      TRUE ~ NA_character_
    )
  )
sample_data(bact_phylo) <- sampdata

label1 <- "non-EBF"
label2 <- "EBF"

var_vec <- sampdata[["BF_cat"]]
names(var_vec) <- rownames(sampdata)

group1_ids <- rownames(sampdata)[var_vec %in% label1]
group2_ids <- rownames(sampdata)[var_vec %in% label2]

bact_g1 <- prune_samples(group1_ids, bact_phylo)
bact_g2 <- prune_samples(group2_ids, bact_phylo)

counts_g1 <- t(as(otu_table(bact_g1), "matrix"))
counts_g2 <- t(as(otu_table(bact_g2), "matrix"))

X_counts <- rbind(counts_g1, counts_g2)
y_groups <- factor(
  c(rep(label1, nrow(counts_g1)), rep(label2, nrow(counts_g2))),
  levels = c(label1, label2)
)

dim(X_counts)
table(y_groups)
```

## Reference selection (dacomp)

```{r dacomp_refs}
set.seed(42)

res_refs <- dacomp.select_references(
  X = X_counts,
  minimal_TA = 50,
  verbose = TRUE
)
length(res_refs$selected_references)
```

## Runtime benchmarking

### Helper functions

```{r runtime_helpers}
q_DSFDR <- 0.1

run_dacomp <- function(n_perm) {
  dacomp.test(
    X                  = X_counts,
    y                  = y_groups,
    ind_reference_taxa = res_refs,
    test               = DACOMP.TEST.NAME.WILCOXON,
    nr_perm            = n_perm,
    q                  = q_DSFDR,
    verbose            = FALSE,
    return_rarefied_values = FALSE,  # runtime only
    Test_All           = TRUE
  )
}

## Build obs_stats / perm_stats from a dacomp result
extract_stats_for_permApprox <- function(res_dacomp) {
  stats_mat  <- res_dacomp$stats_matrix
  obs_stats  <- stats_mat[1, ]
  perm_stats <- stats_mat[-1, , drop = FALSE]
  list(
    obs_stats = obs_stats,
    perm_stats = perm_stats
  )
}

run_permApprox_constrained <- function(obs_stats, perm_stats, sampsize,
                                      exceed0, exceed0_min = 250) {

  gpd_ctrl_constr <- make_gpd_ctrl(
    fit_method    = "lme",
    include_obs   = FALSE,
    constraint    = "support_at_obs",
    sample_size   = sampsize,
    thresh_method = "rob_ftr",
    exceed0       = exceed0,
    exceed0_min   = exceed0_min,
    exceed_min    = 10,
    thresh_step   = 10
  )

  perm_approx(
    obs_stats     = obs_stats,
    perm_stats    = perm_stats,
    alternative   = "two_sided",
    approx_thresh = 0.1,
    method        = "gpd",
    gpd_ctrl      = gpd_ctrl_constr,
    adjust_method = "BH",
    cores         = 1,
    verbose       = FALSE
  )
}
```

### Measure runtimes

```{r measure_runtimes, eval=FALSE}
## Settings requested
perm_grid_all  <- c(1e3, 1e4, 1e5, 1e6)          # for dacomp and permApprox(250)
perm_grid_025  <- c(1e3, 1e4, 1e5)               # for permApprox(25%)
new_perm_extrap <- c(1e7, 1e8)                   # extend lines to 1e8

sampsize <- min(nrow(counts_g1), nrow(counts_g2))  # for SLLS rule

## store results
dacomp_res_list <- list()

## --- 1) dacomp runtimes + store results (needed for permApprox timings) ----
runtime_dacomp_tbl <- purrr::map_dfr(perm_grid_all, function(n_perm) {
  set.seed(42)
  cat("dacomp n_perm:", as.character(n_perm), "\n")

  t <- system.time({
    dacomp_res_list[[as.character(n_perm)]] <<- run_dacomp(n_perm)
  })

  tibble(
    method    = "dacomp",
    setting   = "dacomp",
    n_perm    = n_perm,
    elapsed_s = unname(t["elapsed"])
  )
})

## --- 2) permApprox constrained, exceed0 = 250 (run for all) ----------------
runtime_pApp_250_tbl <- purrr::map_dfr(perm_grid_all, function(n_perm) {
  set.seed(42)
  cat("permApprox(250) n_perm:", as.character(n_perm), "\n")

  res_dacomp <- dacomp_res_list[[as.character(n_perm)]]
  st <- extract_stats_for_permApprox(res_dacomp)

  ## IMPORTANT: set colnames for perm_stats (permApprox expects them to match obs)
  colnames(st$perm_stats) <- names(st$obs_stats)

  t <- system.time({
    invisible(
      run_permApprox_constrained(
        obs_stats  = st$obs_stats,
        perm_stats = st$perm_stats,
        sampsize   = sampsize,
        exceed0    = 250,
        exceed0_min = 250
      )
    )
  })

  tibble(
    method    = "dacomp + permApprox",
    setting   = "SLLS, exceed0 = 250",
    n_perm    = n_perm,
    elapsed_s = unname(t["elapsed"])
  )
})

## --- 3) permApprox constrained, exceed0 = 0.25 (run only for 1e3, 1e4) -----
runtime_pApp_025_tbl <- purrr::map_dfr(perm_grid_025, function(n_perm) {
  set.seed(42)
  cat("permApprox(0.25) n_perm:", as.character(n_perm), "\n")

  res_dacomp <- dacomp_res_list[[as.character(n_perm)]]
  st <- extract_stats_for_permApprox(res_dacomp)
  colnames(st$perm_stats) <- names(st$obs_stats)

  t <- system.time({
    invisible(
      run_permApprox_constrained(
        obs_stats  = st$obs_stats,
        perm_stats = st$perm_stats,
        sampsize   = sampsize,
        exceed0    = 0.25,
        exceed0_min = 250
      )
    )
  })

  tibble(
    method    = "dacomp + permApprox",
    setting   = "SLLS, exceed0 = 0.25",
    n_perm    = n_perm,
    elapsed_s = unname(t["elapsed"])
  )
})

## Combine and compute TOTAL runtime for combined methods = dacomp + permApprox
runtime_raw_tbl <- bind_rows(runtime_dacomp_tbl, runtime_pApp_250_tbl, runtime_pApp_025_tbl)

# 1) dacomp baseline per n_perm
dacomp_base_tbl <- runtime_raw_tbl %>%
  filter(method == "dacomp") %>%
  select(n_perm, elapsed_s_dacomp = elapsed_s)

# 2) total runtime rows (dacomp-only + dacomp+permApprox)
runtime_total_tbl <- runtime_raw_tbl %>%
  # attach dacomp baseline to ALL rows (including permApprox rows)
  left_join(dacomp_base_tbl, by = "n_perm") %>%
  mutate(
    elapsed_s_total = case_when(
      method == "dacomp" ~ elapsed_s_dacomp,                 # dacomp only
      TRUE              ~ elapsed_s_dacomp + elapsed_s       # dacomp + permApprox
    ),
    line = case_when(
      method == "dacomp" ~ "dacomp only",
      setting == "SLLS, exceed0 = 250" ~ "dacomp + permApprox (SLLS, exceed0=250)",
      setting == "SLLS, exceed0 = 0.25" ~ "dacomp + permApprox (SLLS, exceed0=0.25)",
      TRUE ~ paste0("dacomp + permApprox (", setting, ")")
    )
  ) %>%
  select(line, method, setting, n_perm, elapsed_s_total)

runtime_total_tbl

saveRDS(runtime_total_tbl, file = path_target("runtime_total_tbl.rds"))
saveRDS(runtime_dacomp_tbl, file = path_target("runtime_dacomp_tbl.rds"))
saveRDS(runtime_pApp_250_tbl, file = path_target("runtime_permApprox_250_tbl.rds"))
saveRDS(runtime_pApp_025_tbl, file = path_target("runtime_permApprox_025_tbl.rds"))
saveRDS(dacomp_res_list, file = path_target("dacomp_reslist_runtime.rds"))
```

```{r load_or_refit}
fname_total <- path_target("runtime_total_tbl.rds")

runtime_total_tbl <- readRDS(fname_total)

# Rename settings
runtime_total_tbl <- runtime_total_tbl %>%
  dplyr::mutate(
    line = dplyr::case_when(
      line == "dacomp + permApprox (SLLS, exceed0=0.25)" ~
        "dacomp + permApprox (SLLS, k0=0.25B)",
      line == "dacomp + permApprox (SLLS, exceed0=250)" ~
        "dacomp + permApprox (SLLS, k0=250)",
      TRUE ~ line
    )
  )

runtime_total_tbl
```


## Add 1e07

We add from another script the 1e07 permutations case. 

```{r}
runtime_tbl_other <- readRDS(path_source("02_micro_diff_abund/dacomp_runtime.rds"))
runtime_tbl_other
```

```{r}
runtime_total_tbl <- bind_rows(
  runtime_total_tbl,
  tibble(
    line = "dacomp only",
    method = "dacomp",
    setting = "dacomp",
    n_perm = 1e7,
    elapsed_s_total = as.numeric(runtime_tbl_other[5, "elapsed"])
  )
) %>%
  arrange(desc(line), n_perm)
```


```{r}
# Remove lines
runtime_total_tbl <- runtime_total_tbl[-9,]
```


```{r}
# Add minutes and hours
runtime_total_tbl <- runtime_total_tbl %>%
  mutate(
    elapsed_s   = round(elapsed_s_total, 2),
    elapsed_min = round(elapsed_s_total / 60, 2),
    elapsed_h   = round(elapsed_s_total / 3600, 3)
  ) %>%
  select(line, n_perm, elapsed_s, elapsed_min, elapsed_h) %>%
  arrange(desc(line), n_perm)
```

```{r}
kable(runtime_total_tbl)
```


## Plot

```{r}
runtime_plot_tbl <- runtime_total_tbl %>%
  mutate(
    elapsed_s   = round(elapsed_s, 2),
    elapsed_min = round(elapsed_s / 60, 2),
    elapsed_h   = round(elapsed_s / 3600, 3),
    line = factor(
      line,
      levels = c(
        "dacomp only",
        "dacomp + permApprox (SLLS, k0=250)",
        "dacomp + permApprox (SLLS, k0=0.25B)"
      )
    )
  ) %>%
  filter(!is.na(elapsed_s), !is.na(n_perm))
runtime_plot_tbl
```


```{r runtime_plot, fig.width=7.5, fig.height=3.6, dpi=300}
library(dplyr)
library(ggplot2)
library(scales)

# --- custom grids you want ---
x_grid <- sort(unique(runtime_plot_tbl$n_perm))

minute_grid <- c(1, 2, 5, 10, 30, 60, 120, 300)
second_grid <- c(1, 3, 10, minute_grid * 60)

ggplot(runtime_plot_tbl, aes(x = n_perm, y = elapsed_s, colour = line, group = line)) +
  # horizontal minute-gridlines (in seconds)
  geom_hline(yintercept = second_grid, color = "grey85", linewidth = 0.3) +
  # vertical gridlines at your n_perm grid
  geom_vline(xintercept = x_grid, color = "grey90", linewidth = 0.3) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.2, stroke = 0.9) +
  scale_x_log10(
    breaks = x_grid,
    labels = scales::label_scientific()
  ) +
  scale_y_log10(
    name = "Runtime (seconds)",
    breaks = second_grid,                  # <- ticks/grid match your custom seconds
    labels = scales::label_number(),
    sec.axis = sec_axis(
      trans  = ~ . / 60,
      name   = "Runtime (minutes)",
      breaks = minute_grid,                # <- custom minutes, aligned
      labels = scales::label_number()
    )
  ) +
  scale_colour_manual(
    values = c(
      "dacomp only" = "black",
      "dacomp + permApprox (SLLS, k0=250)" = "darkorange2",
      "dacomp + permApprox (SLLS, k0=0.25B)" = "#1f77b4"
    )
  ) +
  labs(x = "Number of permutations", colour = NULL) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    # optional: turn off default y major grid so only your custom hlines show
    panel.grid.major.y = element_blank(),
    legend.position = c(1, 0.1),
    legend.justification = c("right", "bottom"),
    legend.background = element_rect(fill = "white", color = "grey80", linewidth = 0.3),
    legend.key = element_rect(fill = "white")
  )

```


